### 2017-05-17 paper.tex
(*) First commit.

% Sample document for SBGames papers
% Uses a slightly modified IEEE VGTC template in conference mode

\documentclass{vgtc}                          % final (conference style)

%% These three lines bring in essential packages: ``mathptmx'' for Type 1 
%% typefaces, ``graphicx'' for inclusion of EPS figures. and ``times''
%% for proper handling of the times font family.

\usepackage{mathptmx}
\usepackage{graphicx}
\usepackage{times}
\usepackage{xspace}
\usepackage{url}
\usepackage{verbatim}

\usepackage{listings}
\usepackage{color}
    \definecolor{light}{gray}{0.97}
    \definecolor{dark}{gray}{0.30}
\lstset{
%columns=fullflexible,
%basicstyle=\ttfamily,
escapeinside={||},
    %mathescape=true,
    language=C, % choose the language of the code
    basicstyle=\fontfamily{pcr}\selectfont\scriptsize\color{black},
    keywordstyle=\color{black}\bfseries, % style for keywords
    numbers=none, % where to put the line-numbers
    numberstyle=\tiny, % the size of the fonts that are used for the line-numbers
    backgroundcolor=\color{light},
    showspaces=false, % show spaces adding particular underscores
    showstringspaces=false, % underline spaces within strings
    showtabs=false, % show tabs within strings adding particular underscores
    %frame=single, % adds a frame around the code
    tabsize=2, % sets default tabsize to 2 spaces
    %rulesepcolor=\color{gray}
    captionpos=b, % sets the caption-position to bottom
    breaklines=false, % sets automatic line breaking
    %breakatwhitespace=false,
    numbersep=2em,
    % C was used in the blocksworld example to refer to block C and nowhere else
    emph={par,or,hor,do,end,loop,await,emit,input,event,call,with,%
          var,and,then,else,return,pure,deterministic,nohold,finalize,%
          class, every, FOREVER, this, spawn, in, pool, watching, until, 
          interface, each, abort, when, signal, PROC, CHAN, SIGNAL, PAR, not,
          bool, data, tag, escape, new, traverse,implementation,output,
          native,@const,@pure,@safe,define},
    emphstyle={\bfseries},
    commentstyle=\color{dark}\scriptsize,
    %xleftmargin=20pt,
    %xrightmargin=20pt,
    framesep=20pt,
    %upquote=true,
    %aboveskip={1.5\baselineskip},
}

\newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
\newcommand{\code}[1] {{\small{\texttt{#1}}}}
\newcommand{\ax}{\code{[a]}\xspace}
\newcommand{\bx}{\code{[b]}\xspace}

%% We encourage the use of mathptmx for consistent usage of times font
%% throughout the proceedings. However, if you encounter conflicts
%% with other math-related packages, you may want to disable it.

%% If you are submitting a paper to a conference for review with a double
%% blind reviewing process, please replace the value ``0'' below with your
%% OnlineID. Otherwise, you may safely leave it at ``0''.
\onlineid{0}

%% declare the category of your paper, only shown in review mode
\vgtccategory{Research}

%% Paper title.

\title{Structured Synchronous Reactive Programming for Game Development
        \\ \Large{Case Study: On Rewriting Pingus from C++ to \CEU}}

\author{Francisco Sant'Anna
        \\ Departamento de Inform\'atica e Ci\^encia da Computa\c{c}\~ao, UERJ
        \\ francisco@ime.uerj.br}

\abstract{Abstract.

\smallskip

\noindent \textbf{Keywords:} Radiosity, global illumination, constant time.
}

%% Copyright space is enabled by default as required by guidelines.
%% It is disabled by the 'review' option or via the following command:
% \nocopyrightspace

\begin{document}

\firstsection{Introduction}

\maketitle

Pingus%
\footnote{Pingus: \url{http://pingus.seul.org/}}
is an open-source clone of Lemmings%
\footnote{Lemmings: \url{https://en.wikipedia.org/wiki/Lemmings_(video_game)}},
a puzzle-platformer video game.      
The objective of the game is to guide a group of penguins through a number of
obstacles towards a designated exit%
\footnote{Pingus gameplay: \url{https://www.youtube.com/watch?v=MKrJgIFtJX0}}.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{pingus}
\caption{Pingus gameplay.
\label{fig.sweeney}
}
\end{figure}

Pingus is developed in standard object-oriented C++, the \emph{lingua franca}
of game development \cite{games.patterns}.
The codebase is about 40.000 lines of code (LoCs)%
\footnote{Pingus repository: \url{https://github.com/Pingus/pingus/tree/7b255840c201d028fd6b19a2185ccf7df3a2cd6e/src}}, divided into
the engine, level editor, auxiliary libraries, and the game logic itself.

According to Tim Sweeney (of Unreal Engine fame), about half the complexity in
game development resides in \emph{simulation} (aka \emph{game logic}), but
which accounts for only 10\% of the CPU budget~\cite{games.sweeney}.
The game logic ``models the state of the game world as interacting objects
evolve over time''.
The high development costs contrasting with the low impact on performance
appeals for alternatives with productivity in mind, especially considering that
it is the game logic that varies the most between projects.
Sweeney states that ``will gladly sacrifice 10\% of our performance for 10\%
higher productivity''.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{sweeney}
\caption{Three kinds of code~\cite{games.sweeney}.
\label{fig.sweeney}
}
\end{figure}

Object-oriented games use the \emph{observer pattern}~\cite{games.patterns}
in the game logic to handle events from the environment (e.g., key presses and
timers) and also as a notification mechanism between game entities.
%
The observers are short-lived callbacks which must execute as fast as possible
and in real time to keep the game reactive to incoming events.
%
For this reason, callbacks cannot contain long-lasting locals and loops, which
are elementary capabilities of classical structured
programming~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
%
In this sense, callbacks actually disrupt structured programming, becoming
``our generation's \code{goto}''.%
\footnote{``Callbacks as our Generations' Go To Statement'':
\url{http://tirania.org/blog/archive/2013/Aug-15.html}}%
\footnote{``Escape from Callback Hell'':
\url{http://elm-lang.org/learn/Escape-from-Callback-Hell.elm}}

\CEU~\cite{ceu.sensys13,ceu.mod15} is a programming language that aims to offer
a concurrent and expressive alternative to C/C++ with the characteristics that
follow:
%
\begin{itemize}
\item \emph{Reactive}: code only executes in reactions to events.
\item \emph{Structured}: programs use structured control mechanisms, such as
      \code{await} (to suspend a line of execution), and \code{par} (to combine
      multiple lines of execution).
\item \emph{Synchronous}: reactions run atomically and to completion on each
      line of execution, i.e., there's no implicit preemption or real
      parallelism.
\end{itemize}
%
Structured reactive programming eliminates callbacks, letting programmers write
code in direct and sequential style and recover from the inversion of control
imposed by the observer pattern~\cite{rp.deprecating}.
%
\CEU supports logical parallelism with a resource-efficient implementation in
terms of memory and CPU usage~\cite{ceu.sensys13}.
The runtime is single threaded and the language requires no garbage
collection.

Contributions:
    - patterns
    - solutions

\section{Control-Flow Patterns}

The rewriting process consisted of identifying sets of callbacks implementing
*control-flow behaviors* in the game and translating them to \CEU using
appropriate structured constructs.
As an example, a double mouse click is characterized by a first click, followed
by a maximum amount of time, followed by a second click.
This behavior depends on different events (clicks and timers) which have to
occur in a particular order.
In C++, the implementation involves callbacks crossing reactions to successive
events which manipulate state variables explicitly.

We can identify control-flow behaviors in C++ by looking for class members with
identifiers resembling verbs, statuses, and counters (e.g.,
\code{pressed},
\code{particle\_thrown},
\code{mode}, and
\code{delay\_count}).
Good chances are that variables with these ``suspicious names'' encode some
form of control-flow progression that cross multiple callback invocations.

We selected 9 representative game behaviors and describe their
implementations in C++ and \CEU.
We also categorized these examples in 5 abstract C++ control-flow patterns that
likely apply to other games:

\begin{enumerate}
\item \emph{Finite State Machines}:
    State machines describe the behavior of entities by mapping event
    occurrences to transitions between states that trigger appropriate actions.
\item \emph{Continuation Passing}:
    The completion of a long-lasting activity may carry a continuation, i.e.,
    some action to execute next.
\item \emph{Dispatching Hierarchies}:
    Entities typically form a dispatching hierarchy in which a container that
    receives a stimulus automatically forwards it to its managed children.
\item \emph{Lifespan Hierarchies}:
    Entities typically form a lifespan hierarchy in which a terminating
    container entity automatically destroys its managed children.
\item \emph{Signaling Mechanisms}:
    Entities often need to communicate explicitly through signaling mechanisms,
    especially if there is no hierarchy relationship between them.
\end{enumerate}

\subsection{Finite State Machines \\ Case Study: The Armageddon Button Double Click}
%\label{sec.patterns.fsm}

State machines describe the behavior of entities by mapping event occurrences
to transitions between states that trigger appropriate actions.

%\subsection{The Armageddon Button Double Click}
%\label{sec.patterns.fsm.armageddon}

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{double-click-opt}
\caption{Double click detection.
\label{fig.armageddon}
}
\end{figure}

In Pingus, a double click in the \emph{Armageddon} button at the bottom right
of the screen literally explodes all pingus (Figure~\ref{fig.armageddon}).
Figure~\ref{fig.armageddon.code} compares the implementations in C++ and \CEU.

\begin{figure*}
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
ArmageddonButton::ArmageddonButton(<...>):
    RectComponent(<...>),
    pressed(false); // button initially not pressed
    press_time(0);    // how long since 1st click?
    <...>
{
    <...>
}

void ArmageddonButton::draw (<...>) {
    <...>
}

void ArmageddonButton::update (float delta) {
    <...>
    if (pressed) {
        press_time += delta;
        if (press_time > 1.0f) {
            pressed = false; // give up, 1st click
            press_time = 0;  // was too long ago
        }
    } else {
        <...>
        press_time = 0;
    }
}

void ArmageddonButton::on_click (<...>) {
    if (pressed) {
        server->send_armageddon_event();
    } else {
        pressed = true;
    }
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
do
    var& RectComponent c = <...>;
    <...>
    loop do
        await c.component.on_click;
        watching 1s do
            await c.component.on_click;
            break;
        end
    end
    <...>
    emit outer.game.go_armageddon;
end

.
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different 
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure*}

In C++, the class \code{ArmageddonButton} implements methods for rendering the
button and handling mouse and timer events.
The listing focus on the double click detection, hiding unrelated parts with
\code{<...>}.
%
The methods \code{update} (ln. 14--26) and \code{on\_click} (ln. 28--34) are
examples of \emph{short-lived callbacks}, which are pieces of code that execute
atomically in reaction to external input events.
The callback \code{on\_click} reacts to mouse clicks detected by the button
base class \code{RectComponent} (ln. 2), while the callback \code{update}
continuously reacts to the passage of time, frame by frame.
Callbacks are short lived because they must react to input as fast as possible
to let other callbacks execute, keeping the game with real-time responsiveness.
%
The class first initializes the variable \code{pressed} to track the first
click (ln. 3,32).
It also initializes the variable \code{press\_time} to count the time since the
first click (ln. 4, 17).
If another click occurs within 1 second, the class signals the double click to
the application (ln. 30).
Otherwise, the \code{pressed} and \code{press\_time} state variables are reset
(ln. 19--20). 
%
Figure~\ref{fig.armageddon.fsm} illustrates how we can model the double-click 
behavior in C++ as a state machine.
The circles represent the state of the variables in the class, while the arrows 
represent the callbacks manipulating state.
%
Note in the code how the accesses to the state variables are spread
across the entire class.
For instance, the distance between the initialization of \code{pressed} (ln.
3) and the last access to it (ln. 32) is over 40 lines in the original file.
Arguably, this dispersion of code across methods makes the understanding and 
maintenance of the double-click behavior more difficult.
Also, even though the state variables are private, unrelated methods such as 
\code{draw}`, which is defined in middle of the class (ln. 10--12), can
potentially access them.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{double-click}
\caption{State machine for the \emph{Armageddon} double click.
\label{fig.armageddon.fsm}
}
\end{figure}

\CEU provides structured constructs to deal with events, aiming to eradicate
explicit manipulation of state variables for control-flow purposes.
%
The loop detection (ln. 4--10) awaits the first click (ln. 5) and then, while
watching 1 second (ln. 6--9), awaits the second click (ln. 7).
If the second click occurs within 1 second, the \code{break} terminates the
loop (ln. 8) and the \code{emit} signals the double click to the application
(ln. 12).
Otherwise, the \code{watching} block as a whole aborts and restarts the loop, 
falling back to the first click \code{await} (ln. 5).
%
Double click detection in \CEU doesn't require state variables and is entirely
self-contained in the \code{loop} body (ln. 4--10).
Furthermore, these 7 lines of code \emph{only} detect the double click, leaving
the actual effect to happen outside the loop (ln. 12).

\subsection{Continuation Passing \\ Case Study: Advancing Pages in the Story Screen}

The completion of a long-lasting activity may carry a continuation, i.e., some
action to execute next.

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{story-anim}
\caption{The Story screen.
\label{fig.story}
}
\end{figure}

The clickable \emph{blue dots} in the campaign world map transit to ambience
story screens (Figure~\ref{fig.story}).
A story is composed of multiple pages and, inside each page, the words of the
story appear incrementally over time.
A first click in the button \code{>>>} fast forwards the words to show the full 
page.
A second click advances to the next page, until the story terminates.
If the page completes before a click (due to the time elapsing), a first click 
advances to the next page.

\begin{figure*}
\begin{minipage}[t]{0.55\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
StoryScreenComponent::StoryScreenComponent (<...>) :
    <...>
{
    pages        = <...>; // vector with loaded pages
    current_page = pages.back(); // first loaded page
    displayed    = false; // if current is complete
    <...>
}

<...>   // draw page over time

void StoryScreenComponent::update (<...>) {
    <...>
    if (&lt;all-words-appearing&gt;) {
        displayed = true;
    }
}

void StoryScreenComponent::next_text() {
    if (!displayed) {
        displayed = true;
        <...>     // remove current page
    } else {
        pages.pop_back();
        if (!pages.empty()) { // next page
            current_page = pages.back();
            displayed    = false;
            <...>
        } else {
            <...> // terminates the story screen
        }
    }
}
\end{lstlisting}
\centering\small{\ax Implementation in C++}
\end{minipage}
%
\begin{minipage}[t]{0.45\linewidth}
\begin{lstlisting}[numbers=left,xleftmargin=3em]
code/await Story (void) -> bool do
    <...>
    event void next_text; // clicks in >>>

    { pages = <...>; } // same as in C++
    loop i in [0 <- {pages.size()}[ do
        par/or do
            watching next_text do
                <...> // advance text
            end
            await next_text;
        with
            <...> // redraw _pages[i]
        end
    end
end

.
\end{lstlisting}
\centering\small{\bx Implementation in \CEU}
\end{minipage}
%\rule{8.4cm}{0.37pt}
\caption{ Shared-memory concurrency in \CEU:
example \ax is safe because the trails access \code{x} atomically in different 
reactions;
example \bx is unsafe because both trails access \code{y} in the same reaction.
\label{lst.shared}
}
\end{figure*}

\begin{figure}
\centering
\includegraphics[width=\columnwidth]{story}
\caption{State machine for the Story screen.
\label{fig.story}
}
\end{figure}

In C++, the class \code{StoryScreenComponent} implements the method
\code{next\_text}, which is a callback for clicks in \code{>>>}.
%
The variable `pages` (ln. 4--5, 24--26) is a vector holding each page, but
which also encodes \emph{continuations} for the story progress:
each call to \code{next\_text} that advances the story (ln. 23--32) removes the 
current page (ln. 24) and sets the next action to perform (i.e., ``display a
new page'') in the variable \code{current\_page} (ln. 26).
Figure~\ref{fig.story} illustrates the continuation mechanism to advance 
pages and also a state machine for fast forwarding words (inside the dashed
rectangle).
The state variable \code{displayed} (ln. 6,15,20,21,27) switches between the
behaviors ``advancing text'' and ``advancing pages'', which are both handled
intermixed inside the method \code{next\_text}.

The code in \CEU uses the internal event \code{next\_text}, which is emitted
from clicks in \code{>>>}.
%
The sequential navigation from page to page uses a loop in direct style
(ln. 6--15) instead of explicit state variables for the continuation and state
machine.
While the text advances in an inner loop (hidden in ln. 9), we watch the
\code{next\_text} event that fast forwards it.
The loop may also eventually terminate with the time elapsing normally.
This way, we do not need a variable (such as `displayed` in C++) to switch 
between the states ``advancing text'' and ``advancing pages''.
The \code{par/or} makes the page advance logic to execute in parallel with the
redrawing code (ln. 13).
Whenever the page advances, the redrawing code is automatically aborted
(due to the \code{or} modifier).
The \code{await next\_text} in sequence (ln. 11) is the condition to advance to
the next page.
%
Note that, unlike the implementation in C++, the ``advancing text'' behavior is
not intermixed with the ``advancing pages'' behavior, instead, it is
encapsulated inside the inner loop nested with a deeper indentation (ln. 9).

\bibliographystyle{abbrv}
\bibliography{my,other}
\end{document}

### 2017-05-18 paper.tex
...

@@
 
+\subsection{Dispatching Hierarchies \\ Case Study: Bomber `draw` and `update` callbacks}
+
+Entities typically form a dispatching hierarchy in which a container that
+receives a stimulus automatically forwards it to its managed children.
+
+\begin{figure*}
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+class Bomber : public PinguAction {
+    <...>
+    Sprite sprite;
+}
+
+Bomber::Bomber (<...>) : <...> {
+    sprite.load(<...>);
+    <...>
+}
+
+void Bomber::update () {
+    sprite.update ();
+}
+
+void Bomber::draw (SceneContext& gc) {
+    <...>
+    gc.color().draw(sprite, <...>);
+}
+\end{lstlisting}
+\centering\small{\ax Implementation in C++}
+\end{minipage}
+%
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+code/await Bomber (void) -> ActionName do
+    <...>
+    var&? Sprite sprite = spawn Sprite(<...>);
+    <...>
+end
+\end{lstlisting}
+\centering\small{\bx Implementation in \CEU}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ Shared-memory concurrency in \CEU:
+example \ax is safe because the trails access \code{x} atomically in different 
+reactions;
+example \bx is unsafe because both trails access \code{y} in the same reaction.
+\label{lst.shared}
+}
+\end{figure*}
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{story}
+\caption{State machine for the Story screen.
+\label{fig.story}
+}
+\end{figure}
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{hierarchy}
+\caption{Dispatching chain for \code{update}.
+\label{fig.hier}
+}
+\end{figure}
+
+In C++, the class \code{Bomber} declares a \code{sprite} member to handle its
+animation frames (Figure~\ref{fig.TODO}).
+%
+The \code{Sprite} class is part of the game engine and knows how to update and
+render itself.
+However, the \code{Bomber} still has to respond to \code{update} and
+\code{draw} requests from the game and forward them to the sprite
+(ln. 11--13 and 15--18).
+%
+To understand how the \code{update} callback flows from the original
+environment stimulus from the game down to the sprite, we need to follow a long
+chain of 7 method dispatches (Figure~\ref{fig.hier}):
+%
+\begin{enumerate}
+\item \code{ScreenManager::display} in the main game loop calls \code{update}.
+\item \code{ScreenManager::update} calls \code{last\_screen->update} for the
+      active game screen (a \code{GameSession} instance, considering the
+      \code{Bomber}).
+\item \code{GameSession::update} calls \code{world->update}.
+\item \code{World::update} calls \code{obj->update} for each object in the
+      world.
+\item \code{PinguHolder::update} calls \code{pingu->update} for each pingu
+      alive.
+\item \code{Pingu::update} calls \code{action->update} for the active pingu
+      action.
+\item \code{Bomber::update} calls \code{sprite.update}.
+      \code{Sprite::update} finally updates the animation frame.
+\end{enumerate}
+%
+Each dispatching step in the chain is necessary considering the game
+architecture:
+%
+\begin{itemize}
+\item With a single assignment to \code{last\_screen}, we can easily deactivate
+      the current screen and redirect all dispatches to a new screen.
+\item The \code{World} class manages and dispatches events to all game
+      entities, such as all pingus and traps, with the common interface
+      \code{WorldObj}.
+\item Since it is common to iterate only over the pingus (vs. all world
+      objects), the container \code{PinguHolder} manages all pingus.
+\item Since a single pingu can change between actions during lifetime, the
+      \code{action} member decouples them with another level of indirection.
+\item Sprites are part of the game engine and are reusable everywhere (e.g., UI
+      buttons, world objects, etc.), so it is also convenient to decouple them
+      from actions.
+\end{itemize}
+%
+The \code{draw} callback flows through the same dispatching hierarchy until
+reaching the \code{Sprite} class.
+
+In \CEU, the \code{Bomber} action spawns a \code{Sprite} animation instance on
+its body.
+%
+The \code{Sprite} instance (ln. 3) can react directly to external \code{dt} and
+\code{redraw} events (which are analogous to \code{update} and \code{redraw}
+callbacks, respectively), bypassing the program hierarchy entirely.
+While and \emph{only while} the bomber abstraction is alive, the sprite
+animation is also alive.
+The radical decoupling between the program hierarchy and reactions to events
+eliminates dispatching chains entirely.
+
+\subsection{Lifespan Hierarchies \\ Case Study: The Pingus Container}
+
+Entities typically form a lifespan hierarchy in which a terminating
+container entity automatically destroys its managed children.
+
+\begin{figure*}
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+Pingu* PinguHolder::create_pingu (<...>) {
+    <...>
+    Pingu* pingu = new Pingu (<...>);
+    pingus.push_back(pingu);
+    <...>
+}
+
+void PinguHolder::update() {
+    <...>
+    while(pingu != pingus.end()) {
+        (*pingu)->update();
+        if ((*pingu)->get_status() == Pingu::PS_DEAD) {
+            pingu = pingus.erase(pingu);
+        }
+        <...>
+        ++pingu;
+    }
+}
+\end{lstlisting}
+\centering\small{\ax Implementation in C++}
+\end{minipage}
+%
+\begin{minipage}[t]{0.40\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+code/await Game (void) do
+    <...>
+    pool[] Pingu pingus;
+    code/await Pingu_Spawn (<...>) do
+        <...>
+        spawn Pingu(<...>) in outer.pingus;
+    end
+    <...>   // code invoking Pingu_Spawn
+end
+
+TODO
+
+code/await Pingu (<...>) do
+    <...>
+    loop do
+        await outer.game.dt;
+        if call Pingu_Rel_Getpixel(0,-1) == {Groundtype::GP_OUTOFSCREEN} then
+            <...>
+            escape {PS_DEAD};
+        end
+    end
+end
+\end{lstlisting}
+\centering\small{\bx Implementation in \CEU}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ Shared-memory concurrency in \CEU:
+example \ax is safe because the trails access \code{x} atomically in different 
+reactions;
+example \bx is unsafe because both trails access \code{y} in the same reaction.
+\label{lst.shared}
+}
+\end{figure*}
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{story}
+\caption{State machine for the Story screen.
+\label{fig.story}
+}
+\end{figure}
+
+In C++, the class \code{PinguHolder} is a container that holds all pingus
+alive.
+%
+The method \code{PinguHolder::create\_pingu} (ln. 1--6) is called periodically
+to create a new \code{Pingu} and add it to the \code{pingus} collection
+(ln. 3--4).
+The method \code{PinguHolder::update} (ln. 8--18) checks the state of all
+pingus on every frame, removing those with the dead status (ln. 12--14).
+%
+Entities with dynamic lifespan in C++ require explicit \code{add} and
+\code{remove} calls associated to a container (ln. 4,13).
+Without the \code{erase} call above, a dead pingu would remain in the
+collection with updates on every frame (ln. 11).
+Since the \code{redraw} behavior for a dead pingu is innocuous, the death could
+go unnoticed but the program would keep consuming memory and CPU time.
+This problem is known as the \emph{lapsed listener}~\cite{TODO} and also occurs
+in languages with garbage collection:
+A container typically holds a strong reference to a child (sometimes the only 
+reference to it), and the runtime cannot magically detect it as garbage.
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{pool}
+\caption{Lifespan of dynamic instances.
+\label{fig.pool}
+}
+\end{figure}
+
+\CEU supports \code{pool} declarations to hold dynamic abstraction instances.
+Additionally, the \code{spawn} statement supports a pool identifier to
+associate the new instance with a pool.
+%
+The game screen spawns a new \code{Pingu} on every invocation of
+\code{Pingu\_Spawn}.
+%
+The \code{spawn} statement (ln. 6) specifies the pool declared at the top-level
+block of the game screen (ln. 3).
+In this case, the lifespan of the new instances follows the scope of the pool
+(ln. 1--9) instead of the enclosing scope of the \code{spawn} statement
+(ln. 4--7).
+Since pools are also subject to lexical scope, the lifespan of all dynamically
+allocated pingus is constrained to the game screen.
+%
+Lexical scopes handle memory and event dispatching automatically for static
+instances and also for pools.
+However, the lifespan of a dynamic instance does not necessarily have to match
+the lifespan of its associated pool (Figure~\ref{fig.pool}).
+In \CEU, when the execution block of a dynamic instance terminates, which
+characterizes its *natural termination*, the instance is automatically removed
+from its pool.
+Therefore, dynamic instances do not require any extra bookkeeping related to 
+containers or explicit deallocation.
+%
+To remove a pingu from the game in \CEU, we just need to terminate its execution
+block according to the appropriate conditions:
+%
+The \code{escape} statement (ln. TODO) aborts the execution block of the
+\code{Pingu} instance, removing it from its associated pool automatically.
+Hence, a dynamic instance that terminates naturally leaves no traces in the 
+program.
+
+\subsection{Signaling Mechanisms\\ Case Study: Global Keys and the Options Menu}
+
+Entities often need to communicate explicitly through signaling mechanisms,
+especially if there is no hierarchy relationship between them.
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{options-anim-opt}
+\caption{The \emph{Mouse Grab} configuration option.
+\label{fig.options}
+}
+\end{figure}
+
+The *Mouse Grab* option restricts the mouse movement to the game window
+boundaries (Figure~\ref{fig.options}).
+The option can be set anywhere in the game by pressing *Ctrl-G*.
+In addition, the *Options* menu has a check box to toggle the *Mouse Grab*
+option with mouse clicks while still responding to *Ctrl-G* presses.
+
+\begin{figure}
+\centering
+\includegraphics[width=\columnwidth]{options-anim-opt}
+\caption{Mutual dependency between signals.
+\label{fig.events}
+}
+\end{figure}
+
+The implementations in C++ and \CEU use a signalling mechanism to connect the
+key presses, the check box, and a configuration manager that applies the
+appropriate side effects in the game (i.e., restrict the mouse movement).
+Figure~\ref{fig.events} illustrates how the mutual notifications create a 
+dependency cycle between the configuration manager and the check box.
+
+\begin{figure*}
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+void GlobalEvent::on_button_press (<...>) {
+    <...>
+    switch (event.keysym.sym) {
+        case SDLK_g:
+            if (keystate[SDLK_LCTRL] || keystate[SDLK_RCTRL]) {
+                config_manager.set_mouse_grab(          @ctrl_g_cpp
+                    !config_manager.get_mouse_grab());
+            }
+            break;
+        <...>
+    }
+}
+
+///
+
+boost::signals2::signal<void(bool)> on_mouse_grab_change;   // definition in `config_manager.h` @signal_def
+
+void ConfigManager::set_mouse_grab (bool v) {   @set_mouse_grab
+    <...>
+    if (v != get_mouse_grab()) {                @if_1
+        <...>   // the actual "grab" effect
+        on_mouse_grab_change(v);                @signal
+    }                                           @if_2
+}
+
+///
+
+boost::signals2::signal<void (bool)> on_change;   // definition in `check_box.hpp`
+
+void CheckBox::set_state (bool is_on, bool send_signal) {   @last_argument
+    <...>   // switches the check box state
+    if (send_signal) {      @if_cb_1
+        on_change(is_on);
+    }                       @if_cb_2
+}
+
+///
+
+typedef std::vector&lt;boost::signals2::connection&gt; Connections;   // definition in `option_menu.hpp`
+Connections connections;                                        // definition in `option_menu.hpp`
+
+OptionMenu::OptionMenu() :
+    connections(),
+    mousegrab_box(),
+    <...>
+{
+    mousegrab_box = new CheckBox(<...>);
+    connections.push_back(                              @bind_11
+        config_manager.on_mouse_grab_change.connect(
+            std::bind(&CheckBox::set_state, mousegrab_box, <...>, false); @bind_false
+        )
+    );                                                  @bind_12
+    connections.push_back(                              @bind_21
+        mousegrab_box->on_change.connect(
+            std::bind(&ConfigManager::set_mouse_grab, &config_manager, <...>);
+        )
+    );                                                  @bind_22
+    <...>
+
+}
+
+OptionMenu::~OptionMenu() {     @destr_1
+    for (Connections::iterator i=connections.begin(); i!=connections.end(); ++i) {
+        (*i).disconnect();
+    }
+}                               @destr_2
+
+\end{lstlisting}
+\centering\small{\ax Implementation in C++}
+\end{minipage}
+%
+\begin{minipage}[t]{0.40\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+spawn do
+    var _SDL_KeyboardEvent&& e;
+    every e in SDL_KEYDOWN do
+        var _u8&& keystate = _SDL_GetKeyState(null);
+        <...>
+        if e:keysym.sym == {SDLK_g} then
+            if ((keystate[{SDLK_LCTRL}] as bool) or (keystate[{SDLK_RCTRL}] as bool)) then
+                emit config_manager.go_mouse_grab(    @ctrl_g_ceu
+                        not ({config_manager.get_mouse_grab()} as bool));
+            end
+        end
+        <...>
+    end
+end
+
+///
+
+data ConfigManager with
+    event bool go_mouse_grab;
+end
+var ConfigManager config_manager = val ConfigManager(_);
+
+spawn do
+    var bool v;
+    every v in config_manager.go_mouse_grab do
+        <...>   // the actual "grab" effect
+    end
+end
+
+///
+
+data ICheckBox with
+    var   bool is_on;
+    event bool go_click;
+end
+
+code/await CheckBox (<...>) -> (var ICheckBox checkbox) -> FOREVER do
+    checkbox = val ICheckBox(<...>);
+    <...>
+    par do
+        every c.component.on_click do                   @every_1
+            emit checkbox.go_click(not checkbox.is_on); @dir_class_app
+        end                                             @every_2
+    with
+        loop do                                         @loop_1
+            <...>   // switches the check box state
+            checkbox.is_on = await checkbox.go_click;   @dir_app_class
+        end                                             @loop_2
+    end
+end
+
+///
+
+code/await OptionMenu <...> do
+    <...>
+    var& CheckBox b2 = <...>;
+    spawn do
+        par do
+            var bool v;
+            every v in outer.config_manager.go_mouse_grab do    @loop_11
+                emit b2.checkbox.go_click(v);
+            end                                                 @loop_12
+        with
+            var bool v;
+            every v in b2.checkbox.go_click do                  @loop_21
+                emit outer.config_manager.go_mouse_grab(v);
+            end                                                 @loop_22
+        end
+    end
+    <...>
+end
+
+\end{lstlisting}
+\centering\small{\bx Implementation in \CEU}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ Shared-memory concurrency in \CEU:
+example \ax is safe because the trails access \code{x} atomically in different 
+reactions;
+example \bx is unsafe because both trails access \code{y} in the same reaction.
+\label{lst.shared}
+}
+\end{figure*}
+
+In C++, the class `GlobalEvent` detects *Ctrl-G*
+presses and invokes the callback `config\_manager.set\_mouse\_grab`:
+%
+The class `ConfigManager` uses a `boost::signal`
+[[![X]][boost\_signal]] to notify the application when the new configuration is
+applied:
+%
+The `if` enclosing the signal emission @NN(if\_1,-,if\_2) breaks the dependency 
+cycle of Figure~\ref{fig.events} and prevents an infinite execution loop.
+%
+The class `CheckBox` also uses a `boost::signal` to 
+notify the application on changes:
+%
+Again, the `if` enclosing the signal emission @NN(if\_cb\_1,-,if\_cb\_2) breaks the 
+dependency cycle of Figure~\ref{fig.events} to avoid infinite execution.
+%
+The class `OptionMenu` creates the dependency loop by
+connecting the two signals:
+%
+The constructor binds
+the signal `config\_manager.on\_mouse\_grab\_change` to the callback method
+           `mousegrab\_box->set\_state`
+           @NN(bind\_11,-,bind\_12),
+and also
+the signal `mousegrab\_box->on\_change` to the callback method
+           `config\_manager.set\_mouse\_grab`
+           @NN(bind\_21,-,bind\_22).
+This way, every time the `ConfigManager` signals `on\_mouse\_grab\_change`
+(`ConfigManager`, ln. @N(signal) [up](\#cpp\_config-manager)), `set\_state` is
+implicitly called.
+The same happens between the signal `on\_change` in the `CheckBox` and the 
+method `set\_mouse\_grab` in the `ConfigManager`
+(`ConfigManager`, ln. @N(set\_mouse\_grab) [up](\#cpp\_config-manager)).
+%
+Note that the signal binding to call `CheckBox::set\_state` @NN(bind\_false) 
+receives a fixed value `false` as the last argument to prevent infinite
+execution (`CheckBox`, ln. @N(last\_argument) [up](\#cpp\_check-box)).
+%
+The destructor @NN(destr\_1,-,destr\_2) breaks the connections explicitly when
+the *Option* screen terminates.
+
+In \CEU, a *Ctrl-G* key press broadcasts the internal event
+`config\_manager.go\_mouse\_grab` to the application [[![X]][ceu\_global\_event]]:
+%
+The configuration manager [[![X]][ceu\_config\_manager]] just needs to react to
+`go\_mouse\_grab` continuously to perform the *grab* effect:
+%
+The `CheckBox` [[![X]][ceu\_check\_box]] exposes the event `go\_click` for 
+notifications in both directions, i.e., from the abstraction to the application
+and *vice versa*:
+%
+The abstraction reacts to external clicks continuously @NN(every\_1,-,every\_2)
+to broadcast the event `go\_click` @NN(dir\_class\_app).
+It also reacts continuously to `go\_click` in another line of execution
+@NN(loop\_1,-,loop\_2), which awakes from notifications from the first line of
+execution or from the application.
+%
+The `OptionMenu` [[![X]][ceu\_option\_menu]] connects the two events as 
+follows:
+%
+The two loops in parallel handle the connections in opposite directions:
+from the configuration manager to the check box @NN(loop\_11,-,loop\_12);
+and
+from the check box to the configuration manager @NN(loop\_21,-,loop\_22).
+%
+When the *Option* screen terminates, the connections break automatically since
+the body is automatically aborted.
+%
+Note that the implementation in \CEU does not check event emits to break the
+dependency cycle and prevent infinite execution.
+Due to the [stack-based execution for internal events][ceu\_stack] in \CEU,
+programs with mutually-dependent events do not create infinite execution loops.
+
+\section{Related Work}
+
+\section{Conclusion}
+
+We promote the \emph{structured synchronous reactive} programming model of
+\CEU for the development of games.
+We present in-depth use cases categorized in 5 control-flow patterns applied to
+\emph{Pingus} (an open-source \emph{Lemmings} clone) that likely apply to other
+games.
+
+We show how the standard way to program games with objects and callbacks in C++
+hinders structured programming techniques, such as support for sequential
+execution, long-lasting loops, and persisting local variables.
+In this sense, callbacks actually disrupt structured programming, becoming
+["our generation’s goto"][goto] according to Miguel de Icaza.
+
+Overall, we believe that most difficulties in implementing control behavior in 
+game logic is not inherent to this domain, but a result of accidental
+complexity due to the lack of structured abstractions and an appropriate
+concurrency model to handle event-based applications.
+
+[goto]: http://tirania.org/blog/archive/2013/Aug-15.html
+
+\section{Acknowledgments}
+
+[Leonardo Kaplan](https://github.com/leokaplan/) and
+[Alexander Tkachov](https://github.com/Tkachov/)
+for early explorations and prototypes.
+
 \bibliographystyle{abbrv}

### 2017-05-19 paper.tex
...

@@
     % C was used in the blocksworld example to refer to block C and nowhere else
-    emph={par,or,hor,do,end,loop,await,emit,input,event,call,with,%
+    emph={par,or,hor,do,end,loop,code,await,emit,input,event,call,with,%
           var,and,then,else,return,pure,deterministic,nohold,finalize,%
@@
 
-\begin{figure}
+\begin{figure}[t]
 \centering
@@
 
+In this work, we advocate structured synchronous reactive programming as an
+expressive and productive alternative for game logic development.
+In Pingus, the game logic also accounts for almost half the size of the
+codebase (18.173 from 39.362 LoCs, or 46\%).
+
+rewriting Pingus to \CEU is to
+
 Contributions:
+    - present SSRP as an effective alternative for game logic dev
+    - qualitative analysis
         - patterns
             - solutions
+
+\section{The Pingus Codebase}
+
+In Pingus, the game logic also accounts for almost half the size of the
+codebase (18.173 from 39.362 LoCs, or 46\%).
+
+We rewrote 126 of the 272 files (46\%) which account for 9.186 of the 18.173
+LoCs (51\%) comprising the game logic of Pingus [[![X]][cpp\_uncompressed]].
+Half of the game logic relates to non-reactive code, such as configurations and
+options, saved games and serialization, maps and levels descriptions, string
+formatting, collision detection, graph algorithms, etc.
+This part remains unchanged and relies on the seamless integration between \CEU
+and C/C++ to remain usable.
+From the 9.186 touched LoCs, we removed all headers, declarations, trivial
+getters \& setters, and other innocuous statements, resulting in 70
+implementation files with 4.135 dense LoCs originally written in C++
+[[![X]][cpp\_compressed]].
+We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs
+[[![X]][ceu\_compressed]].
+Figure~\ref{tab.tree} summarizes the condensed codebase in the two
+implementations:
+
+This work focuses on a qualitative analysis for the programming techniques
+that we applied during the rewriting process.
+Not all techniques result in reduction in LoCs (especially considering the
+verbose syntax of \CEU), but have other effects such as eliminating shared
+variables and dependencies between classes.
+<!--, helping on encapsulation and cohesion.-->
+Nonetheless, the lowest ratio numbers above correlate to the parts of the game
+logic that we consider more susceptible to structured reactive programming.
+For instance, the *Pingu* behavior (*ratio 0.80*) contains complex animations
+that are affected by timers, game rules, and user interaction.
+In contrast, the *Option* screen (*ratio 0.97*) is a simple UI grid with
+trivial mouse interactions.
+
+\begin{figure*}[t]
+\begin{verbatim}
+Path              Ceu      C++   Ceu/C++    Descritpion
+------------     ----     ----     ----     --------------------------------------------
+game/            2064     2268     0.91     the main gameplay
+  ./              710      679     1.05         main functionality
+  objs/           470      478     0.98         world objects (tiles, traps, etc)
+  pingu/          884     1111     0.80         pingu behaviors
+    ./            343      458     0.75             main functionality
+    actions/      541      653     0.83             pingu actions (bomber, climber, etc)
+worldmap/         468      493     0.95     campaign worldmap
+screens/         1109     1328     0.84     menus and screens
+    option/       347      357     0.97         option menu
+    others/       762      971     0.78         other menus and screens
+misc/              56       46     1.22     miscellaneous functionality
+                 ----     ----     ----
+                 3697     4135     0.89
+\end{verbatim}
+\caption{The Pingus codebase directory tree.
+\label{tab.tree}
+}
+\end{figure*}
+
+%[cpp_uncompressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/CPP
+%[cpp_compressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/ALL
+%[ceu_compressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/ALL
 
@@
 
-We selected 9 representative game behaviors and describe their
+We selected 6 representative game behaviors and describe their
 implementations in C++ and \CEU.
@@
 
-\subsection{Finite State Machines \\ Case Study: The Armageddon Button Double Click}
-%\label{sec.patterns.fsm}
-
-State machines describe the behavior of entities by mapping event occurrences
-to transitions between states that trigger appropriate actions.
-
-%\subsection{The Armageddon Button Double Click}
-%\label{sec.patterns.fsm.armageddon}
-
-\begin{figure}
+\subsection{Finite State Machines}
+
+\textbf{Case Study: Detecting double-clicks in the \emph{Armageddon button}}
+
+\begin{figure}[t]
 \centering
@@
 
-In Pingus, a double click in the \emph{Armageddon} button at the bottom right
+In Pingus, a double click in the \emph{Armageddon button} at the bottom right
 of the screen literally explodes all pingus (Figure~\ref{fig.armageddon}).
-Figure~\ref{fig.armageddon.code} compares the implementations in C++ and \CEU.
-
-\begin{figure*}
+Figure~\ref{lst.armageddon} compares the implementations in C++ and \CEU.
+
+\begin{figure*}[t]
 \begin{minipage}[t]{0.55\linewidth}
@@
     <...>
-    emit outer.game.go_armageddon;
+    emit game.go_armageddon;
 end
@@
 %\rule{8.4cm}{0.37pt}
-\caption{ Shared-memory concurrency in \CEU:
-example \ax is safe because the trails access \code{x} atomically in different 
-reactions;
-example \bx is unsafe because both trails access \code{y} in the same reaction.
-\label{lst.shared}
+\caption{ Detecting double-clicks in the \emph{Armageddon button}.
+\label{lst.armageddon}
 }
@@
 
-\begin{figure}
+\begin{figure}[t]
 \centering
 \includegraphics[width=\columnwidth]{double-click}
-\caption{State machine for the \emph{Armageddon} double click.
+\caption{State machine for detecting double-clicks in the
+         \emph{Armageddon button}.
 \label{fig.armageddon.fsm}
@@
 
-\subsection{Continuation Passing \\ Case Study: Advancing Pages in the Story Screen}
-
-The completion of a long-lasting activity may carry a continuation, i.e., some
-action to execute next.
-
-\begin{figure}
+\textbf{Case Study: The \emph{Bomber action} sequence}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=\columnwidth]{bomber-anim}
+\caption{ The \emph{Bomber action}.
+\label{fig.bomber.anim}
+}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=\columnwidth]{bomber-state}
+\caption{ State machine for the *Bomber* animation.
+\label{fig.bomber.state}
+}
+\end{figure}
+
+The \emph{Bomber action} explodes the clicked pingu, throwing particles around
+and also destroying the terrain under its radius (Figure~\ref{fig.bomber.anim}).
+%
+We can model the explosion animation with a sequential state machine
+(Figure~\ref{fig.bomber.state}) with actions associated to specific frames as
+follows:
+%
+\begin{enumerate}
+\item 0th frame:  plays a "Oh no!" sound.
+\item 10th frame: plays a "Bomb!" sound.
+\item 13th frame: throws particles, destroys the terrain, and shows an
+                  explosion sprite.
+\item Game tick:  hides the explosion sprite.
+\item Last frame: kills the pingu.
+\end{enumerate}
+
+TODO
+%*([This video][youtube_bomber] presents the sound effects.)*
+%[youtube_bomber]: https://youtu.be/QLXIT59il6o?t=306
+
+\begin{figure*}[t]
+\begin{minipage}[t]{0.55\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+Bomber::Bomber (Pingu* p) :
+    <...>
+    sprite(<...>),              // bomber sprite
+    sound_played(false),        // tracks state 2               @def_1
+    particle_thrown(false),     // tracks state 3
+    colmap_exploded(false),     // tracks state 3
+    gfx_exploded(false)         // tracks state 4               @def_2
+{
+    <...>
+    // 1. 0th frame: plays a "Oh no!" sound.
+    get_world()->play_sound("ohno", pingu->get_pos());          @sound_ohno
+}
+
+void Bomber::update () {                                        @update_ini
+    sprite.update();                                            @update_1
+    <...>   // pingu movement                                   @update_2
+
+    // 2. 10th frame: plays a "Bomb!" sound.                    @sound_bomb_1
+    if (sprite.get_current_frame()==10 && !sound_played) {
+        sound_played = true;
+        get_world()->play_sound("plop", pingu->get_pos());
+    }                                                           @sound_bomb_2
+
+    // 3. 13th frame: throws particles, destroys the terrain, shows an explosion sprite
+    if (sprite.get_current_frame()==13 && !particle_thrown) {   @state_3_1
+        particle_thrown = true;
+        get_world()->get_pingu_particle_holder()->add_particle(...);
+    }
+    if (sprite.get_current_frame()==13 && !colmap_exploded) {
+        colmap_exploded = true;
+        get_world()->remove(bomber_radius, <...>);
+    }                                                           @state_3_2
+
+    // 5. Last frame: kills the Pingu
+    if (sprite.is_finished ()) {                                @die_1
+        pingu->set_status(Pingu::PS_DEAD);
+    }                                                           @die_2
+}                                                               @update_end
+
+void Bomber::draw (SceneContext& gc) {
+    // 3. 13th frame: throws particles, destroys the terrain, shows an explosion sprite
+    // 4. Game tick: hides the explosion sprite
+    if (sprite.get_current_frame()==13 && !gfx_exploded) {      @state_3_3
+        gfx_exploded = true;                                    @state_3_4
+        gc.color().draw(explo_surf, <...>);                     @state_4
+    }
+    gc.color().draw(sprite, pingu->get_pos());
+}
+\end{lstlisting}
+\centering\small{\ax Implementation in C++}
+\end{minipage}
+%
+\begin{minipage}[t]{0.45\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+code/await Bomber (void) -> _ActionName__Enum
+do
+    <...>
+    spawn Mover();                          // spawn the pingu movement to execute in the "background"      @move
+
+    var&? Sprite s = spawn Sprite(<...>);   // spawn the bomber animation to execute in the "background"    @sprite
+    watching s do                                                   @watch-1
+        // 1. 0th frame: plays a "Oh no!" sound.                    @anim_1
+        {Sound::PingusSound::play_sound("ohno", 0.5, 0.0)};
+
+        // 2. 10th frame: plays a "Bomb!" sound.
+        await outer.game.dt until s!.sprite.frame == 10;            @frame_1
+        {Sound::PingusSound::play_sound("plop", 0.5, 0.0)};
+
+        // 3. 13th frame: throws particles, destroys the terrain, shows an explosion sprite
+        await outer.game.dt until s!.sprite.frame == 13;            @frame_2
+        spawn PinguParticles(<...>) in outer.pingu_particles;       @particles
+        call Game_Remove({&bomber_radius}, <...>);
+        do                                                          @do
+            <...>
+            spawn Sprite(<...>);            // explosion            @explo
+
+            // 4. Game tick: hides the explosion sprite
+            await outer.game.dt;                                    @frame_3
+        end                                                         @end
+        await FOREVER;
+    end                                                             @watch-2
+
+    // 5. Last frame: kills the pingu
+    escape {ActionName::DEAD};                                      @anim_2
+end
+\end{lstlisting}
+\centering\small{\bx Implementation in \CEU}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ Advancing pages in the \emph{Story Screen}.
+\label{lst.story}
+}
+\end{figure*}
+
+In C++, the class \code{Bomber} defines the callbacks \code{draw} and
+\code{update} to manage the state machine described above:
+%
+The class first defines one state variable for each action to perform
+(ln. 4--7).
+The ``Oh no!'' sound plays as soon as the object starts in \emph{state-1} 
+(ln. 11).
+The \code{update} callback (ln. 14--38) updates the pingu animation and
+movement on every frame regardless of its current state (ln. 15--16).
+When the animation reaches the 10th frame, it plays the ``Bomb!'' sound and 
+switches to \emph{state-2} (ln. 18--22).
+The state variable \code{sound\_played} is required because the sprite frame
+doesn't necessarily advance on every \code{update} invocation (e.g.,
+\code{update} may execute twice during the 10th frame).
+The same reasoning and technique applies to \emph{state-3} (ln. 25--32 and
+43--44).
+The explosion sprite appears in a single frame in \emph{state-4} (ln. 45).
+Finally, the pingu dies after the animation frames terminate (ln. 35--37).
+%
+Note that a single numeric state variable suffices to track the states, but the
+original authors probably chose to encode each state in an independent boolean 
+variable to rearrange and experiment with them during development.
+Still, due to the short-lived nature of callbacks, state variables are 
+unavoidable and are actually the essence of object-oriented programming
+(i.e., *methods + mutable state*).
+%
+Like the C++ implementation for detecting double clicks, note also that the
+state machine is encoded across 3 different methods, each intermixing code with
+unrelated functionality.
+
+The equivalent code for the \emph{Bomber action} in \CEU doesn't require state
+variables and reflects the sequential state machine implicitly, using
+\code{await} statements in direct style to separate the actions.
+%
+The \code{Bomber} is a \code{code/await} abstraction of \CEU, which is similar
+to a coroutine~\cite{sync_async.cooperative}: a subroutine that retains runtime
+state, such as local variables and the program counter, across reactions to
+events (i.e., across \code{await} statements).
+The pingu movement and sprite animation are isolated in two other abstractions
+and execute in separate through the \code{spawn} primitive (ln. 4,6).
+The event \code{game.dt} (ln. 12,16,24) is analogous to the \code{update}
+callback of C++ and occurs on every frame.
+%
+The code tracks the animation instance (ln. 7--27), performing the last bomber
+action on termination (ln. 30).
+As soon as the animation starts, the code performs the first action (ln. 8).
+The intermediate actions are performed when the corresponding conditions occur
+(ln. 12,16,24).
+The \code{do-end} block (ln. 19--25), restricts the lifespan of the
+single-frame explosion sprite (ln. 21): after the next game tick (ln. 24), the
+block terminates and automatically destroys the spawned abstraction (removing
+it from the screen).
+%
+In contrast with the implementation in C++, all actions happen in a contiguous
+chunk of code (ln. 6--30) which handles no extra functionality.
+
+\subsection{Continuation Passing}
+
+\textbf{Case Study: Advancing Pages in the \emph{Story screen}}
+
+\begin{figure}[t]
 \centering
 \includegraphics[width=\columnwidth]{story-anim}
-\caption{The Story screen.
+\caption{The \emph{Story screen}.
 \label{fig.story}
@@
 
-\begin{figure*}
+\begin{figure*}[t]
 \begin{minipage}[t]{0.55\linewidth}
@@
     <...>
-    if (&lt;all-words-appearing&gt;) {
+    if (<all-words-appearing>) {
         displayed = true;
@@
 %\rule{8.4cm}{0.37pt}
-\caption{ Shared-memory concurrency in \CEU:
-example \ax is safe because the trails access \code{x} atomically in different 
-reactions;
-example \bx is unsafe because both trails access \code{y} in the same reaction.
-\label{lst.shared}
+\caption{ Advancing pages in the \emph{Story Screen}.
+\label{lst.story}
 }
@@
 
-\begin{figure}
+\begin{figure}[t]
 \centering
 \includegraphics[width=\columnwidth]{story}
-\caption{State machine for the Story screen.
+\caption{State machine for advancing pages in the \emph{Story screen}.
 \label{fig.story}
@@
 
-\subsection{Dispatching Hierarchies \\ Case Study: Bomber `draw` and `update` callbacks}
-
-Entities typically form a dispatching hierarchy in which a container that
-receives a stimulus automatically forwards it to its managed children.
-
-\begin{figure*}
+\subsection{Dispatching Hierarchies}
+
+\textbf{Case Study: Bomber `draw` and `update` callbacks}
+
+\begin{figure*}[t]
 \begin{minipage}[t]{0.50\linewidth}
@@
 end
+
+.
 \end{lstlisting}
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=\columnwidth]{story}
-\caption{State machine for the Story screen.
-\label{fig.story}
-}
-\end{figure}
-
-\begin{figure}
+\begin{figure}[t]
 \centering
@@
 
-\subsection{Lifespan Hierarchies \\ Case Study: The Pingus Container}
-
-Entities typically form a lifespan hierarchy in which a terminating
-container entity automatically destroys its managed children.
-
-\begin{figure*}
+\subsection{Lifespan Hierarchies}
+
+\textbf{Case Study: The Pingus Container}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=\columnwidth]{pingus_create_die-anim}
+\caption{Creation and death of pingus.
+\label{fig.pingus_create_die}
+}
+\end{figure}
+
+A pingu is a dynamic entity created periodically and destroyed under certain
+conditions, such as falling from a high altitude
+(Figure~\ref{pingus_create_die}).
+
+\begin{figure*}[t]
 \begin{minipage}[t]{0.50\linewidth}
@@
         (*pingu)->update();
-        if ((*pingu)->get_status() == Pingu::PS_DEAD) {
+        if ((*pingu)->get_status() == PS_DEAD) {
             pingu = pingus.erase(pingu);
@@
 }
+
+.
 \end{lstlisting}
@@
 %
-\begin{minipage}[t]{0.40\linewidth}
+\begin{minipage}[t]{0.50\linewidth}
 \begin{lstlisting}[numbers=left,xleftmargin=3em]
@@
         <...>
-        spawn Pingu(<...>) in outer.pingus;
+        spawn Pingu(<...>) in pingus;
     end
@@
 
-TODO
-
 code/await Pingu (<...>) do
@@
     loop do
-        await outer.game.dt;
-        if call Pingu_Rel_Getpixel(0,-1) == {Groundtype::GP_OUTOFSCREEN} then
+        await game.dt;
+        if Pingu_Is_Out_Of_Screen() then
             <...>
@@
 
-\begin{figure}
-\centering
-\includegraphics[width=\columnwidth]{story}
-\caption{State machine for the Story screen.
-\label{fig.story}
-}
-\end{figure}
-
 In C++, the class \code{PinguHolder} is a container that holds all pingus
@@
 go unnoticed but the program would keep consuming memory and CPU time.
-This problem is known as the \emph{lapsed listener}~\cite{TODO} and also occurs
-in languages with garbage collection:
+This problem is known as the \emph{lapsed listener}~\cite{games.patters} and
+also occurs in languages with garbage collection:
 A container typically holds a strong reference to a child (sometimes the only 
@@
 %
-The \code{escape} statement (ln. TODO) aborts the execution block of the
+The \code{escape} statement (ln. 17) aborts the execution block of the
 \code{Pingu} instance, removing it from its associated pool automatically.
@@
 
-\subsection{Signaling Mechanisms\\ Case Study: Global Keys and the Options Menu}
-
-Entities often need to communicate explicitly through signaling mechanisms,
-especially if there is no hierarchy relationship between them.
-
-\begin{figure}
+\subsection{Signaling Mechanisms}
+
+\textbf{Case Study: Global Keys and the Options Menu}
+
+\begin{figure}[t]
 \centering
@@
 
-\begin{figure}
+\begin{figure}[t]
 \centering
-\includegraphics[width=\columnwidth]{options-anim-opt}
+\includegraphics[width=\columnwidth]{events}
 \caption{Mutual dependency between signals.
@@
 
-\begin{figure*}
+\begin{figure*}[t]
 \begin{minipage}[t]{0.50\linewidth}
 \begin{lstlisting}[numbers=left,xleftmargin=3em]
-void GlobalEvent::on_button_press (<...>) {
+void GlbEvt::on_button_press (<...>) {
     <...>
     switch (event.keysym.sym) {
-        case SDLK_g:
-            if (keystate[SDLK_LCTRL] || keystate[SDLK_RCTRL]) {
-                config_manager.set_mouse_grab(          @ctrl_g_cpp
-                    !config_manager.get_mouse_grab());
+        case K_g:
+            if (keys[K_LCTRL] || keys[K_RCTRL]) {
+                cfgmgr.set_grab(
+                    !cfgmgr.get_grab());
             }
@@
 
-boost::signals2::signal<void(bool)> on_mouse_grab_change;   // definition in `config_manager.h` @signal_def
-
-void ConfigManager::set_mouse_grab (bool v) {   @set_mouse_grab
-    <...>
-    if (v != get_mouse_grab()) {                @if_1
+boost::signal<void(bool)> on_grab_change;
+
+void CfgMgr::set_grab (bool v) {
+    <...>
+    if (v != get_grab()) {
         <...>   // the actual "grab" effect
-        on_mouse_grab_change(v);                @signal
-    }                                           @if_2
+        on_grab_change(v);
+    }
 }
@@
 
-boost::signals2::signal<void (bool)> on_change;   // definition in `check_box.hpp`
-
-void CheckBox::set_state (bool is_on, bool send_signal) {   @last_argument
+boost::signal<void (bool)> on_change;
+
+void ChkBox::set (bool on, bool sndsig) {
     <...>   // switches the check box state
-    if (send_signal) {      @if_cb_1
-        on_change(is_on);
-    }                       @if_cb_2
+    if (sndsig) {
+        on_change(on);
+    }
 }
@@
 
-typedef std::vector&lt;boost::signals2::connection&gt; Connections;   // definition in `option_menu.hpp`
-Connections connections;                                        // definition in `option_menu.hpp`
+typedef std::vector<boost::connection> Conns;
+Conns conns;
 
 OptionMenu::OptionMenu() :
-    connections(),
-    mousegrab_box(),
+    conns(),
+    grab_box(),
     <...>
 {
-    mousegrab_box = new CheckBox(<...>);
-    connections.push_back(                              @bind_11
-        config_manager.on_mouse_grab_change.connect(
-            std::bind(&CheckBox::set_state, mousegrab_box, <...>, false); @bind_false
+    grab_box = new ChkBox(<...>);
+    conns.push_back(
+        cfgmgr.on_grab_change.connect(
+            std::bind(&ChkBox::set, grab_box, <...>, false);
         )
-    );                                                  @bind_12
-    connections.push_back(                              @bind_21
-        mousegrab_box->on_change.connect(
-            std::bind(&ConfigManager::set_mouse_grab, &config_manager, <...>);
+    );
+    conns.push_back(
+        grab_box->on_change.connect(
+            std::bind(&CfgMgr::set_grab, &cfgmgr, <...>);
         )
-    );                                                  @bind_22
-    <...>
-
-}
-
-OptionMenu::~OptionMenu() {     @destr_1
-    for (Connections::iterator i=connections.begin(); i!=connections.end(); ++i) {
+    );
+    <...>
+
+}
+
+OptionMenu::~OptionMenu() {
+    for (Conns::iterator i=conns.begin();
+                         i!=conns.end();
+                         ++i) {
         (*i).disconnect();
     }
-}                               @destr_2
+}
 
@@
     every e in SDL_KEYDOWN do
-        var _u8&& keystate = _SDL_GetKeyState(null);
-        <...>
-        if e:keysym.sym == {SDLK_g} then
-            if ((keystate[{SDLK_LCTRL}] as bool) or (keystate[{SDLK_RCTRL}] as bool)) then
-                emit config_manager.go_mouse_grab(    @ctrl_g_ceu
-                        not ({config_manager.get_mouse_grab()} as bool));
+        var _u8&& keys = _SDL_GetKeyState(null);
+        <...>
+        if e:keysym.sym == K_g then
+            if keys[K_LCTRL] or keys[K_RCTRL] then
+                emit cfgmgr.go_grab(
+                        not {cfgmgr.get_grab()});
             end
@@
 
-data ConfigManager with
-    event bool go_mouse_grab;
-end
-var ConfigManager config_manager = val ConfigManager(_);
+data CfgMgr with
+    event bool go_grab;
+end
+var CfgMgr cfgmgr = val CfgMgr(_);
 
@@
     var bool v;
-    every v in config_manager.go_mouse_grab do
+    every v in cfgmgr.go_grab do
         <...>   // the actual "grab" effect
@@
 
-data ICheckBox with
+data IChkBox with
     var   bool is_on;
@@
 
-code/await CheckBox (<...>) -> (var ICheckBox checkbox) -> FOREVER do
-    checkbox = val ICheckBox(<...>);
+code/await ChkBox (<...>) -> (var IChkBox chkbox) -> FOREVER do
+    chkbox = val IChkBox(<...>);
     <...>
     par do
-        every c.component.on_click do                   @every_1
-            emit checkbox.go_click(not checkbox.is_on); @dir_class_app
-        end                                             @every_2
+        every c.component.on_click do
+            emit chkbox.go_click(not chkbox.is_on);
+        end
     with
-        loop do                                         @loop_1
+        loop do
             <...>   // switches the check box state
-            checkbox.is_on = await checkbox.go_click;   @dir_app_class
-        end                                             @loop_2
+            chkbox.is_on = await chkbox.go_click;
+        end
     end
@@
     <...>
-    var& CheckBox b2 = <...>;
+    var& ChkBox b2 = <...>;
     spawn do
@@
             var bool v;
-            every v in outer.config_manager.go_mouse_grab do    @loop_11
-                emit b2.checkbox.go_click(v);
-            end                                                 @loop_12
+            every v in cfgmgr.go_grab do
+                emit b2.chkbox.go_click(v);
+            end
         with
             var bool v;
-            every v in b2.checkbox.go_click do                  @loop_21
-                emit outer.config_manager.go_mouse_grab(v);
-            end                                                 @loop_22
+            every v in b2.chkbox.go_click do
+                emit cfgmgr.go_grab(v);
+            end
         end
@@
 
-In C++, the class `GlobalEvent` detects *Ctrl-G*
-presses and invokes the callback `config\_manager.set\_mouse\_grab`:
-%
-The class `ConfigManager` uses a `boost::signal`
-[[![X]][boost\_signal]] to notify the application when the new configuration is
-applied:
-%
-The `if` enclosing the signal emission @NN(if\_1,-,if\_2) breaks the dependency 
+In C++, the class `GlbEvt` detects \emph{Ctrl-G} presses and invokes the
+callback \code{config\_manager.set\_grab} (ln. 5--8).
+%
+The class `CfgMgr` uses a \code{boost::signal} (ln. 16) to notify the
+application when the new configuration is applied (ln. 22).
+%
+The \code{if} enclosing the signal emission (ln. 20--23) breaks the dependency 
 cycle of Figure~\ref{fig.events} and prevents an infinite execution loop.
 %
-The class `CheckBox` also uses a `boost::signal` to 
-notify the application on changes:
-%
-Again, the `if` enclosing the signal emission @NN(if\_cb\_1,-,if\_cb\_2) breaks the 
+The class \code{ChkBox} also uses a \code{boost::signal} (ln. 28) to notify
+the application on changes (ln. 33).
+%
+Again, the \code{if} enclosing the signal emission (ln. 32--34) breaks the 
 dependency cycle of Figure~\ref{fig.events} to avoid infinite execution.
 %
-The class `OptionMenu` creates the dependency loop by
-connecting the two signals:
+The class \code{OptionMenu} creates the dependency loop by connecting the two
+signals.
 %
 The constructor binds
-the signal `config\_manager.on\_mouse\_grab\_change` to the callback method
-           `mousegrab\_box->set\_state`
-           @NN(bind\_11,-,bind\_12),
+the signal \code{config\_manager.on\_grab\_change} to the callback
+    method \code{grab\_box->set} (ln. 48--52),
 and also
-the signal `mousegrab\_box->on\_change` to the callback method
-           `config\_manager.set\_mouse\_grab`
-           @NN(bind\_21,-,bind\_22).
-This way, every time the `ConfigManager` signals `on\_mouse\_grab\_change`
-(`ConfigManager`, ln. @N(signal) [up](\#cpp\_config-manager)), `set\_state` is
-implicitly called.
-The same happens between the signal `on\_change` in the `CheckBox` and the 
-method `set\_mouse\_grab` in the `ConfigManager`
-(`ConfigManager`, ln. @N(set\_mouse\_grab) [up](\#cpp\_config-manager)).
-%
-Note that the signal binding to call `CheckBox::set\_state` @NN(bind\_false) 
-receives a fixed value `false` as the last argument to prevent infinite
-execution (`CheckBox`, ln. @N(last\_argument) [up](\#cpp\_check-box)).
-%
-The destructor @NN(destr\_1,-,destr\_2) breaks the connections explicitly when
-the *Option* screen terminates.
-
-In \CEU, a *Ctrl-G* key press broadcasts the internal event
-`config\_manager.go\_mouse\_grab` to the application [[![X]][ceu\_global\_event]]:
-%
-The configuration manager [[![X]][ceu\_config\_manager]] just needs to react to
-`go\_mouse\_grab` continuously to perform the *grab* effect:
-%
-The `CheckBox` [[![X]][ceu\_check\_box]] exposes the event `go\_click` for 
-notifications in both directions, i.e., from the abstraction to the application
-and *vice versa*:
-%
-The abstraction reacts to external clicks continuously @NN(every\_1,-,every\_2)
-to broadcast the event `go\_click` @NN(dir\_class\_app).
-It also reacts continuously to `go\_click` in another line of execution
-@NN(loop\_1,-,loop\_2), which awakes from notifications from the first line of
-execution or from the application.
-%
-The `OptionMenu` [[![X]][ceu\_option\_menu]] connects the two events as 
-follows:
+the signal \code{grab\_box->on\_change} to the callback method
+    \code{config\_manager.set\_grab} (ln. 53--57).
+This way, every time the \code{CfgMgr} signals
+\code{on\_grab\_change} (ln. 22), the method \code{set} is implicitly called.
+The same happens between the signal \code{on\_change} in the \code{ChkBox}
+and the method \code{set\_grab} in the \code{CfgMgr} (ln. 18).
+%
+Note that the signal binding to call \code{ChkBox::set} (ln. 50) receives a
+fixed value \code{false} as the last argument to prevent infinite execution
+(ln. 30).
+%
+The \code{OptionMenu} destructor (ln. 62--66) breaks the connections explicitly
+when the \emph{Option screen} terminates.
+
+In \CEU, a \emph{Ctrl-G} key press broadcasts the internal event
+\code{config\_manager.go\_grab} to the application (ln. 8).
+%
+The configuration manager just needs to react to \code{go\_grab}
+continuously to perform the *grab* effect (ln. 25--27).
+%
+The \code{ChkBox} exposes the event \code{go\_click} for notifications in
+both directions, i.e., from the abstraction to the application and
+\emph{vice versa}:
+%
+The abstraction reacts to external clicks continuously (ln. 41--43) to
+broadcast the event \code{go\_click} (ln. 47).
+It also reacts continuously to \code{go\_click} in another line of execution
+(ln. 45--48), which awakes from notifications from the first line of execution
+or from the application.
+%
+The \code{OptionMenu} connects the two events as follows:
 %
 The two loops in parallel handle the connections in opposite directions:
-from the configuration manager to the check box @NN(loop\_11,-,loop\_12);
+from the configuration manager to the check box (ln. 60--62);
 and
-from the check box to the configuration manager @NN(loop\_21,-,loop\_22).
-%
-When the *Option* screen terminates, the connections break automatically since
-the body is automatically aborted.
+from the check box to the configuration manager (ln. 65--67).
+%
+When the \emph{Option screen} terminates, the connections break automatically
+since the body with the two loops is automatically aborted.
 %
@@
 dependency cycle and prevent infinite execution.
-Due to the [stack-based execution for internal events][ceu\_stack] in \CEU,
-programs with mutually-dependent events do not create infinite execution loops.
+Due to the stack-based execution for internal events in
+\CEU~\cite{ceu.sensys13)}, programs with mutually-dependent events do not
+create infinite execution loops.
 
@@
 
-[Leonardo Kaplan](https://github.com/leokaplan/) and
-[Alexander Tkachov](https://github.com/Tkachov/)
-for early explorations and prototypes.
+We would like to thank Leonardo Kaplan and Alexander Tkachov for early
+explorations and prototypes of the game rewrite.
 

### 2017-05-19 paper.tex
...

@@
 
-Pingus%
-\footnote{Pingus: \url{http://pingus.seul.org/}}
-is an open-source clone of Lemmings%
-\footnote{Lemmings: \url{https://en.wikipedia.org/wiki/Lemmings_(video_game)}},
-a puzzle-platformer video game.      
-The objective of the game is to guide a group of penguins through a number of
-obstacles towards a designated exit%
-\footnote{Pingus gameplay: \url{https://www.youtube.com/watch?v=MKrJgIFtJX0}}.
-
 \begin{figure}[t]
@@
 \caption{Pingus gameplay.
-\label{fig.sweeney}
+\label{fig.pingus}
 }
@@
 
+Pingus is an open-source clone of Lemmings%
+\footnote{Lemmings: \url{https://en.wikipedia.org/wiki/Lemmings_(video_game)}},
+a puzzle-platformer video game (Figure~\ref{fig.pingus}).      
+The objective of the game is to guide a group of penguins through a number of
+obstacles towards a designated exit%
+\footnote{Pingus gameplay video: \url{https://www.youtube.com/watch?v=MKrJgIFtJX0}}.
+%
 Pingus is developed in standard object-oriented C++, the \emph{lingua franca}
 of game development \cite{games.patterns}.
-The codebase is about 40.000 lines of code (LoCs)%
-\footnote{Pingus repository: \url{https://github.com/Pingus/pingus/tree/7b255840c201d028fd6b19a2185ccf7df3a2cd6e/src}}, divided into
+The codebase%
+\footnote{Pingus repository: \url{https://github.com/Pingus/pingus/tree/7b255840c201d028fd6b19a2185ccf7df3a2cd6e/src}}
+is about 40.000 lines of code (LoCs), divided into
 the engine, level editor, auxiliary libraries, and the game logic itself.
@@
 game development resides in \emph{simulation} (aka \emph{game logic}), but
-which accounts for only 10\% of the CPU budget~\cite{games.sweeney}.
+which accounts for only 10\% of the CPU budget~\cite{games.sweeney}
+(Figure~\ref{fig.sweeney}).
 The game logic ``models the state of the game world as interacting objects
@@
 \section{The Pingus Codebase}
-
-In Pingus, the game logic also accounts for almost half the size of the
-codebase (18.173 from 39.362 LoCs, or 46\%).
-
-We rewrote 126 of the 272 files (46\%) which account for 9.186 of the 18.173
-LoCs (51\%) comprising the game logic of Pingus [[![X]][cpp\_uncompressed]].
-Half of the game logic relates to non-reactive code, such as configurations and
-options, saved games and serialization, maps and levels descriptions, string
-formatting, collision detection, graph algorithms, etc.
-This part remains unchanged and relies on the seamless integration between \CEU
-and C/C++ to remain usable.
-From the 9.186 touched LoCs, we removed all headers, declarations, trivial
-getters \& setters, and other innocuous statements, resulting in 70
-implementation files with 4.135 dense LoCs originally written in C++
-[[![X]][cpp\_compressed]].
-We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs
-[[![X]][ceu\_compressed]].
-Figure~\ref{tab.tree} summarizes the condensed codebase in the two
-implementations:
-
-This work focuses on a qualitative analysis for the programming techniques
-that we applied during the rewriting process.
-Not all techniques result in reduction in LoCs (especially considering the
-verbose syntax of \CEU), but have other effects such as eliminating shared
-variables and dependencies between classes.
-<!--, helping on encapsulation and cohesion.-->
-Nonetheless, the lowest ratio numbers above correlate to the parts of the game
-logic that we consider more susceptible to structured reactive programming.
-For instance, the *Pingu* behavior (*ratio 0.80*) contains complex animations
-that are affected by timers, game rules, and user interaction.
-In contrast, the *Option* screen (*ratio 0.97*) is a simple UI grid with
-trivial mouse interactions.
 
@@
 
-%[cpp_uncompressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/CPP
+In Pingus, the game logic also accounts for almost half the size of the
+codebase (18.173 from 39.362 LoCs, or 46\%).
+
+We rewrote 126 of the 272 files (46\%) which account for 9.186 of the 18.173
+LoCs (51\%) comprising the game logic of Pingus [[![X]][cpp\_uncompressed]].
+Half of the game logic relates to non-reactive code, such as configurations and
+options, saved games and serialization, maps and levels descriptions, string
+formatting, collision detection, graph algorithms, etc.
+This part remains unchanged and relies on the seamless integration between \CEU
+and C/C++ to remain usable.
+From the 9.186 touched LoCs, we removed all headers, declarations, trivial
+getters \& setters, and other innocuous statements, resulting in 70
+implementation files with 4.135 dense LoCs originally written in C++
+[[![X]][cpp\_compressed]].
+We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs
+[[![X]][ceu\_compressed]].
+Figure~\ref{tab.tree} summarizes the condensed codebase in the two
+implementations:
+
+\footnote{ Game logic code: \url{https://github.com/fsantanna/pingus/tree/ceu/cmp/CPP} }.
 %[cpp_compressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/ALL
@@
 
+This work focuses on a qualitative analysis for the programming techniques
+that we applied during the rewriting process.
+Not all techniques result in reduction in LoCs (especially considering the
+verbose syntax of \CEU), but have other effects such as eliminating shared
+variables and dependencies between classes.
+<!--, helping on encapsulation and cohesion.-->
+Nonetheless, the lowest ratio numbers above correlate to the parts of the game
+logic that we consider more susceptible to structured reactive programming.
+For instance, the \emph{Pingu} behavior (\emph{ratio 0.80}) contains complex
+animations that are affected by timers, game rules, and user interaction.
+In contrast, the \emph{Option screen} (\emph{ratio 0.97}) is a simple UI grid
+with trivial mouse interactions.
+
 \section{Control-Flow Patterns}
@@
 The rewriting process consisted of identifying sets of callbacks implementing
-*control-flow behaviors* in the game and translating them to \CEU using
+\emph{control-flow behaviors} in the game and translating them to \CEU using
 appropriate structured constructs.
@@
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=\columnwidth]{double-click-opt}
-\caption{Double click detection.
-\label{fig.armageddon}
-}
-\end{figure}
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=\columnwidth]{double-click-opt}
+%\caption{Double click detection.
+%\label{fig.armageddon}
+%}
+%\end{figure}
+%(Figure~\ref{fig.armageddon}).
 
 In Pingus, a double click in the \emph{Armageddon button} at the bottom right
-of the screen literally explodes all pingus (Figure~\ref{fig.armageddon}).
+of the screen literally explodes all pingus.%
+\footnote{Double click animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#1} }
 Figure~\ref{lst.armageddon} compares the implementations in C++ and \CEU.
@@
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=\columnwidth]{bomber-anim}
-\caption{ The \emph{Bomber action}.
-\label{fig.bomber.anim}
-}
-\end{figure}
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=\columnwidth]{bomber-anim}
+%\caption{ The \emph{Bomber action}.
+%\label{fig.bomber.anim}
+%}
+%\end{figure}
 
@@
 \includegraphics[width=\columnwidth]{bomber-state}
-\caption{ State machine for the *Bomber* animation.
-\label{fig.bomber.state}
+\caption{ State machine for the \emph{Bomber animation}.
+\label{fig.bomber}
 }
@@
 
+\begin{figure*}[th!]
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+Bomber::Bomber (Pingu* p) :
+  <...>
+  sprite(<...>),          // bomber sprite
+  sound_played(false),    // tracks state 2
+  particle_thrown(false), // tracks state 3
+  colmap_exploded(false), // tracks state 3
+  gfx_exploded(false)     // tracks state 4
+{
+  <...>
+  // 1. plays a "Oh no!" sound.
+  get_world()->play_sound("ohno");
+}
+
+void Bomber::update () {
+  sprite.update();
+  <...>   // pingu movement
+
+  // 2. plays a "Bomb!" sound.
+  if (sprite.frame()==10 && !sound_played) {
+    sound_played = true;
+    get_world()->play_sound("plop");
+  }
+
+  // 3. particles, terrain, explosion sprite
+  if (sprite.frame()==13 && !particle_thrown) {
+    particle_thrown = true;
+    get_world()->get_particles()->add(...);
+  }
+  if (sprite.frame()==13 && !colmap_exploded) {
+    colmap_exploded = true;
+    get_world()->remove(bomber_radius, <...>);
+  }
+
+  // 5. kills the Pingu
+  if (sprite.is_finished ()) {
+    pingu->set_status(PS_DEAD);
+  }
+}
+
+void Bomber::draw (SceneContext& gc) {
+  // 3. particles, terrain, explosion sprite
+  // 4. tick: hides the explosion sprite
+  if (sprite.frame()==13 && !gfx_exploded) {
+    gfx_exploded = true;
+    gc.color().draw(explo_surf, <...>);
+  }
+  gc.color().draw(sprite, pingu->get_pos());
+}
+\end{lstlisting}
+\centering\small{\ax Implementation in C++}
+\end{minipage}
+%
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+code/await Bomber (void) -> ActionName
+do
+  <...>
+  spawn Mover();  // movement in the background
+  var&? Sprite s = spawn Sprite(<...>);
+                  // animation in the background
+  watching s do
+    // 1. plays a "Oh no!" sound.
+    {Sound::PingusSound::play_sound("ohno")};
+
+    // 2. plays a "Bomb!" sound.
+    await game.dt until s!.sprite.frame == 10;
+    {Sound::PingusSound::play_sound("plop")};
+
+    // 3. particles, terrain, explosion sprite
+    await game.dt until s!.sprite.frame == 13;
+    spawn PinguParticles(<...>) in particles;
+    call Game_Remove({&bomber_radius}, <...>);
+    do
+      <...>
+      spawn Sprite(<...>);      // explosion
+
+      // 4. tick: hides the explosion sprite
+      await game.dt;
+    end
+    await FOREVER;
+  end
+
+  // 5. kills the pingu
+  escape DEAD;
+end
+
+.
+\end{lstlisting}
+\centering\small{\bx Implementation in \CEU}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ The \emph{Bomber action} sequence.
+\label{lst.bomber}
+}
+\end{figure*}
+
 The \emph{Bomber action} explodes the clicked pingu, throwing particles around
-and also destroying the terrain under its radius (Figure~\ref{fig.bomber.anim}).
+and also destroying the terrain under its radius.%
+\footnote{Bomber action animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#2} }
 %
 We can model the explosion animation with a sequential state machine
-(Figure~\ref{fig.bomber.state}) with actions associated to specific frames as
-follows:
+(Figure~\ref{fig.bomber}) with actions associated to specific frames as
+follows%
+\footnote{State machine animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#3} }:
 %
@@
 
-TODO
-%*([This video][youtube_bomber] presents the sound effects.)*
-%[youtube_bomber]: https://youtu.be/QLXIT59il6o?t=306
-
-\begin{figure*}[t]
-\begin{minipage}[t]{0.55\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
-Bomber::Bomber (Pingu* p) :
-    <...>
-    sprite(<...>),              // bomber sprite
-    sound_played(false),        // tracks state 2               @def_1
-    particle_thrown(false),     // tracks state 3
-    colmap_exploded(false),     // tracks state 3
-    gfx_exploded(false)         // tracks state 4               @def_2
-{
-    <...>
-    // 1. 0th frame: plays a "Oh no!" sound.
-    get_world()->play_sound("ohno", pingu->get_pos());          @sound_ohno
-}
-
-void Bomber::update () {                                        @update_ini
-    sprite.update();                                            @update_1
-    <...>   // pingu movement                                   @update_2
-
-    // 2. 10th frame: plays a "Bomb!" sound.                    @sound_bomb_1
-    if (sprite.get_current_frame()==10 && !sound_played) {
-        sound_played = true;
-        get_world()->play_sound("plop", pingu->get_pos());
-    }                                                           @sound_bomb_2
-
-    // 3. 13th frame: throws particles, destroys the terrain, shows an explosion sprite
-    if (sprite.get_current_frame()==13 && !particle_thrown) {   @state_3_1
-        particle_thrown = true;
-        get_world()->get_pingu_particle_holder()->add_particle(...);
-    }
-    if (sprite.get_current_frame()==13 && !colmap_exploded) {
-        colmap_exploded = true;
-        get_world()->remove(bomber_radius, <...>);
-    }                                                           @state_3_2
-
-    // 5. Last frame: kills the Pingu
-    if (sprite.is_finished ()) {                                @die_1
-        pingu->set_status(Pingu::PS_DEAD);
-    }                                                           @die_2
-}                                                               @update_end
-
-void Bomber::draw (SceneContext& gc) {
-    // 3. 13th frame: throws particles, destroys the terrain, shows an explosion sprite
-    // 4. Game tick: hides the explosion sprite
-    if (sprite.get_current_frame()==13 && !gfx_exploded) {      @state_3_3
-        gfx_exploded = true;                                    @state_3_4
-        gc.color().draw(explo_surf, <...>);                     @state_4
-    }
-    gc.color().draw(sprite, pingu->get_pos());
-}
-\end{lstlisting}
-\centering\small{\ax Implementation in C++}
-\end{minipage}
-%
-\begin{minipage}[t]{0.45\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
-code/await Bomber (void) -> _ActionName__Enum
-do
-    <...>
-    spawn Mover();                          // spawn the pingu movement to execute in the "background"      @move
-
-    var&? Sprite s = spawn Sprite(<...>);   // spawn the bomber animation to execute in the "background"    @sprite
-    watching s do                                                   @watch-1
-        // 1. 0th frame: plays a "Oh no!" sound.                    @anim_1
-        {Sound::PingusSound::play_sound("ohno", 0.5, 0.0)};
-
-        // 2. 10th frame: plays a "Bomb!" sound.
-        await outer.game.dt until s!.sprite.frame == 10;            @frame_1
-        {Sound::PingusSound::play_sound("plop", 0.5, 0.0)};
-
-        // 3. 13th frame: throws particles, destroys the terrain, shows an explosion sprite
-        await outer.game.dt until s!.sprite.frame == 13;            @frame_2
-        spawn PinguParticles(<...>) in outer.pingu_particles;       @particles
-        call Game_Remove({&bomber_radius}, <...>);
-        do                                                          @do
-            <...>
-            spawn Sprite(<...>);            // explosion            @explo
-
-            // 4. Game tick: hides the explosion sprite
-            await outer.game.dt;                                    @frame_3
-        end                                                         @end
-        await FOREVER;
-    end                                                             @watch-2
-
-    // 5. Last frame: kills the pingu
-    escape {ActionName::DEAD};                                      @anim_2
-end
-\end{lstlisting}
-\centering\small{\bx Implementation in \CEU}
-\end{minipage}
-%\rule{8.4cm}{0.37pt}
-\caption{ Advancing pages in the \emph{Story Screen}.
-\label{lst.story}
-}
-\end{figure*}
+Figure~\ref{lst.bomber} compares the implementations in C++ and \CEU.
 
@@
 unavoidable and are actually the essence of object-oriented programming
-(i.e., *methods + mutable state*).
+(i.e., \emph{methods + mutable state}).
 %
@@
 The pingu movement and sprite animation are isolated in two other abstractions
-and execute in separate through the \code{spawn} primitive (ln. 4,6).
+and execute in separate through the \code{spawn} primitive (ln. 4--5).
 The event \code{game.dt} (ln. 12,16,24) is analogous to the \code{update}
@@
 In contrast with the implementation in C++, all actions happen in a contiguous
-chunk of code (ln. 6--30) which handles no extra functionality.
+chunk of code (ln. 5--30) which handles no extra functionality.
 
@@
 \textbf{Case Study: Advancing Pages in the \emph{Story screen}}
+
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=\columnwidth]{story-anim}
+%\caption{The \emph{Story screen}.
+%\label{fig.story}
+%}
+%\end{figure}
 
@@
 \centering
-\includegraphics[width=\columnwidth]{story-anim}
-\caption{The \emph{Story screen}.
+\includegraphics[width=\columnwidth]{story}
+\caption{State machine for advancing pages in the \emph{Story screen}.
 \label{fig.story}
@@
 \end{figure}
-
-The clickable \emph{blue dots} in the campaign world map transit to ambience
-story screens (Figure~\ref{fig.story}).
-A story is composed of multiple pages and, inside each page, the words of the
-story appear incrementally over time.
-A first click in the button \code{>>>} fast forwards the words to show the full 
-page.
-A second click advances to the next page, until the story terminates.
-If the page completes before a click (due to the time elapsing), a first click 
-advances to the next page.
 
@@
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=\columnwidth]{story}
-\caption{State machine for advancing pages in the \emph{Story screen}.
-\label{fig.story}
-}
-\end{figure}
+The clickable \emph{blue dots} in the campaign world map transit to ambience
+story screens%
+\footnote{Story screen animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#4} }.
+A story is composed of multiple pages and, inside each page, the words of the
+story appear incrementally over time.
+A first click in the button \code{>>>} fast forwards the words to show the full 
+page.
+A second click advances to the next page, until the story terminates.
+If the page completes before a click (due to the time elapsing), a first click 
+advances to the next page.
+%
+Figure~\ref{lst.story} compares the implementations in C++ and \CEU.
 
@@
 
-\textbf{Case Study: Bomber `draw` and `update` callbacks}
+\textbf{Case Study: \emph{Bomber action} \code{draw} and \code{update} dispatching}
+
+TODO
+
+Figure~\ref{lst.hier} compares the implementations in C++ and \CEU.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=\columnwidth]{hierarchy}
+\caption{Dispatching chain for \code{update}.
+\label{fig.hier}
+}
+\end{figure}
 
@@
 %\rule{8.4cm}{0.37pt}
-\caption{ Shared-memory concurrency in \CEU:
-example \ax is safe because the trails access \code{x} atomically in different 
-reactions;
-example \bx is unsafe because both trails access \code{y} in the same reaction.
-\label{lst.shared}
+\caption{ \emph{Bomber action} \code{draw} and \code{update} dispatching.
+\label{lst.hier}
 }
@@
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=\columnwidth]{hierarchy}
-\caption{Dispatching chain for \code{update}.
-\label{fig.hier}
-}
-\end{figure}
-
 In C++, the class \code{Bomber} declares a \code{sprite} member to handle its
-animation frames (Figure~\ref{fig.TODO}).
+animation frames (Figure~\ref{fig.bomber}).
 %
@@
 
-\textbf{Case Study: The Pingus Container}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=\columnwidth]{pingus_create_die-anim}
-\caption{Creation and death of pingus.
-\label{fig.pingus_create_die}
-}
-\end{figure}
+\textbf{Case Study: Managing the Pingus Lifecycle}
+
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=\columnwidth]{pingus_create_die-anim}
+%\caption{Creation and death of pingus.
+%\label{fig.pingus_create_die}
+%}
+%\end{figure}
 
 A pingu is a dynamic entity created periodically and destroyed under certain
-conditions, such as falling from a high altitude
-(Figure~\ref{pingus_create_die}).
+conditions, such as falling from a high altitude%
+\footnote{Death of pingus animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#5} }.
+%
+Figure~\ref{lst.pingus} compares the implementations in C++ and \CEU.
 
@@
 %\rule{8.4cm}{0.37pt}
-\caption{ Shared-memory concurrency in \CEU:
-example \ax is safe because the trails access \code{x} atomically in different 
-reactions;
-example \bx is unsafe because both trails access \code{y} in the same reaction.
-\label{lst.shared}
+\caption{ Managing the pingus lifecycle.
+\label{lst.pingus}
 }
@@
 In \CEU, when the execution block of a dynamic instance terminates, which
-characterizes its *natural termination*, the instance is automatically removed
-from its pool.
+characterizes its \emph{natural termination}, the instance is automatically
+removed from its pool.
 Therefore, dynamic instances do not require any extra bookkeeping related to 
@@
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=\columnwidth]{options-anim-opt}
-\caption{The \emph{Mouse Grab} configuration option.
-\label{fig.options}
-}
-\end{figure}
-
-The *Mouse Grab* option restricts the mouse movement to the game window
-boundaries (Figure~\ref{fig.options}).
-The option can be set anywhere in the game by pressing *Ctrl-G*.
-In addition, the *Options* menu has a check box to toggle the *Mouse Grab*
-option with mouse clicks while still responding to *Ctrl-G* presses.
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=\columnwidth]{options-anim-opt}
+%\caption{The \emph{Mouse Grab} configuration option.
+%\label{fig.options}
+%}
+%\end{figure}
 
@@
 \end{figure}
-
-The implementations in C++ and \CEU use a signalling mechanism to connect the
-key presses, the check box, and a configuration manager that applies the
-appropriate side effects in the game (i.e., restrict the mouse movement).
-Figure~\ref{fig.events} illustrates how the mutual notifications create a 
-dependency cycle between the configuration manager and the check box.
 
@@
 %\rule{8.4cm}{0.37pt}
-\caption{ Shared-memory concurrency in \CEU:
-example \ax is safe because the trails access \code{x} atomically in different 
-reactions;
-example \bx is unsafe because both trails access \code{y} in the same reaction.
-\label{lst.shared}
+\caption{ TODO.
+\label{lst.grab}
 }
 \end{figure*}
+
+The \emph{mouse grab option} restricts the mouse movement to the game window
+boundaries%
+\footnote{Death of pingus animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#5} }.
+The option can be set anywhere in the game by pressing \emph{Ctrl-G}.
+In addition, the \emph{Options menu} has a check box to toggle the
+\emph{mouse grab option} with mouse clicks while still responding to
+\emph{Ctrl-G} presses.
+%
+Figure~\ref{lst.grab} compares the implementations in C++ and \CEU.
+
+The implementations in C++ and \CEU use a signalling mechanism to connect the
+key presses, the check box, and a configuration manager that applies the
+appropriate side effects in the game (i.e., restrict the mouse movement).
+Figure~\ref{fig.events} illustrates how the mutual notifications create a 
+dependency cycle between the configuration manager and the check box.
 
@@
 The configuration manager just needs to react to \code{go\_grab}
-continuously to perform the *grab* effect (ln. 25--27).
+continuously to perform the \emph{grab} effect (ln. 25--27).
 %

### 2017-05-19 paper.tex
...

@@
 
-\abstract{Abstract.
+\abstract{TODO.
 
@@
 
-\noindent \textbf{Keywords:} Radiosity, global illumination, constant time.
+\noindent \textbf{Keywords:} TODO, TODO, TODO.
 }
@@
 
-Pingus is an open-source clone of Lemmings%
-\footnote{Lemmings: \url{https://en.wikipedia.org/wiki/Lemmings_(video_game)}},
-a puzzle-platformer video game (Figure~\ref{fig.pingus}).      
+Pingus is an open-source puzzle-platform video game based on Lemmings.
 The objective of the game is to guide a group of penguins through a number of
-obstacles towards a designated exit%
-\footnote{Pingus gameplay video: \url{https://www.youtube.com/watch?v=MKrJgIFtJX0}}.
+obstacles towards a designated exit (Figure~\ref{fig.pingus}).
+% \footnote{Pingus gameplay video: \url{https://www.youtube.com/watch?v=MKrJgIFtJX0}}.
 %
@@
 (Figure~\ref{fig.sweeney}).
-The game logic ``models the state of the game world as interacting objects
-evolve over time''.
+%The game logic ``models the state of the game world as interacting objects
+%evolve over time''.
 The high development costs contrasting with the low impact on performance
@@
 
-Object-oriented games use the \emph{observer pattern}~\cite{games.patterns}
-in the game logic to handle events from the environment (e.g., key presses and
-timers) and also as a notification mechanism between game entities.
-%
-The observers are short-lived callbacks which must execute as fast as possible
-and in real time to keep the game reactive to incoming events.
+Object-oriented games rely on the \emph{observer pattern}~\cite{games.patterns}
+to handle events from the environment (e.g., key presses and timers) and also
+as a notification mechanism between entities in the game logic.
+%
+The observers are short-lived callbacks that have to execute as fast as
+possible to keep the game reactive to incoming events in real time.
 %
@@
 \footnote{``Callbacks as our Generations' Go To Statement'':
-\url{http://tirania.org/blog/archive/2013/Aug-15.html}}%
-\footnote{``Escape from Callback Hell'':
-\url{http://elm-lang.org/learn/Escape-from-Callback-Hell.elm}}
-
-\CEU~\cite{ceu.sensys13,ceu.mod15} is a programming language that aims to offer
-a concurrent and expressive alternative to C/C++ with the characteristics that
+\url{http://tirania.org/blog/archive/2013/Aug-15.html}}
+%\footnote{``Escape from Callback Hell'':
+%\url{http://elm-lang.org/learn/Escape-from-Callback-Hell.elm}}
+
+\CEU~\cite{ceu.sensys13,ceu.mod15} is a programming language that offers a
+concurrent and expressive alternative to C/C++ with the characteristics that
 follow:
@@
 terms of memory and CPU usage~\cite{ceu.sensys13}.
-The runtime is single threaded and the language requires no garbage
-collection.
+The runtime is single threaded and does not rely on garbage collection.
 
@@
 expressive and productive alternative for game logic development.
-In Pingus, the game logic also accounts for almost half the size of the
-codebase (18.173 from 39.362 LoCs, or 46\%).
-
-rewriting Pingus to \CEU is to
-
-Contributions:
-    - present SSRP as an effective alternative for game logic dev
-    - qualitative analysis
-        - patterns
-            - solutions
+We present a case study of rewriting the game logic of Pingus from C++ to \CEU
+with the contributions that follow:
+%
+\begin{itemize}
+\item Identification of five recurrent control-flow patterns in the game logic
+      that likely apply to other games.
+\item Idiomatic solutions in \CEU for six selected behaviors in the game logic.
+\item An in-depth qualitative analysis of the proposed solutions in comparison
+      to C++.
+\end{itemize}
 
@@
 In Pingus, the game logic also accounts for almost half the size of the
-codebase (18.173 from 39.362 LoCs, or 46\%).
-
-We rewrote 126 of the 272 files (46\%) which account for 9.186 of the 18.173
-LoCs (51\%) comprising the game logic of Pingus [[![X]][cpp\_uncompressed]].
-Half of the game logic relates to non-reactive code, such as configurations and
-options, saved games and serialization, maps and levels descriptions, string
-formatting, collision detection, graph algorithms, etc.
+codebase: 18.173 from 39.362 LoCs (46\%) spread in 272 files.
+%
+However, about half of the game logic relates to non-reactive code, such as
+configurations and options, saved games and serialization, maps and levels
+descriptions, string formatting, collision detection, graph algorithms, etc.
 This part remains unchanged and relies on the seamless integration between \CEU
-and C/C++ to remain usable.
-From the 9.186 touched LoCs, we removed all headers, declarations, trivial
-getters \& setters, and other innocuous statements, resulting in 70
-implementation files with 4.135 dense LoCs originally written in C++
-[[![X]][cpp\_compressed]].
-We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs
-[[![X]][ceu\_compressed]].
-Figure~\ref{tab.tree} summarizes the condensed codebase in the two
-implementations:
-
-\footnote{ Game logic code: \url{https://github.com/fsantanna/pingus/tree/ceu/cmp/CPP} }.
-%[cpp_compressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/ALL
-%[ceu_compressed]: https://github.com/fsantanna/pingus/tree/ceu/cmp/ALL
+and C/C++~\cite{ceu.sensys13}.
+%
+Therefore, we rewrote 9.186 LoCs spread in 126 files%
+\footnote{Game logic codebase: \url{https://github.com/sbgames-anon/paper-2017/tree/master/cpp/}}.
+%
+In order to only consider effective code in the analysis, we then removed all
+headers, declarations, trivial getters \& setters, and other innocuous
+statements, resulting in 70 implementation files with 4.135 dense LoCs
+originally written in C++%
+\footnote{``Effective'' C++ codebase: \url{https://github.com/sbgames-anon/paper-2017/tree/master/all/}}.
+We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs%
+\footnote{``Effective'' \CEU codebase: \url{https://github.com/sbgames-anon/paper-2017/tree/master/all/}}.
+%
+Figure~\ref{tab.tree} summarizes the effective codebase in the two
+implementations.
 
@@
 variables and dependencies between classes.
-<!--, helping on encapsulation and cohesion.-->
-Nonetheless, the lowest ratio numbers above correlate to the parts of the game
-logic that we consider more susceptible to structured reactive programming.
+%<!--, helping on encapsulation and cohesion.-->
+Nonetheless, the lines with lower ratio numbers above correlate to the parts of
+the game logic that we consider more susceptible to structured reactive
+programming.
 For instance, the \emph{Pingu} behavior (\emph{ratio 0.80}) contains complex
@@
 implementations in C++ and \CEU.
-We also categorized these examples in 5 abstract C++ control-flow patterns that
+We also categorize these behaviors in 5 abstract control-flow patterns that
 likely apply to other games:
@@
 \item \emph{Finite State Machines}:
-    State machines describe the behavior of entities by mapping event
-    occurrences to transitions between states that trigger appropriate actions.
+    Event occurrences map to transitions between states that trigger
+    appropriate actions comprising the behavior of a game entity.
 \item \emph{Continuation Passing}:
     The completion of a long-lasting activity may carry a continuation, i.e.,
-    some action to execute next.
+    some action to execute next in the game.
 \item \emph{Dispatching Hierarchies}:
@@
 boundaries%
-\footnote{Death of pingus animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#5} }.
+\footnote{Mouse grab animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#6} }.
 The option can be set anywhere in the game by pressing \emph{Ctrl-G}.
@@
 
+TODO: non reactive, C++ integration
+
 We promote the \emph{structured synchronous reactive} programming model of

### 2017-05-20 paper.tex
...

@@
 Therefore, we rewrote 9.186 LoCs spread in 126 files%
-\footnote{Game logic codebase: \url{https://github.com/sbgames-anon/paper-2017/tree/master/cpp/}}.
+\footnote{Game logic codebase: \url{https://github.com/an000/p/tree/master/cpp/}}.
 %
@@
 originally written in C++%
-\footnote{``Effective'' C++ codebase: \url{https://github.com/sbgames-anon/paper-2017/tree/master/all/}}.
+\footnote{``Effective'' C++ codebase: \url{https://github.com/an000/p/tree/master/all/}}.
 We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs%
-\footnote{``Effective'' \CEU codebase: \url{https://github.com/sbgames-anon/paper-2017/tree/master/all/}}.
+\footnote{``Effective'' \CEU codebase: \url{https://github.com/an000/p/tree/master/all/}}.
 %
@@
 of the screen literally explodes all pingus.%
-\footnote{Double click animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#1} }
+\footnote{Double click animation: \url{https://github.com/an000/p/blob/master/README.md#1} }
 Figure~\ref{lst.armageddon} compares the implementations in C++ and \CEU.
@@
 and also destroying the terrain under its radius.%
-\footnote{Bomber action animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#2} }
+\footnote{Bomber action animation: \url{https://github.com/an000/p/blob/master/README.md#2} }
 %
@@
 follows%
-\footnote{State machine animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#3} }:
+\footnote{State machine animation: \url{https://github.com/an000/p/blob/master/README.md#3} }:
 %
@@
 story screens%
-\footnote{Story screen animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#4} }.
+\footnote{Story screen animation: \url{https://github.com/an000/p/blob/master/README.md#4} }.
 A story is composed of multiple pages and, inside each page, the words of the
@@
 conditions, such as falling from a high altitude%
-\footnote{Death of pingus animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#5} }.
+\footnote{Death of pingus animation: \url{https://github.com/an000/p/blob/master/README.md#5} }.
 %
@@
 boundaries%
-\footnote{Mouse grab animation: \url{https://github.com/sbgames-anon/paper-2017/blob/master/README.md#6} }.
+\footnote{Mouse grab animation: \url{https://github.com/an000/p/blob/master/README.md#6} }.
 The option can be set anywhere in the game by pressing \emph{Ctrl-G}.

### 2017-05-22 paper.tex
...

@@
 obstacles towards a designated exit (Figure~\ref{fig.pingus}).
-% \footnote{Pingus gameplay video: \url{https://www.youtube.com/watch?v=MKrJgIFtJX0}}.
+% \footnote{Pingus gameplay video: \url{www.youtube.com/watch?v=MKrJgIFtJX0}}.
 %
@@
 The codebase%
-\footnote{Pingus repository: \url{https://github.com/Pingus/pingus/tree/7b255840c201d028fd6b19a2185ccf7df3a2cd6e/src}}
+\footnote{Pingus repository: \url{github.com/Pingus/pingus/}}
 is about 40.000 lines of code (LoCs), divided into
@@
 \footnote{``Callbacks as our Generations' Go To Statement'':
-\url{http://tirania.org/blog/archive/2013/Aug-15.html}}
+\url{tirania.org/blog/archive/2013/Aug-15.html}}
 %\footnote{``Escape from Callback Hell'':
-%\url{http://elm-lang.org/learn/Escape-from-Callback-Hell.elm}}
+%\url{elm-lang.org/learn/Escape-from-Callback-Hell.elm}}
 
@@
 expressive and productive alternative for game logic development.
-We present a case study of rewriting the game logic of Pingus from C++ to \CEU
-with the contributions that follow:
+We present a case study of rewriting Pingus from C++ to \CEU with the
+contributions as follows:
 %
 \begin{itemize}
-\item Identification of five recurrent control-flow patterns in the game logic
-      that likely apply to other games.
-\item Idiomatic solutions in \CEU for six selected behaviors in the game logic.
-\item An in-depth qualitative analysis of the proposed solutions in comparison
-      to C++.
+\item Identification of five recurrent control-flow patterns that likely apply
+      to other games.
+\item Idiomatic code in \CEU for six selected behaviors in the game logic.
+\item An in-depth qualitative analysis of the implementations in \CEU in
+      comparison to C++.
 \end{itemize}
 
+TODO: falar mais sobre as 3 contribuicoes:
+- control-flow patterns
+    - o que eh isso
+        - definir controle
+    - como eh diferente de GoF
+        - behavioral patterns
+- ideia geral das solucoes
+    - eliminar estados explicitos com estruturas de controle apropriadas
+- o que eh analise qualitativa
+    - como medimos
+
+The rest of the paper is organized as follows:
+Section~\ref{sec.codebase} gives an overview of the Pingus codebases in C++ and
+\CEU.
+Section~\ref{sec.patterns} describes the five control-flow patterns and
+discusses six case studies.
+Section~\ref{sec.related} discusses related work.
+Section~\ref{sec.conclusion} concludes the paper.
+
 \section{The Pingus Codebase}
+\label{sec.codebase}
 
@@
 In Pingus, the game logic also accounts for almost half the size of the
-codebase: 18.173 from 39.362 LoCs (46\%) spread in 272 files.
+codebase: 18.173 from 39.362 LoCs (46\%) spread across 272 files.
 %
@@
 %
-Therefore, we rewrote 9.186 LoCs spread in 126 files%
-\footnote{Game logic codebase: \url{https://github.com/an000/p/tree/master/cpp/}}.
+Therefore, we rewrote 9.186 LoCs spread across 126 files%
+\footnote{Complete codebase: \url{github.com/an000/p/tree/master/cpp}}.
 %
@@
 headers, declarations, trivial getters \& setters, and other innocuous
-statements, resulting in 70 implementation files with 4.135 dense LoCs
+statements, resulting 4.135 dense LoCs spread across 70 implementation files
 originally written in C++%
-\footnote{``Effective'' C++ codebase: \url{https://github.com/an000/p/tree/master/all/}}.
+\footnote{C++ codebase: \url{github.com/an000/p/tree/master/all}}.
 We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs%
-\footnote{``Effective'' \CEU codebase: \url{https://github.com/an000/p/tree/master/all/}}.
+\footnote{\CEU codebase: \url{github.com/an000/p/tree/master/all}}.
 %
@@
 
-\section{Control-Flow Patterns}
+\section{Control-Flow Patterns \& Case Studies}
+\label{sec.patterns}
+
+TODO: state pattern, FSM
+      strategy/visitor pattern, polymorhpism, hierarchies
+      observer pattern, signaling
 
@@
 \item \emph{Dispatching Hierarchies}:
-    Entities typically form a dispatching hierarchy in which a container that
-    receives a stimulus automatically forwards it to its managed children.
+    Entities form a dispatching hierarchy in which a container that receives a
+    stimulus automatically forwards it to its managed children.
 \item \emph{Lifespan Hierarchies}:
-    Entities typically form a lifespan hierarchy in which a terminating
-    container entity automatically destroys its managed children.
+    Entities form a lifespan hierarchy in which a terminating container entity
+    automatically destroys its managed children.
 \item \emph{Signaling Mechanisms}:
@@
 \subsection{Finite State Machines}
+
+TODO: falar do state pattern, citar game.patterns, falar que usa estado explicito
 
@@
 of the screen literally explodes all pingus.%
-\footnote{Double click animation: \url{https://github.com/an000/p/blob/master/README.md#1} }
+\footnote{Double click animation: \url{github.com/an000/p/blob/master/README.md#1} }
 Figure~\ref{lst.armageddon} compares the implementations in C++ and \CEU.
@@
 and also destroying the terrain under its radius.%
-\footnote{Bomber action animation: \url{https://github.com/an000/p/blob/master/README.md#2} }
+\footnote{Bomber action animation: \url{github.com/an000/p/blob/master/README.md#2} }
 %
@@
 follows%
-\footnote{State machine animation: \url{https://github.com/an000/p/blob/master/README.md#3} }:
+\footnote{State machine animation: \url{github.com/an000/p/blob/master/README.md#3} }:
 %
@@
 story screens%
-\footnote{Story screen animation: \url{https://github.com/an000/p/blob/master/README.md#4} }.
+\footnote{Story screen animation: \url{github.com/an000/p/blob/master/README.md#4} }.
 A story is composed of multiple pages and, inside each page, the words of the
@@
 conditions, such as falling from a high altitude%
-\footnote{Death of pingus animation: \url{https://github.com/an000/p/blob/master/README.md#5} }.
+\footnote{Death of pingus animation: \url{github.com/an000/p/blob/master/README.md#5} }.
 %
@@
 
-OptionMenu::OptionMenu() :
+OptMenu::OptMenu() :
   conns(),
@@
   conns.push_back(
-        cfgmgr.on_grab_change.connect(
-            std::bind(&ChkBox::set, grab_box, <...>, false);
+    cfgmgr.on_grab_change.connect( std::bind(
+      &ChkBox::set, grab_box, <...>, false) );
     )
@@
   conns.push_back(
-        grab_box->on_change.connect(
-            std::bind(&CfgMgr::set_grab, &cfgmgr, <...>);
+    grab_box->on_change.connect( std::bind(
+      &CfgMgr::set_grab, &cfgmgr, <...>) );
     )
@@
 
-OptionMenu::~OptionMenu() {
+OptMenu::~OptMenu() {
   for (Conns::iterator i=conns.begin();
@@
 
+.
 \end{lstlisting}
@@
 
-code/await ChkBox (<...>) -> (var IChkBox chkbox) -> FOREVER do
-    chkbox = val IChkBox(<...>);
+code/await ChkBox (<...>) -> (var IChkBox box) -> FOREVER do
+  box = val IChkBox(<...>);
   <...>
@@
     every c.component.on_click do
-            emit chkbox.go_click(not chkbox.is_on);
+      emit box.go_click(not box.is_on);
     end
@@
       <...>   // switches the check box state
-            chkbox.is_on = await chkbox.go_click;
+      box.is_on = await box.go_click;
     end
@@
 
-code/await OptionMenu <...> do
+code/await OptMenu <...> do
   <...>
@@
       every v in cfgmgr.go_grab do
-                emit b2.chkbox.go_click(v);
+        emit b2.box.go_click(v);
       end
@@
       var bool v;
-            every v in b2.chkbox.go_click do
+      every v in b2.box.go_click do
         emit cfgmgr.go_grab(v);
@@
 end
-
 \end{lstlisting}
@@
 boundaries%
-\footnote{Mouse grab animation: \url{https://github.com/an000/p/blob/master/README.md#6} }.
+\footnote{Mouse grab animation: \url{github.com/an000/p/blob/master/README.md#6} }.
 The option can be set anywhere in the game by pressing \emph{Ctrl-G}.
@@
 %
-The class \code{OptionMenu} creates the dependency loop by connecting the two
+The class \code{OptMenu} creates the dependency loop by connecting the two
 signals.
@@
 %
-The \code{OptionMenu} destructor (ln. 62--66) breaks the connections explicitly
+The \code{OptMenu} destructor (ln. 62--66) breaks the connections explicitly
 when the \emph{Option screen} terminates.
@@
 %
-The \code{OptionMenu} connects the two events as follows:
+The \code{OptMenu} connects the two events as follows:
 %
@@
 \section{Related Work}
+\label{sec.related}
+
+Some previous work~\cite{a,b,c} discuss how the GoF patterns~\cite{gof} apply
+to video game development.
+    - generally?
+        - mostram exemplos abstratos
+            - imagine um jogo...
+                - assim poderiamos aplicar...
+            - sem codigo real
+            - nao aplicam a jogos existentes
+
+- patterns
+    - sbgames
+    - gpp   
+- contol-flow patterns that are recurrent in Pingus and likely apply to other
+games
+- also, not identifying and applying, but applying in idiomatic code in another
+  language
+
+Design patterns for games can be approached from two dif-
+ferent perspectives; firstly as patterns used for describing
+the game mechanics (gameplay and game rules) and sec-
+ondly as the use of object-oriented design patterns in pro-
+gramming games.
+Concerning game mechanics, Bjork et al. [4] introduced
+a set of design patterns which essentially are descriptions
+(employing a unified vocabulary) of reoccurring interac-
+tion schemes relevant to game’s story and gameplay. As
+such, these patterns are not related to the software archi-
+tecture or code. The proposed patterns are collected from
+interviewing professional game programmers, from ana-
+lyzing existing games and from transforming game
+mechanics.
 
 \section{Conclusion}
+\label{sec.conclusion}
 
@@
 
-[goto]: http://tirania.org/blog/archive/2013/Aug-15.html
+[goto]: tirania.org/blog/archive/2013/Aug-15.html
 

### 2017-05-22 paper.tex
...

@@
 
+TODO: OO state + methods
+
 \CEU~\cite{ceu.sensys13,ceu.mod15} is a programming language that offers a
@@
 \footnote{Double click animation: \url{github.com/an000/p/blob/master/README.md#1} }
-Figure~\ref{lst.armageddon} compares the implementations in C++ and \CEU.
 
@@
     if (pressed) {
-        server->send_armageddon_event();
+        send_armageddon_event();
     } else {
@@
 do
-    var& RectComponent c = <...>;
+    var RectComponent c = <...>;
     <...>
@@
 
-In C++, the class \code{ArmageddonButton} implements methods for rendering the
-button and handling mouse and timer events.
-The listing focus on the double click detection, hiding unrelated parts with
+Figure~\ref{lst.armageddon}.a shows the C++ implementation for the class
+\code{ArmageddonButton} with methods for rendering the button and handling
+mouse and timer events.
+The code focus on the double click detection and hides unrelated parts with
 \code{<...>}.
@@
 atomically in reaction to external input events.
-The callback \code{on\_click} reacts to mouse clicks detected by the button
-base class \code{RectComponent} (ln. 2), while the callback \code{update}
-continuously reacts to the passage of time, frame by frame.
+The callback \code{on\_click} reacts to mouse clicks detected by the base class
+\code{RectComponent} (ln. 2), while the callback \code{update} continuously
+reacts to the passage of time, frame by frame.
 Callbacks are short lived because they must react to input as fast as possible
@@
 Also, even though the state variables are private, unrelated methods such as 
-\code{draw}`, which is defined in middle of the class (ln. 10--12), can
+\code{draw}, which is defined in middle of the class (ln. 10--12), can
 potentially access them.
@@
 %
-The loop detection (ln. 4--10) awaits the first click (ln. 5) and then, while
-watching 1 second (ln. 6--9), awaits the second click (ln. 7).
+In Figure~\ref{lst.armageddon}.b, the loop detection (ln. 4--10) awaits the
+first click (ln. 5) and then, while watching 1 second (ln. 6--9), awaits the
+second click (ln. 7).
 If the second click occurs within 1 second, the \code{break} terminates the
@@
 (ln. 12).
-Otherwise, the \code{watching} block as a whole aborts and restarts the loop, 
-falling back to the first click \code{await} (ln. 5).
+Otherwise, the \code{watching} block as a whole aborts after 1 second  and the
+loop restarts, falling back to the first click \code{await} (ln. 5).
 %
@@
 self-contained in the \code{loop} body (ln. 4--10).
-Furthermore, these 7 lines of code \emph{only} detect the double click, leaving
-the actual effect to happen outside the loop (ln. 12).
-
-\textbf{Case Study: The \emph{Bomber action} sequence}
+Also, these 7 lines of code \emph{only} detect the double click, leaving the
+actual effect to happen outside the loop (ln. 12).
+
+TODO: Structured programming, ...
+
+\textbf{Case Study: The \emph{Bomber action} animation sequence}
+
+TODO: descrever o que sao actions, exibir imagem da tela com menu de actions e
+seta apontando para pingu sendo clicado e com acao recem ativada
 
@@
 \includegraphics[width=\columnwidth]{bomber-state}
-\caption{ State machine for the \emph{Bomber animation}.
+\caption{ State machine for the \emph{Bomber animation} sequence.
 \label{fig.bomber}
@@
   // 1. plays a "Oh no!" sound.
-  get_world()->play_sound("ohno");
+  play_sound("ohno");
 }
@@
     sound_played = true;
-    get_world()->play_sound("plop");
+    play_sound("plop");
   }
@@
   spawn Mover();  // movement in the background
-  var&? Sprite s = spawn Sprite(<...>);
+  var Sprite s = spawn Sprite(<...>);
                   // animation in the background
@@
     // 1. plays a "Oh no!" sound.
-    {Sound::PingusSound::play_sound("ohno")};
+    {play_sound("ohno")};
 
     // 2. plays a "Bomb!" sound.
-    await game.dt until s!.sprite.frame == 10;
-    {Sound::PingusSound::play_sound("plop")};
+    await game.dt until s.sprite.frame == 10;
+    {play_sound("plop")};
 
     // 3. particles, terrain, explosion sprite
-    await game.dt until s!.sprite.frame == 13;
+    await game.dt until s.sprite.frame == 13;
     spawn PinguParticles(<...>) in particles;
@@
 In C++, the class \code{Bomber} defines the callbacks \code{draw} and
-\code{update} to manage the state machine described above:
+\code{update} to manage the state machine described above.
 %
@@
 (ln. 11).
-The \code{update} callback (ln. 14--38) updates the pingu animation and
+The \code{update} callback (ln. 14--38) first updates the pingu animation and
 movement on every frame regardless of its current state (ln. 15--16).
@@
 \code{update} may execute twice during the 10th frame).
-The same reasoning and technique applies to \emph{state-3} (ln. 25--32 and
+The same reasoning and technique applies to \emph{state-3} (ln. 24--32 and
 43--44).
 The explosion sprite appears in a single frame in \emph{state-4} (ln. 45).
-Finally, the pingu dies after the animation frames terminate (ln. 35--37).
+Finally, the pingu dies after the animation frames terminate (ln. 34--35).
 %
@@
 %
-Like the C++ implementation for detecting double clicks, note also that the
-state machine is encoded across 3 different methods, each intermixing code with
-unrelated functionality.
+Like double click detection in C++, note that the state machine is encoded
+across 3 different methods, each intermixing code with unrelated functionality.
 
@@
 The \code{Bomber} is a \code{code/await} abstraction of \CEU, which is similar
-to a coroutine~\cite{sync_async.cooperative}: a subroutine that retains runtime
-state, such as local variables and the program counter, across reactions to
-events (i.e., across \code{await} statements).
-The pingu movement and sprite animation are isolated in two other abstractions
-and execute in separate through the \code{spawn} primitive (ln. 4--5).
+to a coroutine or fiber~\cite{sync_async.cooperative}: a subroutine that
+retains runtime state, such as local variables and the program counter, across
+reactions to events (i.e., across \code{await} statements).
+The pingu movement and sprite animation are isolated in two other
+\code{code/await} abstractions and execute in separate through the \code{spawn}
+primitive (ln. 4--5).
 The event \code{game.dt} (ln. 12,16,24) is analogous to the \code{update}
@@
 %
-The code tracks the animation instance (ln. 7--27), performing the last bomber
-action on termination (ln. 30).
-As soon as the animation starts, the code performs the first action (ln. 8).
+The code tracks the animation aliveness (ln. 7--27) and, on termination,
+performs the last bomber action (ln. 30).
+As soon as the animation starts, the code performs the first action (ln. 9).
 The intermediate actions are performed when the corresponding conditions occur
@@
 \subsection{Continuation Passing}
+
+TODO: what is this?
 
@@
 
+TODO:
+    - FSMs
+    - BTs
+    - FRP
+
 \section{Conclusion}

### 2017-05-22 paper.tex
...

@@
 
-TODO: OO state + methods
-
 \CEU~\cite{ceu.sensys13,ceu.mod15} is a programming language that offers a
@@
 \begin{itemize}
-\item Identification of five recurrent control-flow patterns that likely apply
-      to other games.
-\item Idiomatic code in \CEU for six selected behaviors in the game logic.
-\item An in-depth qualitative analysis of the implementations in \CEU in
-      comparison to C++.
+\item Identifying five recurrent control-flow patterns that likely apply to
+      other games:
+        \emph{Finite State Machines},
+        \emph{Continuation Passing},
+        \emph{Dispatching Hierarchies},
+        \emph{Lifespan Hierarchies},
+        \emph{Signaling Mechanisms}.
+\item Applying idiomatic code in \CEU as alternative solutions for six selected
+      behaviors in the game logic.
+\item Presenting an in-depth qualitative analysis of the proposed solutions in
+      comparison to the original implementations in C++.
 \end{itemize}
-
-TODO: falar mais sobre as 3 contribuicoes:
-- control-flow patterns
-    - o que eh isso
-        - definir controle
-    - como eh diferente de GoF
-        - behavioral patterns
-- ideia geral das solucoes
-    - eliminar estados explicitos com estruturas de controle apropriadas
-- o que eh analise qualitativa
-    - como medimos
+%
+A control-flow pattern is a recurring technique to describe dependencies and
+explicit orders between statements (or groups of statements) in a program.
+For instance, consider how a key press stimulus propagates through the game
+entities and also what happens with them if the stimulus causes the end of the
+game.
+%
+\CEU supports primitives that help describing complex control-flow
+relationships in the game logic more concisely.
+%
+The rewriting process consisted of identifying sets of callbacks implementing
+control flow in the game and translating them to \CEU using appropriate
+structured constructs.
+%
+As an example, a double mouse click is characterized by a first click, followed
+by a maximum amount of time, followed by a second click.
+This behavior depends on different events (clicks and timers) which have to
+occur in a particular order.
+In C++, the implementation involves callbacks crossing reactions to successive
+events which manipulate state variables explicitly.
+%
+Although we documented in detail two use cases for each of the five patterns,
+in this work we only present six cases due to space constraints.
+%
+This work focuses on a qualitative analysis for the programming techniques
+that we applied during the rewriting process.
+Not all techniques result in reduction in LoCs (especially considering the
+verbose syntax of \CEU), but have other effects such as eliminating shared
+variables and dependencies between classes.
 
@@
 Section~\ref{sec.patterns} describes the five control-flow patterns and
-discusses six case studies.
+discusses six case studies in detail.
 Section~\ref{sec.related} discusses related work.
@@
 
+%%%
 This work focuses on a qualitative analysis for the programming techniques
@@
 with trivial mouse interactions.
+%%%
 
@@
 
-The rewriting process consisted of identifying sets of callbacks implementing
-\emph{control-flow behaviors} in the game and translating them to \CEU using
-appropriate structured constructs.
-As an example, a double mouse click is characterized by a first click, followed
-by a maximum amount of time, followed by a second click.
-This behavior depends on different events (clicks and timers) which have to
-occur in a particular order.
-In C++, the implementation involves callbacks crossing reactions to successive
-events which manipulate state variables explicitly.
+use cases first -> patterns later
 
@@
 \item \emph{Signaling Mechanisms}:
-    Entities often need to communicate explicitly through signaling mechanisms,
-    especially if there is no hierarchy relationship between them.
+    Entities often need to communicate and affect each other explicitly through
+    signaling mechanisms, especially if there is no hierarchy relationship
+    between them.
 \end{enumerate}
@@
 
+    - como eh diferente de GoF
+        - not much unlike behavioral patterns
+
 Design patterns for games can be approached from two dif-
@@
 TODO: non reactive, C++ integration
+- TODO: OO state + methods
+- eliminar estados explicitos com estruturas de controle apropriadas
 

### 2017-05-22 paper.tex
...

@@
 game development resides in \emph{simulation} (aka \emph{game logic}), but
-which accounts for only 10\% of the CPU budget~\cite{games.sweeney}
-(Figure~\ref{fig.sweeney}).
+which accounts for only 10\% of the CPU budget~\cite{games.sweeney}.
+%(Figure~\ref{fig.sweeney}).
 %The game logic ``models the state of the game world as interacting objects
@@
 
-\begin{figure}[t]
-\centering
-\includegraphics[width=\columnwidth]{sweeney}
-\caption{Three kinds of code~\cite{games.sweeney}.
-\label{fig.sweeney}
-}
-\end{figure}
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=\columnwidth]{sweeney}
+%\caption{Three kinds of code~\cite{games.sweeney}.
+%\label{fig.sweeney}
+%}
+%\end{figure}
 
@@
 
-%%%
-This work focuses on a qualitative analysis for the programming techniques
-that we applied during the rewriting process.
-Not all techniques result in reduction in LoCs (especially considering the
-verbose syntax of \CEU), but have other effects such as eliminating shared
-variables and dependencies between classes.
-%<!--, helping on encapsulation and cohesion.-->
-Nonetheless, the lines with lower ratio numbers above correlate to the parts of
-the game logic that we consider more susceptible to structured reactive
-programming.
+Although the sections that follow compare the codebases a qualitatively, the
+lines with lower ratio numbers above correlate to the parts of the game logic
+that we consider more susceptible to structured reactive programming.
 For instance, the \emph{Pingu} behavior (\emph{ratio 0.80}) contains complex
@@
 with trivial mouse interactions.
-%%%
 

### 2017-05-23 paper.tex
...

@@
 \begin{itemize}
+\item Applying idiomatic code in \CEU as alternative solutions for six selected
+      behaviors in the game logic.
+\item Presenting an in-depth qualitative analysis of the proposed solutions in
+      comparison to the original implementations in C++.
 \item Identifying five recurrent control-flow patterns that likely apply to
@@
         \emph{Signaling Mechanisms}.
-\item Applying idiomatic code in \CEU as alternative solutions for six selected
-      behaviors in the game logic.
-\item Presenting an in-depth qualitative analysis of the proposed solutions in
-      comparison to the original implementations in C++.
 \end{itemize}
@@
 Section~\ref{sec.codebase} gives an overview of the Pingus codebases in C++ and
-\CEU.
-Section~\ref{sec.patterns} describes the five control-flow patterns and
-discusses six case studies in detail.
+\CEU and describes our approach to identify and rewrite the control flow in the
+game.
+Section~\ref{sec.patterns} discusses six case studies in detail which are
+categorized in five control-flow patterns.
 Section~\ref{sec.related} discusses related work.
@@
 
-\section{Control-Flow Patterns \& Case Studies}
-\label{sec.patterns}
-
-TODO: state pattern, FSM
-      strategy/visitor pattern, polymorhpism, hierarchies
-      observer pattern, signaling
-
-use cases first -> patterns later
-
-We can identify control-flow behaviors in C++ by looking for class members with
-identifiers resembling verbs, statuses, and counters (e.g.,
+As a general rewriting rule, we could identify control-flow behaviors in C++ by
+looking for class members with identifiers resembling verbs, statuses, and
+counters (e.g.,
 \code{pressed},
@@
 
-We selected 6 representative game behaviors and describe their
-implementations in C++ and \CEU.
-We also categorize these behaviors in 5 abstract control-flow patterns that
-likely apply to other games:
+During the course of the rewriting process, we could identify more general
+control-flow patterns which likely apply to other games as well.
+
+\section{Control-Flow Patterns \& Case Studies}
+\label{sec.patterns}
+
+In this section, we select six representative game behaviors and describe in
+detail their implementations in C++ and \CEU.
+%
+We also categorize these behaviors in five abstract control-flow patterns as
+follows:
 
@@
 
+TODO: state pattern, FSM
+      strategy/visitor pattern, polymorhpism, hierarchies
+      observer pattern, signaling
+
 \subsection{Finite State Machines}
@@
 
-Some previous work~\cite{a,b,c} discuss how the GoF patterns~\cite{gof} apply
-to video game development.
+The control-flow patterns closely relate to the \emph{GoF} behavioral
+patterns~\cite{gof}, which some previous work discuss in the context of video
+game development~\cite{a,b,c}.
+%
+different
     - generally?
@@
             - nao aplicam a jogos existentes
+    - propose solutions
+        - eliminate
+for instance, the \emph{state pattern}, state machines, sequences
+trivial case of the bomber animation is a series of statements separated
+by await statements
+visitor, adopted in the language
 

### 2017-05-24 paper.tex
...

@@
 game.
-Section~\ref{sec.patterns} discusses six case studies in detail which are
+Section~\ref{sec.pats} discusses six case studies in detail which are
 categorized in five control-flow patterns.
@@
 \section{Control-Flow Patterns \& Case Studies}
-\label{sec.patterns}
+\label{sec.pats}
 
@@
 
-TODO: state pattern, FSM
-      strategy/visitor pattern, polymorhpism, hierarchies
-      observer pattern, signaling
-
 \subsection{Finite State Machines}
+\label{sec.pats.fsms}
 
@@
 \subsection{Continuation Passing}
+\label{sec.pats.cps}
 
@@
 \subsection{Dispatching Hierarchies}
+\label{sec.pats.dispatching}
 
@@
 \subsection{Lifespan Hierarchies}
+\label{sec.pats.lifespan}
 
@@
 go unnoticed but the program would keep consuming memory and CPU time.
-This problem is known as the \emph{lapsed listener}~\cite{games.patters} and
+This problem is known as the \emph{lapsed listener}~\cite{games.patterns} and
 also occurs in languages with garbage collection:
@@
 \subsection{Signaling Mechanisms}
+\label{sec.pats.signaling}
 
@@
 Due to the stack-based execution for internal events in
-\CEU~\cite{ceu.sensys13)}, programs with mutually-dependent events do not
+\CEU~\cite{ceu.sensys13}, programs with mutually-dependent events do not
 create infinite execution loops.
@@
 patterns~\cite{gof}, which some previous work discuss in the context of video
-game development~\cite{a,b,c}.
-%
+game development~\cite{games.patterns,games.gof.2015,games.gof.2007}.
+%
+The original game in C++ uses variations of the
+    \emph{state} (Sections~\ref{sec.pats.fsms}~and~\ref{sec.pats.cps}),
+    \emph{visitor} (Sections~\ref{sec.pats.dispatching}~and~\ref{sec.pats.lifespan}), and
+    \emph{observer} (Section~\ref{sec.pats.signaling} and to handle input in general)
+patterns as implementation details to achieve the desired higher-level
+control-flow patterns.
+%
+\CEU overcomes the need of behavioral patterns with a semantics that supports
+structured control-flow mechanisms and event-based communication via broadcast.
+%
+As an example, the \emph{state pattern} for the bomber animation in
+Section~\ref{sec.pats.fsms} becomes a series of blocks separated by
+\code{await} statements.
+
+A number of domain-specific languages, frameworks, and techniques have been
+proposed for particular subsystems of the game logic, such as
+animations~\cite{games.anims.2006,games.anims.2003,games.anims.1996,games.anims.1982},
+game state and screen progression~\cite{games.fsms.2006.1,games.fsms.2006.2}, and
+behavior and AI modeling~\cite{games.bts,games.bts.unreal}
+%
+In Pingus, we employed \CEU at the core of the game for event dispatching
+(Section~\ref{sec.pats.dispatching}) and memory management of entities
+(Section~\ref{sec.pats.lifespan}), eliminating parts of the original game
+engine.
+We also implemented all entity animations and behaviors
+(Section~\ref{sec.pats.fsms}), screen transitions
+(Section~\ref{sec.pats.cps}), and intermodule communication
+(Section~\ref{sec.pats.signaling}) using the available control mechanisms.
+%
+\CEU is a superset of C targeting reactive systems in general, not only games,
+and has also been adopted in other domains, such as
+    wireless sensor networks~\cite{ceu.sensys13,ceu.terra} and
+    multimedia systems~\cite{ceu.media.webmedia16}.
+
+Functional reactive programming is an
+FRP
+    - interfacing with C
+    - live porting
+
 different
@@
         - eliminate
-for instance, the \emph{state pattern}, state machines, sequences
-trivial case of the bomber animation is a series of statements separated
-by await statements
 visitor, adopted in the language

### 2017-05-24 paper.tex
...

@@
 descriptions, string formatting, collision detection, graph algorithms, etc.
-This part remains unchanged and relies on the seamless integration between \CEU
-and C/C++~\cite{ceu.sensys13}.
+This part remains unchanged and relies on the integration between \CEU and
+C/C++.
 %
@@
 form of control-flow progression that cross multiple callback invocations.
+
+We employed \emph{live code translation}, i.e., starting from the original
+codebase in C++, we reimplemented piece-by-piece without breaking the game
+compilation and execution.
+%
+This was only possible given the seamless integration between \CEU and
+C/C++~\cite{ceu.sensys13}: the type systems are equivalent and the integration
+happens at the source code level.
+This enables trivial sharing of control (calling C/C++ from \CEU and
+vice-versa) and data (accessing C/C++ data from \CEU and vice-versa).
 
@@
 
-Functional reactive programming is an
-FRP
-    - interfacing with C
-    - live porting
-
-different
-    - generally?
-        - mostram exemplos abstratos
-            - imagine um jogo...
-                - assim poderiamos aplicar...
-            - sem codigo real
-            - nao aplicam a jogos existentes
-    - propose solutions
-        - eliminate
-visitor, adopted in the language
-
-- patterns
-    - sbgames
-    - gpp   
-- contol-flow patterns that are recurrent in Pingus and likely apply to other
-games
-- also, not identifying and applying, but applying in idiomatic code in another
-  language
-
-    - como eh diferente de GoF
-        - not much unlike behavioral patterns
-
-Design patterns for games can be approached from two dif-
-ferent perspectives; firstly as patterns used for describing
-the game mechanics (gameplay and game rules) and sec-
-ondly as the use of object-oriented design patterns in pro-
-gramming games.
-Concerning game mechanics, Bjork et al. [4] introduced
-a set of design patterns which essentially are descriptions
-(employing a unified vocabulary) of reoccurring interac-
-tion schemes relevant to game’s story and gameplay. As
-such, these patterns are not related to the software archi-
-tecture or code. The proposed patterns are collected from
-interviewing professional game programmers, from ana-
-lyzing existing games and from transforming game
-mechanics.
-
-TODO:
-    - FSMs
-    - BTs
-    - FRP
+Functional reactive programming (FRP)~\cite{frp.fran} contrasts with
+Structured synchronous reactive programming as a complementary programming
+style for reactive applications.
+%
+We believe that FRP is more suitable for data-intensive applications, while 
+SSRP, for control-intensive applications.
+%
+On the one hand, FRP uses declarative formulas to specify continuous functions 
+over time, such as for physics or data constraints among entities.
+%
+On the other hand, describing a sequence of steps in FRP requires to encode 
+explicit state machines so that functions can switch behavior depending on the 
+current state.
+%
+FRP has been successfully used to implement a 3D first person shooting game
+from scratch, but with performance considerations~\cite{games.frag}.
+%
+We rewrote an existing game in small steps while keeping it working.
+Although we do not provide a performance evaluation (Pingus is not performance
+sensitive), previous work on \CEU shows that it is comparable to C in the
+context of embedded systems~\cite{ceu.sensys13}.
+Nonetheless, given the tight integration between with C/C++, critical parts of
+games can be preserved in C++ if needed.
 

### 2017-05-24 paper.tex
...

@@
 
-TODO: falar do state pattern, citar game.patterns, falar que usa estado explicito
-
-\textbf{Case Study: Detecting double-clicks in the \emph{Armageddon button}}
+\subsubsection{Case Study: Detecting Double-Clicks in the \emph{Armageddon Button}}
 
@@
 
-TODO: Structured programming, ...
-
-\textbf{Case Study: The \emph{Bomber action} animation sequence}
+\subsubsection{Case Study: The \emph{Bomber Action} Animation Sequence}
 
@@
 
+\subsubsection{Summary \& Uses in Pingus}
+
+The structured constructs of \CEU provide some advantages in comparison to 
+explicit state machines:
+%
+\begin{itemize}
+\item They encode all states with direct sequential code, eliminating shared
+      state variables.
+\item They handle all states (and only them) in the same contiguous block,
+      improving code encapsulation.
+\end{itemize}
+%
+Object-oriented games also adopt the \emph{state pattern} to model state
+machines, with subclasses describing each possible state~\cite{games.patterns}.
+However, this approach is not fundamentally different from using \code{switch}
+or \code{if} branches for each possible state.
+
+\begin{figure}[t]
+\begin{verbatim}
+Action       Ceu    C++   Explicit State
+---------   ----   ----   -----------------
+Bomber        23     50   4 state variables
+Bridger       75    100   2 state variables
+Drown          6     15   1 state variable
+Exiter         7     22   2 state variables
+Splashed       6     19   2 state variables
+\end{verbatim}
+\caption{Pingus actions in \CEU and C++.
+\label{tab.actions}
+}
+\end{figure}
+
+Pingus supports 16 actions in the game.
+As Figure~\ref{tab.actions} shows, 5 of them implement at least one state
+machine and are considerable smaller in \CEU in terms of LoCs when eliminating
+the state variables.
+%
+Considering the other 11 actions, the reduction in LoCs is negligible.
+This asymmetry in the implementation of actions illustrates the gains in
+expressiveness when describing state machines in direct style.
+
+%As other examples, detecting mouse dragging in the scenario and in the small
+%map to move the game viewport also involves state machines.
+%State machines also appear in the \emph{FPS counter} and in UI widgets with
+%visual feedback.
+
+Among the 65 implementation files in \CEU, we found 29 cases in 25 files using
+structured mechanisms to substitute states machines.
+They manifest as \code{await} statements in sequence or in aborting constructs
+such as \code{par/or} and \code{watching}.
+
 \subsection{Continuation Passing}
@@
 
-\textbf{Case Study: Advancing Pages in the \emph{Story screen}}
+\subsubsection{Case Study: Advancing Pages in the \emph{Story screen}}
 
@@
 
+\subsubsection{Summary \& Uses in Pingus}
+
+TODO: case 2
+
+The direct style of \CEU has some advantages in comparison to the 
+continuation-passing style:
+%
+\begin{itemize}
+\item It uses structured control flow (i.e., sequences and loops) instead of 
+      explicit data structures (e.g., stacks) or continuation variables.
+\item The activities are decoupled and do not hold references to one another.
+\item A single parent class describes the flow between the activities in a 
+      self-contained block of code.
+\end{itemize}
+
+Continuation passing typically controls the overall structure of the game,
+such as screen transitions in menus and level progressions.
+%
+\CEU uses the direct style techniques in 5 cases involving screen transitions:
+the main menu, the level menu, the level set menu, the world map loop, and
+the gameplay loop.
+%
+It also uses the same technique for the loop that switches the pingu actions
+during gameplay.
+
 \subsection{Dispatching Hierarchies}
@@
 
-\textbf{Case Study: \emph{Bomber action} \code{draw} and \code{update} dispatching}
+\subsubsection{Case Study: \emph{Bomber Action} \code{draw} and \code{update} Dispatching}
 
@@
 
+\subsubsection{Summary \& Uses in Pingus}
+
+TODO: case 2
+
+Passive entities subjected to hierarchies require a dispatching architecture
+that makes the reasoning about the program harder:
+
+\begin{itemize}
+\item The full dispatching chain may go through dozens of files.
+\item The dispatching chain may interleave between classes specific to the game
+      and also classes from the game engine (possibly third-party classes).
+\end{itemize}
+
+In C++, the update subsystem touches 39 files with around 100 lines of code
+just to forward \code{update} methods through the dispatching hierarchy.
+For the drawing subsystem, 50 files with around 300 lines of code.
+The implementation in C++ also relies on dispatching hierarchy for
+\code{resize} callbacks, touching 12 files with around 100 lines of code.
+%
+Most of this code is eliminated in \CEU since abstractions can react directly
+to the environment, not depending on hierarchies spread across multiple files.
+
+Note that dispatching hierarchies cross game engine code, suggesting that most
+games use this control-flow pattern heavily.
+In the case of the Pingus engine, we rewrote 9 files from C++ to \CEU, reducing
+them from 515 to 173 LoCs, mostly due to dispatching code removal.
+
 \subsection{Lifespan Hierarchies}
@@
 
-\textbf{Case Study: Managing the Pingus Lifecycle}
+\subsubsection{Case Study: Managing the Pingus Lifecycle}
 
@@
 
+\subsubsection{Summary \& Uses in Pingus}
+
+TODO: case 2
+
+Lexical lifespan for static instances and natural termination for dynamic
+instances provide some advantages in comparison to lifespan hierarchies through
+containers:
+
+\begin{itemize}
+\item Lexical scope makes an abstraction lifespan explicit in the source code.
+\item The memory for static instances is known at compile time.
+\item Natural termination makes an instance innocuous and, hence, susceptible
+      to immediate reclamation.
+\item Abstraction instances (static or dynamic) never require explicit
+      manipulation of pointers/references.
+\end{itemize}
+
+All entities in a game have an associated lifespan.
+
+The implementation in \CEU has over 200 static instantiations spread across all
+65 files.
+For dynamic entities, it defines 23 pools in 10 files, with almost 96
+instantiations across 37 files.
+Pools are used to hold explosion particles, levels and level sets from files,
+gameplay \& worldmap objects, and UI widgets.
+
 \subsection{Signaling Mechanisms}
@@
 
-\textbf{Case Study: Global Keys and the Options Menu}
+\subsubsection{Case Study: Global Keys and the Options Menu}
 
@@
 
+\subsubsection{Summary \& Uses in Pingus}
+
+TODO: case 2
+
+First-class internal events of \CEU provide some advantages in comparison to 
+Boost signals:
+
+\begin{itemize}
+\item They use the same convenient syntax of external events (e.g.,
+      \code{emit}, \code{await}, \code{every}, etc.).
+\item They never create infinite dependency loops.
+\item They do not require explicit unbinding.
+\end{itemize}
+
+The implementation in \CEU uses 39 events for internal communication defined in
+23 files with over 200 invocations of \code{emit} and \code{await} spread in
+over 50 files.
+
+Internal events are used for resizing entities, broadcasting update and draw
+requests, pausing parts of the game, triggering new pingu actions, signalling
+collisions, signalling UI interactions, among many others cases.
+
 \section{Related Work}

### 2017-05-24 paper.tex
...

@@
       comparison to the original implementations in C++.
-\item Identifying five recurrent control-flow patterns that likely apply to
+\item Identifying four recurrent control-flow patterns that likely apply to
       other games:
@@
         \emph{Continuation Passing},
-        \emph{Dispatching Hierarchies},
-        \emph{Lifespan Hierarchies},
-        \emph{Signaling Mechanisms}.
+        \emph{Dispatching Hierarchies}, and
+        \emph{Lifespan Hierarchies}.
+        %\emph{Signaling Mechanisms}.
 \end{itemize}
@@
 %
-Although we documented in detail two use cases for each of the five patterns,
-in this work we only present six cases due to space constraints.
+We only present six cases distributed in the four patterns due to space
+constraints, but we actually documented in detail a total eleven cases in six
+patterns (we omit the sixth pattern \emph{Signaling Mechanisms} entirely).
 %
@@
 Section~\ref{sec.pats} discusses six case studies in detail which are
-categorized in five control-flow patterns.
+categorized in four control-flow patterns.
 Section~\ref{sec.related} discusses related work.
@@
     automatically destroys its managed children.
-\item \emph{Signaling Mechanisms}:
-    Entities often need to communicate and affect each other explicitly through
-    signaling mechanisms, especially if there is no hierarchy relationship
-    between them.
+%\item \emph{Signaling Mechanisms}:
+    %Entities often need to communicate and affect each other explicitly through
+    %signaling mechanisms, especially if there is no hierarchy relationship
+    %between them.
 \end{enumerate}
@@
 \label{sec.pats.fsms}
+
+    Event occurrences map to transitions between states that trigger
+    appropriate actions comprising the behavior of a game entity.
 
@@
 
-TODO: descrever o que sao actions, exibir imagem da tela com menu de actions e
-seta apontando para pingu sendo clicado e com acao recem ativada
-
 %\begin{figure}[t]
@@
 \centering
+\includegraphics[width=120px]{bomber-01}
+\caption{ Assigning the \emph{Bomber action} to a pingu.
+\label{fig.bomber.action}
+}
+\end{figure}
+
+\begin{figure}[t]
+\centering
 \includegraphics[width=\columnwidth]{bomber-state}
 \caption{ State machine for the \emph{Bomber animation} sequence.
-\label{fig.bomber}
+\label{fig.bomber.fsm}
 }
@@
 
+In Pingus, the player may assign actions to specific pingus, as illustrated in
+Figure~\ref{fig.bomber}.
+%
 The \emph{Bomber action} explodes the clicked pingu, throwing particles around
@@
 We can model the explosion animation with a sequential state machine
-(Figure~\ref{fig.bomber}) with actions associated to specific frames as
+(Figure~\ref{fig.bomber.fsm}) with actions associated to specific frames as
 follows%
@@
 
-TODO: what is this?
+    The completion of a long-lasting activity may carry a continuation, i.e.,
+    some action to execute next in the game.
 
@@
 
-TODO: case 2
-
 The direct style of \CEU has some advantages in comparison to the 
@@
 %
-\CEU uses the direct style techniques in 5 cases involving screen transitions:
+\CEU uses the direct style techniques in five cases involving screen
+transitions:
 the main menu, the level menu, the level set menu, the world map loop, and
@@
 \label{sec.pats.dispatching}
+
+    Entities form a dispatching hierarchy in which a container that receives a
+    stimulus automatically forwards it to its managed children.
 
@@
 In C++, the class \code{Bomber} declares a \code{sprite} member to handle its
-animation frames (Figure~\ref{fig.bomber}).
+animation frames (Figure~\ref{fig.bomber.fsm}).
 %
@@
 
-TODO: case 2
-
 Passive entities subjected to hierarchies require a dispatching architecture
@@
 \label{sec.pats.lifespan}
+
+    Entities form a lifespan hierarchy in which a terminating container entity
+    automatically destroys its managed children.
 
@@
 
-\subsection{Signaling Mechanisms}
-\label{sec.pats.signaling}
-
-\subsubsection{Case Study: Global Keys and the Options Menu}
-
-%\begin{figure}[t]
-%\centering
-%\includegraphics[width=\columnwidth]{options-anim-opt}
-%\caption{The \emph{Mouse Grab} configuration option.
-%\label{fig.options}
-%}
-%\end{figure}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=\columnwidth]{events}
-\caption{Mutual dependency between signals.
-\label{fig.events}
-}
-\end{figure}
-
-\begin{figure*}[t]
-\begin{minipage}[t]{0.50\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
-void GlbEvt::on_button_press (<...>) {
-  <...>
-  switch (event.keysym.sym) {
-    case K_g:
-      if (keys[K_LCTRL] || keys[K_RCTRL]) {
-        cfgmgr.set_grab(
-          !cfgmgr.get_grab());
-      }
-      break;
-    <...>
-  }
-}
-
-///
-
-boost::signal<void(bool)> on_grab_change;
-
-void CfgMgr::set_grab (bool v) {
-  <...>
-  if (v != get_grab()) {
-    <...>   // the actual "grab" effect
-    on_grab_change(v);
-  }
-}
-
-///
-
-boost::signal<void (bool)> on_change;
-
-void ChkBox::set (bool on, bool sndsig) {
-  <...>   // switches the check box state
-  if (sndsig) {
-    on_change(on);
-  }
-}
-
-///
-
-typedef std::vector<boost::connection> Conns;
-Conns conns;
-
-OptMenu::OptMenu() :
-  conns(),
-  grab_box(),
-  <...>
-{
-  grab_box = new ChkBox(<...>);
-  conns.push_back(
-    cfgmgr.on_grab_change.connect( std::bind(
-      &ChkBox::set, grab_box, <...>, false) );
-    )
-  );
-  conns.push_back(
-    grab_box->on_change.connect( std::bind(
-      &CfgMgr::set_grab, &cfgmgr, <...>) );
-    )
-  );
-  <...>
-
-}
-
-OptMenu::~OptMenu() {
-  for (Conns::iterator i=conns.begin();
-             i!=conns.end();
-             ++i) {
-    (*i).disconnect();
-  }
-}
-
-.
-\end{lstlisting}
-\centering\small{\ax Implementation in C++}
-\end{minipage}
-%
-\begin{minipage}[t]{0.50\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
-spawn do
-  var _SDL_KeyboardEvent&& e;
-  every e in SDL_KEYDOWN do
-    var _u8&& keys = _SDL_GetKeyState(null);
-    <...>
-    if e:keysym.sym == K_g then
-      if keys[K_LCTRL] or keys[K_RCTRL] then
-        emit cfgmgr.go_grab(
-            not {cfgmgr.get_grab()});
-      end
-    end
-    <...>
-  end
-end
-
-///
-
-data CfgMgr with
-  event bool go_grab;
-end
-var CfgMgr cfgmgr = val CfgMgr(_);
-
-spawn do
-  var bool v;
-  every v in cfgmgr.go_grab do
-    <...>   // the actual "grab" effect
-  end
-end
-
-///
-
-data IChkBox with
-  var   bool is_on;
-  event bool go_click;
-end
-
-code/await ChkBox (<...>) -> (var IChkBox box) -> FOREVER do
-  box = val IChkBox(<...>);
-  <...>
-  par do
-    every c.component.on_click do
-      emit box.go_click(not box.is_on);
-    end
-  with
-    loop do
-      <...>   // switches the check box state
-      box.is_on = await box.go_click;
-    end
-  end
-end
-
-///
-
-code/await OptMenu <...> do
-  <...>
-  var& ChkBox b2 = <...>;
-  spawn do
-    par do
-      var bool v;
-      every v in cfgmgr.go_grab do
-        emit b2.box.go_click(v);
-      end
-    with
-      var bool v;
-      every v in b2.box.go_click do
-        emit cfgmgr.go_grab(v);
-      end
-    end
-  end
-  <...>
-end
-\end{lstlisting}
-\centering\small{\bx Implementation in \CEU}
-\end{minipage}
-%\rule{8.4cm}{0.37pt}
-\caption{ TODO.
-\label{lst.grab}
-}
-\end{figure*}
-
-The \emph{mouse grab option} restricts the mouse movement to the game window
-boundaries%
-\footnote{Mouse grab animation: \url{github.com/an000/p/blob/master/README.md#6} }.
-The option can be set anywhere in the game by pressing \emph{Ctrl-G}.
-In addition, the \emph{Options menu} has a check box to toggle the
-\emph{mouse grab option} with mouse clicks while still responding to
-\emph{Ctrl-G} presses.
-%
-Figure~\ref{lst.grab} compares the implementations in C++ and \CEU.
-
-The implementations in C++ and \CEU use a signalling mechanism to connect the
-key presses, the check box, and a configuration manager that applies the
-appropriate side effects in the game (i.e., restrict the mouse movement).
-Figure~\ref{fig.events} illustrates how the mutual notifications create a 
-dependency cycle between the configuration manager and the check box.
-
-In C++, the class `GlbEvt` detects \emph{Ctrl-G} presses and invokes the
-callback \code{config\_manager.set\_grab} (ln. 5--8).
-%
-The class `CfgMgr` uses a \code{boost::signal} (ln. 16) to notify the
-application when the new configuration is applied (ln. 22).
-%
-The \code{if} enclosing the signal emission (ln. 20--23) breaks the dependency 
-cycle of Figure~\ref{fig.events} and prevents an infinite execution loop.
-%
-The class \code{ChkBox} also uses a \code{boost::signal} (ln. 28) to notify
-the application on changes (ln. 33).
-%
-Again, the \code{if} enclosing the signal emission (ln. 32--34) breaks the 
-dependency cycle of Figure~\ref{fig.events} to avoid infinite execution.
-%
-The class \code{OptMenu} creates the dependency loop by connecting the two
-signals.
-%
-The constructor binds
-the signal \code{config\_manager.on\_grab\_change} to the callback
-    method \code{grab\_box->set} (ln. 48--52),
-and also
-the signal \code{grab\_box->on\_change} to the callback method
-    \code{config\_manager.set\_grab} (ln. 53--57).
-This way, every time the \code{CfgMgr} signals
-\code{on\_grab\_change} (ln. 22), the method \code{set} is implicitly called.
-The same happens between the signal \code{on\_change} in the \code{ChkBox}
-and the method \code{set\_grab} in the \code{CfgMgr} (ln. 18).
-%
-Note that the signal binding to call \code{ChkBox::set} (ln. 50) receives a
-fixed value \code{false} as the last argument to prevent infinite execution
-(ln. 30).
-%
-The \code{OptMenu} destructor (ln. 62--66) breaks the connections explicitly
-when the \emph{Option screen} terminates.
-
-In \CEU, a \emph{Ctrl-G} key press broadcasts the internal event
-\code{config\_manager.go\_grab} to the application (ln. 8).
-%
-The configuration manager just needs to react to \code{go\_grab}
-continuously to perform the \emph{grab} effect (ln. 25--27).
-%
-The \code{ChkBox} exposes the event \code{go\_click} for notifications in
-both directions, i.e., from the abstraction to the application and
-\emph{vice versa}:
-%
-The abstraction reacts to external clicks continuously (ln. 41--43) to
-broadcast the event \code{go\_click} (ln. 47).
-It also reacts continuously to \code{go\_click} in another line of execution
-(ln. 45--48), which awakes from notifications from the first line of execution
-or from the application.
-%
-The \code{OptMenu} connects the two events as follows:
-%
-The two loops in parallel handle the connections in opposite directions:
-from the configuration manager to the check box (ln. 60--62);
-and
-from the check box to the configuration manager (ln. 65--67).
-%
-When the \emph{Option screen} terminates, the connections break automatically
-since the body with the two loops is automatically aborted.
-%
-Note that the implementation in \CEU does not check event emits to break the
-dependency cycle and prevent infinite execution.
-Due to the stack-based execution for internal events in
-\CEU~\cite{ceu.sensys13}, programs with mutually-dependent events do not
-create infinite execution loops.
-
-\subsubsection{Summary \& Uses in Pingus}
-
-TODO: case 2
-
-First-class internal events of \CEU provide some advantages in comparison to 
-Boost signals:
-
-\begin{itemize}
-\item They use the same convenient syntax of external events (e.g.,
-      \code{emit}, \code{await}, \code{every}, etc.).
-\item They never create infinite dependency loops.
-\item They do not require explicit unbinding.
-\end{itemize}
-
-The implementation in \CEU uses 39 events for internal communication defined in
-23 files with over 200 invocations of \code{emit} and \code{await} spread in
-over 50 files.
-
-Internal events are used for resizing entities, broadcasting update and draw
-requests, pausing parts of the game, triggering new pingu actions, signalling
-collisions, signalling UI interactions, among many others cases.
-
 \section{Related Work}
@@
 \CEU for the development of games.
-We present in-depth use cases categorized in 5 control-flow patterns applied to
-\emph{Pingus} (an open-source \emph{Lemmings} clone) that likely apply to other
-games.
+We present in-depth use cases categorized in four control-flow patterns applied
+to \emph{Pingus} (an open-source \emph{Lemmings} clone) that likely apply to
+other games.
 

### 2017-05-24 paper.tex
...

@@
 We only present six cases distributed in the four patterns due to space
-constraints, but we actually documented in detail a total eleven cases in six
+constraints, but we actually documented in detail a total of eleven cases in six
 patterns (we omit the sixth pattern \emph{Signaling Mechanisms} entirely).
@@
 \subsubsection{Case Study: The \emph{Bomber Action} Animation Sequence}
+\label{sec.pats.fsms.2}
 
@@
 
+\subsubsection{Case Study: Game UI Widgets}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=\columnwidth]{game-session-arrows}
+\caption{UI children with static lifespan.
+\label{fig.ui}
+}
+\end{figure}
+
+\begin{figure*}[t]
+\begin{minipage}[t]{0.54\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+GameSession::GameSession(<...>) :
+    <...>
+{
+    <...>      // these widgets are always active...
+    btpanel  = new ButtonPanel(<...>);
+    pcounter = new PingusCounter(<...>);
+    smallmap = new SmallMap(<...>);
+    <...>
+    uimgr->add(btpanel);  // ...but are added
+    uimgr->add(pcounter); //  dynamically to the
+    uimgr->add(smallmap); //  dispatching hierarchy
+    <...>
+}
+\end{lstlisting}
+\centering\small{\ax Implementation in C++}
+\end{minipage}
+%
+\begin{minipage}[t]{0.46\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+code/await Game (void) do
+    <...> // other coexisting functionality
+    spawn ButtonPanel(<...>);
+    spawn PingusCounter(<...>);
+    spawn SmallMap(<...>);
+    <...> // other coexisting functionality
+end
+\end{lstlisting}
+\centering\small{\bx Implementation in \CEU}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ Managing the UI widgets lifecycle.
+\label{lst.pingus}
+}
+\end{figure*}
+
+Figure~\ref{fig.ui} shows the game UI widgets holding the buttons, pingus
+counter, and small map which coexist with the game screen during its whole
+lifespan.
+%
+Figure~\ref{lst.ui} compares the implementations in C++ and \CEU.
+
+In C++, the widgets are created in the constructor of the class
+\code{GameSession} (ln. 5--7), added to a UI container (ln. 9--11), and are
+never removed since they must always be visible.
+Arguably, to better express the intent of making them coexist with the game
+screen, the widgets could be declared as top-level automatic (non-dynamic)
+members.
+However, the class uses a container to automate \code{draw} and \code{update}
+dispatching to the widgets, as discussed in Section~\ref{sec.pats.dispatching}.
+In turn, the container method \code{add} expects dynamically allocated children
+only because they are automatically deallocated inside the container
+destructor.
+%
+The dynamic nature of containers in C++ demand extra caution from programmers:
+%
+\begin{itemize}
+\item When containers are part of a dispatching chain, it gets even harder to
+      know which objects are dispatched:
+      one has to ``simulate'' the program execution and track calls to
+      \code{add} and \code{remove}.
+\item For objects with dynamic lifespan, calls to \code{add} must always have
+      matching calls to \code{remove}:
+      missing calls to \code{remove} lead to memory and CPU leaks (to be
+      discussed as the \emph{lapsed listener} problem in
+      Section~\ref{sec.pats.lifespan.2}).
+\end{itemize}
+
+In Céu, entities that coexist just have to be created in the same lexical
+block.
+%
+Since abstractions can react independently, they do not require a dispatching
+container.
+%
+Lexical lifespan never requires containers, allocation and deallocation, or
+explicit references.
+In addition, all required memory is known at compile time, similarly to
+stack-allocated local variables.
+%
+The \emph{Bomber action} of Section~\ref{sec.pats.fsms.2} also relies on
+lexical scope to delimit the lifespan of the explosion sprite to a single
+frame.
+
 \subsubsection{Case Study: Managing the Pingus Lifecycle}
+\label{sec.pats.lifespan.2}
 
@@
 
-TODO: case 2
-
 Lexical lifespan for static instances and natural termination for dynamic

### 2017-05-24 paper.tex
...

@@
           bool, data, tag, escape, new, traverse,implementation,output,
-          native,@const,@pure,@safe,define},
+          native,@const,@pure,@safe,define,public,private},
     emphstyle={\bfseries},
@@
 \newcommand{\code}[1] {{\small{\texttt{#1}}}}
+\newcommand{\Code}[1] {{\texttt{#1}}}
 \newcommand{\ax}{\code{[a]}\xspace}
@@
     // 2. plays a "Bomb!" sound.
-    await game.dt until s.sprite.frame == 10;
+    await game.update until s.sprite.frame == 10;
     {play_sound("plop")};
@@
     // 3. particles, terrain, explosion sprite
-    await game.dt until s.sprite.frame == 13;
+    await game.update until s.sprite.frame == 13;
     spawn PinguParticles(<...>) in particles;
@@
       // 4. tick: hides the explosion sprite
-      await game.dt;
+      await game.update;
     end
@@
 
-Figure~\ref{lst.bomber} compares the implementations in C++ and \CEU.
-
-In C++, the class \code{Bomber} defines the callbacks \code{draw} and
-\code{update} to manage the state machine described above.
+In C++, the class \code{Bomber} in Figure~\ref{lst.bomber}.a defines the
+callbacks \code{draw} and \code{update} to manage the state machine described
+above.
 %
@@
 
-The equivalent code for the \emph{Bomber action} in \CEU doesn't require state
-variables and reflects the sequential state machine implicitly, using
-\code{await} statements in direct style to separate the actions.
+The equivalent code in \CEU for the \emph{Bomber action} in
+Figure~\ref{lst.bomber}.b doesn't require state variables and reflects the
+sequential state machine implicitly, using \code{await} statements in direct
+style to separate the actions.
 %
@@
 primitive (ln. 4--5).
-The event \code{game.dt} (ln. 12,16,24) is analogous to the \code{update}
+The event \code{game.update} (ln. 12,16,24) is analogous to the \code{update}
 callback of C++ and occurs on every frame.
@@
 %
-Figure~\ref{lst.story} compares the implementations in C++ and \CEU.
-
-In C++, the class \code{StoryScreenComponent} implements the method
-\code{next\_text}, which is a callback for clicks in \code{>>>}.
+
+In C++, the class \code{StoryScreenComponent} in Figure~\ref{lst.story}.a
+implements the method \code{next\_text}, which is a callback for clicks in
+\code{>>>}.
 %
@@
 
-The code in \CEU uses the internal event \code{next\_text}, which is emitted
-from clicks in \code{>>>}.
+The code in \CEU in Figure~\ref{lst.story}.b uses the internal event
+\code{next\_text}, which is emitted from clicks in \code{>>>}.
 %
@@
 
-\subsubsection{Case Study: \emph{Bomber Action} \code{draw} and \code{update} Dispatching}
-
-TODO
-
-Figure~\ref{lst.hier} compares the implementations in C++ and \CEU.
+\subsubsection{Case Study: \emph{Bomber Action} \Code{draw} and \Code{update} Dispatching}
 
@@
 
-In C++, the class \code{Bomber} declares a \code{sprite} member to handle its
-animation frames (Figure~\ref{fig.bomber.fsm}).
+In C++, the class \code{Bomber} presented in Figure~\ref{lst.hier}.a declares a
+\code{sprite} member to handle its animation frames
+(Figure~\ref{fig.bomber.fsm}).
 %
@@
 
-In \CEU, the \code{Bomber} action spawns a \code{Sprite} animation instance on
-its body.
-%
-The \code{Sprite} instance (ln. 3) can react directly to external \code{dt} and
-\code{redraw} events (which are analogous to \code{update} and \code{redraw}
-callbacks, respectively), bypassing the program hierarchy entirely.
+In \CEU, the \code{Bomber} abstraction presented in Figure~\ref{lst.hier}.b
+spawns a \code{Sprite} animation instance on its body.
+%
+The \code{Sprite} instance (ln. 3) can react directly to external \code{update}
+and \code{draw} events, bypassing the program hierarchy entirely.
 While and \emph{only while} the bomber abstraction is alive, the sprite
@@
 \caption{ Managing the UI widgets lifecycle.
-\label{lst.pingus}
+\label{lst.ui}
 }
@@
 
-Figure~\ref{fig.ui} shows the game UI widgets holding the buttons, pingus
-counter, and small map which coexist with the game screen during its whole
-lifespan.
-%
-Figure~\ref{lst.ui} compares the implementations in C++ and \CEU.
-
+Figure~\ref{fig.ui} shows the game UI widgets with action buttons, the pingus
+counter, and a small map which all coexist with the game screen during its
+whole lifespan.
+%
 In C++, the widgets are created in the constructor of the class
-\code{GameSession} (ln. 5--7), added to a UI container (ln. 9--11), and are
-never removed since they must always be visible.
+\code{GameSession} in Figure~\ref{lst.ui}.a (ln. 5--7), added to a UI container
+(ln. 9--11), and are never removed since they must always be visible.
 Arguably, to better express the intent of making them coexist with the game
-screen, the widgets could be declared as top-level automatic (non-dynamic)
-members.
+screen, the widgets could alternatively be declared as top-level automatic
+(non-dynamic) members.
 However, the class uses a container to automate \code{draw} and \code{update}
 dispatching to the widgets, as discussed in Section~\ref{sec.pats.dispatching}.
-In turn, the container method \code{add} expects dynamically allocated children
-only because they are automatically deallocated inside the container
+In turn, the container method \code{add} only expects dynamically allocated
+children because they are automatically deallocated inside the container
 destructor.
@@
 \item When containers are part of a dispatching chain, it gets even harder to
-      know which objects are dispatched:
+      know which objects are dispatched at a given moment:
       one has to ``simulate'' the program execution and track calls to
@@
 In Céu, entities that coexist just have to be created in the same lexical
-block.
+block in Figure~\ref{lst.ui}.b (ln. 3--5).
 %
@@
 %
-Figure~\ref{lst.pingus} compares the implementations in C++ and \CEU.
 
@@
     loop do
-        await game.dt;
+        await game.update;
         if Pingu_Is_Out_Of_Screen() then
@@
 
-In C++, the class \code{PinguHolder} is a container that holds all pingus
-alive.
+In C++, the class \code{PinguHolder} in Figure~\ref{lst.pingus}.a is a
+container that holds all pingus alive.
 %
@@
 collection with updates on every frame (ln. 11).
-Since the \code{redraw} behavior for a dead pingu is innocuous, the death could
+Since the \code{draw} behavior for a dead pingu is innocuous, the death could
 go unnoticed but the program would keep consuming memory and CPU time.
@@
 %
-The game screen spawns a new \code{Pingu} on every invocation of
-\code{Pingu\_Spawn}.
+The game screen in Figure~\ref{lst.pingus}.b spawns a new \code{Pingu} on every
+invocation of \code{Pingu\_Spawn} (ln. 4--7).
 %

### 2017-05-26 paper.tex
...

@@
 %
-For this reason, callbacks cannot contain long-lasting locals and loops, which
-are elementary capabilities of classical structured
+For this reason, callbacks cannot use long-lasting locals and loops, which are
+elementary capabilities of classical structured
 programming~\cite{rp.deprecating,rp.rescala,sync_async.cooperative}.
@@
 
-\CEU~\cite{ceu.sensys13,ceu.mod15} is a programming language that offers a
-concurrent and expressive alternative to C/C++ with the characteristics that
-follow:
+\CEU~\cite{ceu.sensys13,ceu.mod15} is a C/C++ source compatible programming
+language with the following characteristics:
 %
@@
 \item \emph{Reactive}: code only executes in reactions to events.
-\item \emph{Structured}: programs use structured control mechanisms, such as
-      \code{await} (to suspend a line of execution), and \code{par} (to combine
-      multiple lines of execution).
+\item \emph{Structured}: programs use structured control-flow mechanisms, such
+      as \code{await} (to suspend a line of execution), and \code{par} (to
+      combine multiple lines of execution).
 \item \emph{Synchronous}: reactions run atomically and to completion on each
-      line of execution, i.e., there's no implicit preemption or real
-      parallelism.
+      line of execution.
+      There's no implicit preemption or real parallelism, resulting in
+      deterministic execution.
 \end{itemize}
@@
 imposed by the observer pattern~\cite{rp.deprecating}.
+%
+\CEU provides primitives that help describing complex control-flow
+relationships in the game logic more concisely.
 %
@@
 \begin{itemize}
-\item Applying idiomatic code in \CEU as alternative solutions for six selected
-      behaviors in the game logic.
-\item Presenting an in-depth qualitative analysis of the proposed solutions in
-      comparison to the original implementations in C++.
-\item Identifying four recurrent control-flow patterns that likely apply to
-      other games:
+\item Alternative solutions in idiomatic \CEU with gains in productivity for
+      six selected behaviors in the game logic.
+\item An in-depth qualitative analysis of the proposed solutions in comparison
+      to the original implementations in C++.
+\item The identification of four recurrent control-flow patterns that likely
+      apply to other games:
         \emph{Finite State Machines},
@@
 explicit orders between statements (or groups of statements) in a program.
-For instance, consider how a key press stimulus propagates through the game
-entities and also what happens with them if the stimulus causes the end of the
-game.
-%
-\CEU supports primitives that help describing complex control-flow
-relationships in the game logic more concisely.
+%For instance, consider how a key press stimulus propagates through the game
+%entities and also what happens with them if the stimulus causes the end of the
+%game.
 %
@@
 %
-We only present six cases distributed in the four patterns due to space
-constraints, but we actually documented in detail a total of eleven cases in six
-patterns (we omit the sixth pattern \emph{Signaling Mechanisms} entirely).
-%
-This work focuses on a qualitative analysis for the programming techniques
-that we applied during the rewriting process.
+We focus on a qualitative analysis for the programming techniques that we
+applied during the rewriting process.
 Not all techniques result in reduction in LoCs (especially considering the
@@
 However, about half of the game logic relates to non-reactive code, such as
-configurations and options, saved games and serialization, maps and levels
-descriptions, string formatting, collision detection, graph algorithms, etc.
+dealing with configurations and options, saved games and serialization, maps
+and levels descriptions, string formatting, collision detection, graph
+algorithms, etc.
 This part remains unchanged and relies on the integration between \CEU and
@@
 
-Although the sections that follow compare the codebases a qualitatively, the
-lines with lower ratio numbers above correlate to the parts of the game logic
-that we consider more susceptible to structured reactive programming.
+Although the sections that follow compare the codebases qualitatively, the
+lines with lower ratio numbers in Figure~\ref{tab.tree} correlate to the parts
+of the game logic that we consider more susceptible to structured reactive
+programming.
 For instance, the \emph{Pingu} behavior (\emph{ratio 0.80}) contains complex
@@
 happens at the source code level.
-This enables trivial sharing of control (calling C/C++ from \CEU and
-vice-versa) and data (accessing C/C++ data from \CEU and vice-versa).
-
-During the course of the rewriting process, we could identify more general
-control-flow patterns which likely apply to other games as well.
+This enables trivial sharing of control and data, i.e., accessing C/C++ data
+and calling C/C++ from \CEU and vice-versa.
 
@@
 
-In this section, we select six representative game behaviors and describe in
-detail their implementations in C++ and \CEU.
-%
-We also categorize these behaviors in five abstract control-flow patterns as
-follows:
+During the course of the rewriting process, we have identified four abstract
+control-flow patterns which likely apply to other games as well:
 
@@
 \item \emph{Finite State Machines}:
-    Event occurrences map to transitions between states that trigger
-    appropriate actions comprising the behavior of a game entity.
+    Event occurrences lead to transitions between states and trigger actions
+    comprising the behavior of a game entity.
 \item \emph{Continuation Passing}:
-    The completion of a long-lasting activity may carry a continuation, i.e.,
-    some action to execute next in the game.
+    The completion of a long-lasting activity may carry a continuation in the
+    game, i.e., some action to execute next.
 \item \emph{Dispatching Hierarchies}:
@@
 
+We describe six representative game behaviors in detail distributed in the four
+patterns, with their implementations in C++ and \CEU.
+%
+Due to space constraints, we omit five other cases and also a fifth pattern
+\emph{Signaling Mechanisms} entirely.
+
 \subsection{Finite State Machines}
@@
 
-    Event occurrences map to transitions between states that trigger
-    appropriate actions comprising the behavior of a game entity.
+    Event occurrences lead to transitions between states and trigger actions
+    comprising the behavior of a game entity.
 
@@
 do
-    var RectComponent c = <...>;
+    var RectComponent r = <...>;
     <...>
     loop do
-        await c.component.on_click;
+        await r.on_click;
         watching 1s do
-            await c.component.on_click;
+            await r.on_click;
             break;
@@
     <...>
-    emit game.go_armageddon;
+    emit game.armageddon;
 end
@@
 reacts to the passage of time, frame by frame.
-Callbacks are short lived because they must react to input as fast as possible
-to let other callbacks execute, keeping the game with real-time responsiveness.
-%
-The class first initializes the variable \code{pressed} to track the first
-click (ln. 3,32).
-It also initializes the variable \code{press\_time} to count the time since the
-first click (ln. 4, 17).
+%Callbacks are short lived because they must react to input as fast as possible
+%to let other callbacks execute, keeping the game with real-time responsiveness.
+%
+The class first initializes the variable \code{pressed} (ln. 3) to track the
+first click (ln. 32).
+It also initializes the variable \code{press\_time} (ln. 4) to count the time
+since the first click (ln. 16--17).
 If another click occurs within 1 second, the class signals the double click to
-the application (ln. 30).
+the application (ln. 29--30).
 Otherwise, the \code{pressed} and \code{press\_time} state variables are reset
-(ln. 19--20). 
+(ln. 18--21). 
 %
@@
 %
-In Figure~\ref{lst.armageddon}.b, the loop detection (ln. 4--10) awaits the
-first click (ln. 5) and then, while watching 1 second (ln. 6--9), awaits the
-second click (ln. 7).
+In Figure~\ref{lst.armageddon}.b, the loop to detect double clicks (ln. 4--10)
+awaits the first click (ln. 5) and then, while watching 1 second (ln. 6--9),
+awaits the second click (ln. 7).
 If the second click occurs within 1 second, the \code{break} terminates the
-loop (ln. 8) and the \code{emit} signals the double click to the application
-(ln. 12).
+loop (ln. 8) and the \code{emit} in sequence signals the double click to the
+application (ln. 12).
 Otherwise, the \code{watching} block as a whole aborts after 1 second  and the
@@
 Double click detection in \CEU doesn't require state variables and is entirely
-self-contained in the \code{loop} body (ln. 4--10).
+self-contained in the \code{loop} body.
 Also, these 7 lines of code \emph{only} detect the double click, leaving the
-actual effect to happen outside the loop (ln. 12).
+actual effect to happen outside the loop (ln. 12) as well as all unrelated
+code such as redrawing the button.
 
@@
 In Pingus, the player may assign actions to specific pingus, as illustrated in
-Figure~\ref{fig.bomber}.
+Figure~\ref{fig.bomber.action}.
 %
@@
 We can model the explosion animation with a sequential state machine
-(Figure~\ref{fig.bomber.fsm}) with actions associated to specific frames as
+(Figure~\ref{fig.bomber.fsm}) with effects associated to specific frames as
 follows%
@@
 %
-The class first defines one state variable for each action to perform
+The class first defines one state variable for each effect to perform
 (ln. 4--7).
@@
 The explosion sprite appears in a single frame in \emph{state-4} (ln. 45).
-Finally, the pingu dies after the animation frames terminate (ln. 34--35).
+Finally, the pingu dies after the animation frames terminate (ln. 34--37).
 %
@@
 unavoidable and are actually the essence of object-oriented programming
-(i.e., \emph{methods + mutable state}).
+(i.e., \emph{methods with mutable state}).
 %
 Like double click detection in C++, note that the state machine is encoded
-across 3 different methods, each intermixing code with unrelated functionality.
+across 3 different methods, each intermixing code with unrelated functionality
+(e.g., changing frames, moving, and redrawing).
 
 The equivalent code in \CEU for the \emph{Bomber action} in
-Figure~\ref{lst.bomber}.b doesn't require state variables and reflects the
+Figure~\ref{lst.bomber}.b does not use state variables and reflects the
 sequential state machine implicitly, using \code{await} statements in direct
-style to separate the actions.
+style to separate the effects.
 %
@@
 The code tracks the animation aliveness (ln. 7--27) and, on termination,
-performs the last bomber action (ln. 30).
-As soon as the animation starts, the code performs the first action (ln. 9).
-The intermediate actions are performed when the corresponding conditions occur
+performs the last bomber effect, killing the pingu (ln. 30).
+As soon as the animation starts, the code performs the first effect (ln. 9).
+The intermediate effects are performed when the corresponding conditions occur
 (ln. 12,16,24).
@@
 %
-In contrast with the implementation in C++, all actions happen in a contiguous
-chunk of code (ln. 5--30) which handles no extra functionality.
+In contrast with the implementation in C++, all effects occur in a contiguous
+chunk of code (ln. 7--30), which handles no extra functionality.
 
@@
 
-The structured constructs of \CEU provide some advantages in comparison to 
+The structured constructs of \CEU introduce some advantages in comparison to 
 explicit state machines:
@@
 Object-oriented games also adopt the \emph{state pattern} to model state
-machines, with subclasses describing each possible state~\cite{games.patterns}.
-However, this approach is not fundamentally different from using \code{switch}
-or \code{if} branches for each possible state.
+machines with subclasses describing each possible state~\cite{games.patterns}.
+However, this approach is not fundamentally different from Pingus' use of
+\code{switch} or \code{if} branches for each possible state.
 
@@
 As Figure~\ref{tab.actions} shows, 5 of them implement at least one state
-machine and are considerable smaller in \CEU in terms of LoCs when eliminating
+machine and are considerable smaller in \CEU in terms of LoCs after eliminating
 the state variables.
@@
 
-    The completion of a long-lasting activity may carry a continuation, i.e.,
-    some action to execute next in the game.
+    The completion of a long-lasting activity may carry a continuation in the
+    game, i.e., some action to execute next.
 
@@
 \includegraphics[width=\columnwidth]{story}
-\caption{State machine for advancing pages in the \emph{Story screen}.
+\caption{Continuation mechanism and state machine for advancing pages in the \emph{Story screen}.
 \label{fig.story}
@@
 In C++, the class \code{StoryScreenComponent} in Figure~\ref{lst.story}.a
-implements the method \code{next\_text}, which is a callback for clicks in
-\code{>>>}.
-%
-The variable `pages` (ln. 4--5, 24--26) is a vector holding each page, but
+implements the method \code{next\_text}, which is a callback for clicks in the
+button \code{>>>}.
+%
+The variable \code{pages} (ln. 4--5, 24--26) is a vector holding each page, but
 which also encodes \emph{continuations} for the story progress:
@@
 The sequential navigation from page to page uses a loop in direct style
-(ln. 6--15) instead of explicit state variables for the continuation and state
-machine.
+(ln. 6--15) instead of explicit state variables for the continuation mechanism
+and state machine.
 While the text advances in an inner loop (hidden in ln. 9), we watch the
@@
 The loop may also eventually terminate with the time elapsing normally.
-This way, we do not need a variable (such as `displayed` in C++) to switch 
+This way, we do not need a variable (such as \code{displayed} in C++) to switch 
 between the states ``advancing text'' and ``advancing pages''.
@@
 \item It uses structured control flow (i.e., sequences and loops) instead of 
-      explicit data structures (e.g., stacks) or continuation variables.
-\item The activities are decoupled and do not hold references to one another.
+      explicit data structures (e.g., stacks) and continuation variables (e.g.
+      indexes).
+\item The activities in sequence are decoupled and do not hold references to
+      one another. % TODO: esta no outro exemplo
 \item A single parent class describes the flow between the activities in a 
-      self-contained block of code.
+      self-contained block of code. % TODO: esta no outro exemplo
 \end{itemize}
@@
 void Bomber::update () {
-    sprite.update ();
+    sprite.update();
 }
@@
     <...>
-    var&? Sprite sprite = spawn Sprite(<...>);
+    var Sprite sprite = spawn Sprite(<...>);
     <...>
@@
 
+TODO: code for Sprite update?
+
 In C++, the class \code{Bomber} presented in Figure~\ref{lst.hier}.a declares a
-\code{sprite} member to handle its animation frames
+\code{sprite} member (ln. 3) to handle its animation frames
 (Figure~\ref{fig.bomber.fsm}).
@@
 To understand how the \code{update} callback flows from the original
-environment stimulus from the game down to the sprite, we need to follow a long
+environment stimulus to the game down to the sprite, we need to follow a long
 chain of 7 method dispatches (Figure~\ref{fig.hier}):
@@
 \item With a single assignment to \code{last\_screen}, we can easily deactivate
-      the current screen and redirect all dispatches to a new screen.
+      the current screen and redirect all dispatches to a new screen (step 2).
 \item The \code{World} class manages and dispatches events to all game
       entities, such as all pingus and traps, with the common interface
-      \code{WorldObj}.
+      \code{WorldObj} (step 4).
 \item Since it is common to iterate only over the pingus (vs. all world
-      objects), the container \code{PinguHolder} manages all pingus.
+      objects), the container \code{PinguHolder} manages all pingus (step 5).
 \item Since a single pingu can change between actions during lifetime, the
-      \code{action} member decouples them with another level of indirection.
+      \code{action} member decouples them with another level of indirection
+      (step 6).
 \item Sprites are part of the game engine and are reusable everywhere (e.g., UI
       buttons, world objects, etc.), so it is also convenient to decouple them
-      from actions.
+      from actions (step 7).
 \end{itemize}
@@
 In \CEU, the \code{Bomber} abstraction presented in Figure~\ref{lst.hier}.b
-spawns a \code{Sprite} animation instance on its body.
-%
-The \code{Sprite} instance (ln. 3) can react directly to external \code{update}
+spawns a \code{Sprite} animation instance on its body (ln. 3).
+%
+The \code{Sprite} abstraction can react directly to external \code{update}
 and \code{draw} events, bypassing the program hierarchy entirely.
-While and \emph{only while} the bomber abstraction is alive, the sprite
-animation is also alive.
+External events in \CEU are broadcasted to the entire application.
+While (\emph{and only while}) the bomber abstraction is alive, the sprite
+animation remains alive.
 The radical decoupling between the program hierarchy and reactions to events
 eliminates dispatching chains entirely.
+Note that one can still declare a local event and restrict its visibility like
+a local variable.
 
@@
 Note that dispatching hierarchies cross game engine code, suggesting that most
-games use this control-flow pattern heavily.
+games also rely heavily on this control-flow pattern.
 In the case of the Pingus engine, we rewrote 9 files from C++ to \CEU, reducing
-them from 515 to 173 LoCs, mostly due to dispatching code removal.
+them from 515 to 173 LoCs (not shown in Figure~\ref{tab.tree}), mostly due to
+dispatching code removal.
 
@@
 
-Figure~\ref{fig.ui} shows the game UI widgets with action buttons, the pingus
-counter, and a small map which all coexist with the game screen during its
+Figure~\ref{fig.ui} shows the game UI widgets with the action buttons, score
+counters, and a small map which all coexist with the game screen during its
 whole lifespan.
-%
+
 In C++, the widgets are created in the constructor of the class
@@
 (non-dynamic) members.
-However, the class uses a container to automate \code{draw} and \code{update}
-dispatching to the widgets, as discussed in Section~\ref{sec.pats.dispatching}.
-In turn, the container method \code{add} only expects dynamically allocated
-children because they are automatically deallocated inside the container
-destructor.
-%
-The dynamic nature of containers in C++ demand extra caution from programmers:
+However, the class relies on a container to automate \code{draw} and
+\code{update} dispatching to the widgets, as discussed in
+Section~\ref{sec.pats.dispatching}.
+The container method \code{add} expects only dynamically allocated children
+because they are automatically deallocated inside the container destructor.
+%
+However, the dynamic nature of containers in C++ demand extra caution from
+programmers:
 %
@@
 
-In Céu, entities that coexist just have to be created in the same lexical
-block in Figure~\ref{lst.ui}.b (ln. 3--5).
+In \CEU, the UI entities that coexist just have to be created in the same
+lexical block in Figure~\ref{lst.ui}.b (ln. 3--5).
 %
@@
 lexical scope to delimit the lifespan of the explosion sprite to a single
-frame.
+frame (Figure~\ref{lst.bomber}, ln. 19--25).
 
@@
 In C++, the class \code{PinguHolder} in Figure~\ref{lst.pingus}.a is a
-container that holds all pingus alive.
+container that holds all alive pingus.
 %
@@
 Without the \code{erase} call above, a dead pingu would remain in the
-collection with updates on every frame (ln. 11).
+collection still with updates on every frame (ln. 11).
 Since the \code{draw} behavior for a dead pingu is innocuous, the death could
-go unnoticed but the program would keep consuming memory and CPU time.
+go unnoticed when testing it but the program would keep consuming memory and
+CPU time.
 This problem is known as the \emph{lapsed listener}~\cite{games.patterns} and
 also occurs in languages with garbage collection:
-A container typically holds a strong reference to a child (sometimes the only 
+a container typically holds a strong reference to a child (sometimes the only 
 reference to it), and the runtime cannot magically detect it as garbage.
@@
 \item Lexical scope makes an abstraction lifespan explicit in the source code.
+      All entities in a game have an associated lexical lifespan.
 \item The memory for static instances is known at compile time.
@@
 
-All entities in a game have an associated lifespan.
-
 The implementation in \CEU has over 200 static instantiations spread across all
@@
 patterns~\cite{gof}, which some previous work discuss in the context of video
-game development~\cite{games.patterns,games.gof.2015,games.gof.2007}.
-%
-The original game in C++ uses variations of the
+games~\cite{games.patterns,games.gof.2015,games.gof.2007}.
+%
+The original Pingus in C++ uses variations of the
     \emph{state} (Sections~\ref{sec.pats.fsms}~and~\ref{sec.pats.cps}),
     \emph{visitor} (Sections~\ref{sec.pats.dispatching}~and~\ref{sec.pats.lifespan}), and
-    \emph{observer} (Section~\ref{sec.pats.signaling} and to handle input in general)
+    \emph{observer} (to handle events in general)
 patterns as implementation details to achieve the desired higher-level
@@
 %
-As an example, the \emph{state pattern} for the bomber animation in
+As an example, the \emph{state pattern} for the bomber animation in C++ in
 Section~\ref{sec.pats.fsms} becomes a series of blocks separated by
-\code{await} statements.
+\code{await} statements in \CEU.
 
@@
 We also implemented all entity animations and behaviors
-(Section~\ref{sec.pats.fsms}), screen transitions
-(Section~\ref{sec.pats.cps}), and intermodule communication
-(Section~\ref{sec.pats.signaling}) using the available control mechanisms.
-%
-\CEU is a superset of C targeting reactive systems in general, not only games,
-and has also been adopted in other domains, such as
+(Section~\ref{sec.pats.fsms}), and screen transitions
+(Section~\ref{sec.pats.cps})
+%, and intermodule communication (Section~\ref{sec.pats.signaling})
+using the available control mechanisms of \CEU.
+%
+Furthermore, \CEU is a superset of C targeting reactive systems in general, not
+only games, and has also been successfully adopted in other domains, such as
     wireless sensor networks~\cite{ceu.sensys13,ceu.terra} and
@@
 Functional reactive programming (FRP)~\cite{frp.fran} contrasts with
-Structured synchronous reactive programming as a complementary programming
-style for reactive applications.
+structured synchronous reactive programming (SSRP) as a complementary
+programming style for reactive applications.
 %
@@
 %
-We rewrote an existing game in small steps while keeping it working.
+Instead, we rewrote an existing game and did it in small steps while keeping it
+working.
 Although we do not provide a performance evaluation (Pingus is not performance
@@
 
+TODO: rever summaries por advantage qualitativa vs LoCs
+
 [goto]: tirania.org/blog/archive/2013/Aug-15.html

### 2017-05-26 paper.tex
...

@@
 
-\subsubsection{Case Study: Advancing Pages in the \emph{Story screen}}
-
-%\begin{figure}[t]
-%\centering
-%\includegraphics[width=\columnwidth]{story-anim}
-%\caption{The \emph{Story screen}.
-%\label{fig.story}
-%}
-%\end{figure}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=\columnwidth]{story}
-\caption{Continuation mechanism and state machine for advancing pages in the \emph{Story screen}.
-\label{fig.story}
-}
-\end{figure}
+\subsubsection{ Transition to the \emph{Credits Screen} from the \emph{Story Screen} }
 
 \begin{figure*}[t]
-\begin{minipage}[t]{0.55\linewidth}
+\begin{minipage}[t]{0.50\linewidth}
 \begin{lstlisting}[numbers=left,xleftmargin=3em]
-StoryScreenComponent::StoryScreenComponent (<...>) :
-    <...>
+StoryDot::StoryDot(const FileReader& reader) :
+  credits(false), // do not display by default
 {
-    pages        = <...>; // vector with loaded pages
-    current_page = pages.back(); // first loaded page
-    displayed    = false; // if current is complete
-    <...>
-}
-
-<...>   // draw page over time
-
-void StoryScreenComponent::update (<...>) {
-    <...>
-    if (<all-words-appearing>) {
-        displayed = true;
-    }
-}
-
-void StoryScreenComponent::next_text() {
+  <...>
+  reader.read("credits", credits); // from file
+}
+
+void StoryDot::on_click() {
+  <...>
+  push_screen(<StoryScreen>(<...>, credits));
+  <...>
+}
+
+///
+
+StoryScreenComp::StoryScreenComp (<...>) :
+  credits(credits),
+  <...>
+{
+  <...>
+}
+
+<...>   // draw and update page
+
+void StoryScreenComp::next_text() {
   if (!displayed) {
-        displayed = true;
-        <...>     // remove current page
+    <...>
   } else {
-        pages.pop_back();
-        if (!pages.empty()) { // next page
-            current_page = pages.back();
-            displayed    = false;
+    <...>
+    if (!pages.empty()) {
       <...>
     } else {
-            <...> // terminates the story screen
+      if (credits) {
+        replace_screen(<Credits>(<...>));
+      } else {
+        pop_screen();
+      }
     }
@@
 %
-\begin{minipage}[t]{0.45\linewidth}
+\begin{minipage}[t]{0.50\linewidth}
 \begin{lstlisting}[numbers=left,xleftmargin=3em]
-code/await Story (void) -> bool do
-    <...>
-    event void next_text; // clicks in >>>
-
-    { pages = <...>; } // same as in C++
-    loop i in [0 <- {pages.size()}[ do
-        par/or do
-            watching next_text do
-                <...> // advance text
-            end
-            await next_text;
-        with
-            <...> // redraw _pages[i]
-        end
+loop do
+  var int ret = await Worldmap();
+  if ret=={STORY_MAP} or ret=={STORY_CREDITS} then
+    <...>
+    var bool is_click = await Story();
+    if is_click and ret=={STORY_CREDITS} then
+      <...>
+      await Credits();
+    end
+  else
+    <...>
   end
@@
 %\rule{8.4cm}{0.37pt}
-\caption{ Advancing pages in the \emph{Story Screen}.
+\caption{ Transition from \emph{Story} to \emph{Credits screen}.
 \label{lst.story}
@@
 
-The clickable \emph{blue dots} in the campaign world map transit to ambience
-story screens%
-\footnote{Story screen animation: \url{github.com/an000/p/blob/master/README.md#4} }.
-A story is composed of multiple pages and, inside each page, the words of the
-story appear incrementally over time.
-A first click in the button \code{>>>} fast forwards the words to show the full 
-page.
-A second click advances to the next page, until the story terminates.
-If the page completes before a click (due to the time elapsing), a first click 
-advances to the next page.
-%
-
-In C++, the class \code{StoryScreenComponent} in Figure~\ref{lst.story}.a
-implements the method \code{next\_text}, which is a callback for clicks in the
-button \code{>>>}.
-%
-The variable \code{pages} (ln. 4--5, 24--26) is a vector holding each page, but
-which also encodes \emph{continuations} for the story progress:
-each call to \code{next\_text} that advances the story (ln. 23--32) removes the 
-current page (ln. 24) and sets the next action to perform (i.e., ``display a
-new page'') in the variable \code{current\_page} (ln. 26).
-Figure~\ref{fig.story} illustrates the continuation mechanism to advance 
-pages and also a state machine for fast forwarding words (inside the dashed
-rectangle).
-The state variable \code{displayed} (ln. 6,15,20,21,27) switches between the
-behaviors ``advancing text'' and ``advancing pages'', which are both handled
-intermixed inside the method \code{next\_text}.
-
-The code in \CEU in Figure~\ref{lst.story}.b uses the internal event
-\code{next\_text}, which is emitted from clicks in \code{>>>}.
-%
-The sequential navigation from page to page uses a loop in direct style
-(ln. 6--15) instead of explicit state variables for the continuation mechanism
-and state machine.
-While the text advances in an inner loop (hidden in ln. 9), we watch the
-\code{next\_text} event that fast forwards it.
-The loop may also eventually terminate with the time elapsing normally.
-This way, we do not need a variable (such as \code{displayed} in C++) to switch 
-between the states ``advancing text'' and ``advancing pages''.
-The \code{par/or} makes the page advance logic to execute in parallel with the
-redrawing code (ln. 13).
-Whenever the page advances, the redrawing code is automatically aborted
-(due to the \code{or} modifier).
-The \code{await next\_text} in sequence (ln. 11) is the condition to advance to
-the next page.
-%
-Note that, unlike the implementation in C++, the ``advancing text'' behavior is
-not intermixed with the ``advancing pages'' behavior, instead, it is
-encapsulated inside the inner loop nested with a deeper indentation (ln. 9).
+The campaign world map has clickable blue dots for introductory and ending
+ambience stories in the two extremes of the map progress trail.
+For introductory stories, the game returns to the world map after displaying 
+the story pages.
+For ending stories, the game also displays a \emph{Credits screen} before
+returning to the world map.%
+\footnote{Story screen animation: \url{github.com/an000/p/blob/master/README.md#4} }
+
+In C++, the class \code{StoryDot} in Figure~\ref{lst.story}.a (ln. 1--12) reads
+the level file (ln. 5) to check whether its an ending story and should, after
+termination, display the \emph{Credits screen}.
+%
+The boolean variable \code{credits} is passed to the class
+\code{StoryScreen} (ln. 10) and represents the screen continuation, i.e., what
+to do after displaying the story.
+The class \code{StoryScreen} (not shown) then forwards the continuation even
+further to the auxiliary class \code{StoryScreenComp} (ln. 16--40).
+%
+When the method \code{next\_text} has no pages left to display (ln. 32--38),
+it decides where to go next, depending on the continuation flag
+\code{credits} (ln. 33--37).
+
+In \CEU, the \code{loop} of Figure~\ref{lst.story}.b controls the flow between
+the screens to display as a direct sequence of statements.
+%
+We first invoke the \code{Worldmap} (ln. 2), which exhibits the map and let
+the player interact with it until a dot is clicked.
+If the player selects a story dot (ln. 4--9), we invoke the \code{Story}
+and await its termination (ln. 5).
+Finally, we check the returned values (ln. 6) to perhaps display the
+\code{Credits} screen (ln. 8).
+The enclosing loop restores the \code{Worldmap} and repeats the process.
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=\columnwidth]{continuation}
+\caption{ Continuation (C++) vs Direct (\CEU) Styles.
+\label{fig.story}
+}
+\end{figure}
+
+Figure~\ref{fig.story} illustrates the \emph{continuation-passing style} of
+C++ and the \emph{direct style} of \CEU for screen transitions:
+
+\begin{enumerate}
+\item \code{Main Loop} $\longrightarrow$ \code{Worldmap}:
+    \begin{itemize}
+    \item C++ uses an explicit stack to push the \code{Worldmap} screen.
+    \item \CEU invokes the \code{WorldMap} screen expecting a return value
+          (ln. 2).
+    \end{itemize}
+\item \code{Worldmap} (\emph{blue dot click}) $\longrightarrow$ \code{Story}:
+    \begin{itemize}
+    \item C++ pushes the \code{Story} screen passing the continuation flag
+          (ln. 10).
+    \item \CEU stores the \code{Worldmap} return value and invokes the \code{Story} screen
+          (ln. 2,5).
+    \end{itemize}
+\item \code{Story} $\longrightarrow$ \code{Credits}:
+    \begin{itemize}
+    \item C++ replaces the current \code{Story} screen with the \code{Credits}
+          screen (ln. 34).
+    \item \CEU invokes the \code{Credits} screen after the \code{await Story}
+          returns (ln. 8).
+    \end{itemize}
+\item \code{Credits} $\longrightarrow$ \code{Worldmap}:
+    \begin{itemize}
+    \item C++ pops the \code{Credits} screen, going back to the \code{Worldmap}
+          screen.
+          \CEU uses an enclosing \code{loop} to restart the process (ln. 1--13).
+    \end{itemize}
+\end{enumerate}
+
+In contrast with C++, the screens in \CEU are decoupled and only the
+\code{Main Loop} touches them: the \code{Worldmap} has no references to
+\code{Story}, which has no references to \code{Credits}.
 
@@
       explicit data structures (e.g., stacks) and continuation variables (e.g.
-      indexes).
+      boolean flags).
 \item The activities in sequence are decoupled and do not hold references to

### 2017-05-27 paper.tex
...

@@
 
+\usepackage{balance}
+
+% No space between bibliography items:
+\let\oldthebibliography=\thebibliography
+  \let\endoldthebibliography=\endthebibliography
+  \renewenvironment{thebibliography}[1]{%
+    \begin{oldthebibliography}{#1}%
+      \setlength{\parskip}{0ex}%
+      \setlength{\itemsep}{0ex}%
+  }%
+  {%
+    \end{oldthebibliography}%
+  }
+\setlength{\parindent}{5mm}
+
+\newcommand{\compresslist}{
+	\setlength{\itemsep}{1pt}
+	\setlength{\parskip}{0pt}
+	\setlength{\parsep}{0pt}
+}
+
+\newenvironment{itemize*}%
+  {\begin{itemize}%
+    \setlength{\itemsep}{0pt}%
+    \setlength{\parskip}{0pt}}%
+  {\end{itemize}}
+\usepackage{enumitem}
+\setlist{nolistsep}
+
 \begin{document}
@@
 \begin{itemize}
+%\setlength\itemsep{0em}
 \item \emph{Reactive}: code only executes in reactions to events.
@@
 \begin{itemize}
+%\setlength\itemsep{0em}
 \item Alternative solutions in idiomatic \CEU with gains in productivity for
@@
 \begin{enumerate}
+%\setlength\itemsep{0em}
 \item \emph{Finite State Machines}:
@@
 \centering
-\includegraphics[width=\columnwidth]{bomber-state}
+\includegraphics[width=\columnwidth]{states}
 \caption{ State machine for the \emph{Bomber animation} sequence.
@@
   spawn Mover(); // movement in the background
-  var Sprite s = spawn Sprite(<...>);
+  var Sprite sprite = spawn Sprite(<...>);
                  // animation in the background
@@
     // 2. plays a "Bomb!" sound.
-    await game.update until s.sprite.frame == 10;
+    await game.update until sprite.frame == 10;
     {play_sound("plop")};
@@
     // 3. particles, terrain, explosion sprite
-    await game.update until s.sprite.frame == 13;
+    await game.update until sprite.frame == 13;
     spawn PinguParticles(<...>) in particles;
@@
 \begin{enumerate}
+%\setlength\itemsep{0em}
 \item 0th frame:  plays a "Oh no!" sound.
@@
 \begin{itemize}
+%\setlength\itemsep{0em}
 \item They encode all states with direct sequential code, eliminating shared
@@
   var int ret = await Worldmap();
-  if ret=={STORY_MAP} or ret=={STORY_CREDITS} then
+  if ret==STORY_MAP or ret==STORY_CREDITS then
     <...>
     var bool is_click = await Story();
-    if is_click and ret=={STORY_CREDITS} then
+    if is_click and ret==STORY_CREDITS then
       <...>
@@
 returning to the world map.%
-\footnote{Story screen animation: \url{github.com/an000/p/blob/master/README.md#4} }
+\footnote{Credits screen animation: \url{github.com/an000/p/blob/master/README.md#4} }
 
@@
 
-\begin{figure}[t]
+\begin{figure*}[t]
 \centering
-\includegraphics[width=\columnwidth]{continuation}
+\includegraphics[width=\textwidth]{continuation}
 \caption{ Continuation (C++) vs Direct (\CEU) Styles.
@@
 }
-\end{figure}
+\end{figure*}
 
@@
 \begin{enumerate}
+%\setlength\itemsep{0em}
 \item \code{Main Loop} $\longrightarrow$ \code{Worldmap}:
     \begin{itemize}
+    %\setlength\itemsep{0em}
     \item C++ uses an explicit stack to push the \code{Worldmap} screen.
@@
     \begin{itemize}
+    %\setlength\itemsep{0em}
     \item C++ pushes the \code{Story} screen passing the continuation flag
@@
     \begin{itemize}
+    %\setlength\itemsep{0em}
     \item C++ replaces the current \code{Story} screen with the \code{Credits}
@@
     \begin{itemize}
+    %\setlength\itemsep{0em}
     \item C++ pops the \code{Credits} screen, going back to the \code{Worldmap}
@@
 \begin{itemize}
+%\setlength\itemsep{0em}
 \item It uses structured control flow (i.e., sequences and loops) instead of 
@@
 
-\begin{figure}[t]
+\begin{figure*}[t]
 \centering
-\includegraphics[width=\columnwidth]{hierarchy}
+\includegraphics[width=\textwidth]{hierarchy}
 \caption{Dispatching chain for \code{update}.
@@
 \begin{enumerate}
+%\setlength\itemsep{0em}
 \item \code{ScreenManager::display} in the main game loop calls \code{update}.
@@
 \begin{itemize}
+%\setlength\itemsep{0em}
 \item With a single assignment to \code{last\_screen}, we can easily deactivate
@@
 \begin{itemize}
+%\setlength\itemsep{0em}
 \item The full dispatching chain may go through dozens of files.
@@
 \begin{itemize}
+%\setlength\itemsep{0em}
 \item When containers are part of a dispatching chain, it gets even harder to
@@
             <...>
-            escape {PS_DEAD};
+            escape PS_DEAD;
         end
@@
 \begin{itemize}
+%\setlength\itemsep{0em}
 \item Lexical scope makes an abstraction lifespan explicit in the source code.

### 2017-05-27 paper.tex
...

@@
 
-\CEU~\cite{ceu.sensys13,ceu.mod15} is a C/C++ source compatible programming
-language with the following characteristics:
+In this work, we advocate structured synchronous reactive programming as an
+expressive and productive alternative for game logic development.
+We present a case study of rewriting Pingus from C++ to the programming
+language \CEU~\cite{ceu.sensys13,ceu.mod15}, which has the following
+characteristics:
 %
@@
 
-In this work, we advocate structured synchronous reactive programming as an
-expressive and productive alternative for game logic development.
-We present a case study of rewriting Pingus from C++ to \CEU with the
-contributions as follows:
-%
-\begin{itemize}
-%\setlength\itemsep{0em}
-\item Alternative solutions in idiomatic \CEU with gains in productivity for
-      six selected behaviors in the game logic.
-\item An in-depth qualitative analysis of the proposed solutions in comparison
-      to the original implementations in C++.
-\item The identification of four recurrent control-flow patterns that likely
-      apply to other games:
+Our case study demonstrates gains in productivity for six selected behaviors in
+the game logic of Pingus through idiomatic code in \CEU.
+We present an in-depth qualitative analysis of the proposed solutions in
+comparison to the original implementations in C++.
+We also identify four recurrent control-flow patterns that likely apply to
+other games:
         \emph{Finite State Machines},
@@
         \emph{Lifespan Hierarchies}.
-        %\emph{Signaling Mechanisms}.
-\end{itemize}
 %
 A control-flow pattern is a recurring technique to describe dependencies and
-explicit orders between statements (or groups of statements) in a program.
+explicit orders between statements (or groups of statements) execution in a
+program.
 %For instance, consider how a key press stimulus propagates through the game
@@
 %game.
-%
+
 The rewriting process consisted of identifying sets of callbacks implementing
@@
 of the screen literally explodes all pingus.%
-\footnote{Double click animation: \url{github.com/an000/p/blob/master/README.md#1} }
+\footnote{Double click animation: \url{github.com/an000/p/#1} }
 
@@
 and also destroying the terrain under its radius.%
-\footnote{Bomber action animation: \url{github.com/an000/p/blob/master/README.md#2} }
+\footnote{Bomber action animation: \url{github.com/an000/p/#2} }
 %
@@
 follows%
-\footnote{State machine animation: \url{github.com/an000/p/blob/master/README.md#3} }:
+\footnote{State machine animation: \url{github.com/an000/p/#3} }:
 %
@@
 returning to the world map.%
-\footnote{Credits screen animation: \url{github.com/an000/p/blob/master/README.md#4} }
+\footnote{Credits screen animation: \url{github.com/an000/p/#4} }
 
@@
 conditions, such as falling from a high altitude%
-\footnote{Death of pingus animation: \url{github.com/an000/p/blob/master/README.md#5} }.
+\footnote{Death of pingus animation: \url{github.com/an000/p/#5} }.
 %

### 2017-05-27 paper.tex
...

@@
 variables and dependencies between classes.
+%
+We employed \emph{live code translation}, i.e., starting from the original
+codebase in C++, we reimplemented piece-by-piece without breaking the game
+compilation and execution.
+%
+This was only possible given the seamless integration between \CEU and
+C/C++~\cite{ceu.sensys13}: the type systems are equivalent and the integration
+happens at the source code level.
+This enables trivial sharing of control and data, i.e., accessing C/C++ data
+and calling C/C++ from \CEU and vice-versa.
 
@@
 
-In Pingus, the game logic also accounts for almost half the size of the
-codebase: 18.173 from 39.362 LoCs (46\%) spread across 272 files.
+In Pingus, the game logic accounts for almost half the size of the codebase:
+18.173 from 39.362 LoCs (46\%) spread across 272 files.
 %
@@
 %
-Therefore, we rewrote 9.186 LoCs spread across 126 files%
+Therefore, we only rewrote 9.186 LoCs spread across 126 files%
 \footnote{Complete codebase: \url{github.com/an000/p/tree/master/cpp}}.
@@
 %
-Figure~\ref{tab.tree} summarizes the effective codebase in the two
+Figure~\ref{tab.tree} summarizes the effective game logic codebase in the two
 implementations.
 
-Although the sections that follow compare the codebases qualitatively, the
-lines with lower ratio numbers in Figure~\ref{tab.tree} correlate to the parts
-of the game logic that we consider more susceptible to structured reactive
-programming.
+Although our analysis is mostly qualitative, the lines in Figure~\ref{tab.tree}
+with lower ratio numbers correlate to the parts of the game logic that we
+consider more susceptible to structured reactive programming.
 For instance, the \emph{Pingu} behavior (\emph{ratio 0.80}) contains complex
@@
 
-As a general rewriting rule, we could identify control-flow behaviors in C++ by
-looking for class members with identifiers resembling verbs, statuses, and
-counters (e.g.,
+As a general rewriting rule, we could identify control-flow behaviors in the
+C++ codebase by looking for class members with identifiers resembling verbs,
+statuses, and counters (e.g.,
 \code{pressed},
@@
 \code{delay\_count}).
-Good chances are that variables with these ``suspicious names'' encode some
-form of control-flow progression that cross multiple callback invocations.
-
-We employed \emph{live code translation}, i.e., starting from the original
-codebase in C++, we reimplemented piece-by-piece without breaking the game
-compilation and execution.
-%
-This was only possible given the seamless integration between \CEU and
-C/C++~\cite{ceu.sensys13}: the type systems are equivalent and the integration
-happens at the source code level.
-This enables trivial sharing of control and data, i.e., accessing C/C++ data
-and calling C/C++ from \CEU and vice-versa.
+Good chances are that such variables encode some form of control-flow
+progression that cross multiple callback invocations.
 

### 2017-05-27 paper.tex
...

@@
 We describe six representative game behaviors in detail distributed in the four
-patterns, with their implementations in C++ and \CEU.
-%
-Due to space constraints, we omit five other cases and also a fifth pattern
-\emph{Signaling Mechanisms} entirely.
+patterns and analyse their implementations in C++ and \CEU.%
+\footnote{Due to space constraints, we omit a fifth pattern
+\emph{Signaling Mechanisms} and three other game behaviors.}
 
@@
 %
-Note in the code how the accesses to the state variables are spread
-across the entire class.
+Note in Figure~\ref{lst.armageddon} how the accesses to the state variables are
+spread across the entire class.
 For instance, the distance between the initialization of \code{pressed} (ln.
@@
 
-\CEU provides structured constructs to deal with events, aiming to eradicate
+\CEU supports structured constructs to deal with events, aiming to eradicate
 explicit manipulation of state variables for control-flow purposes.
@@
 %
-Double click detection in \CEU doesn't require state variables and is entirely
+Double click detection in \CEU does not rely on state variables and is entirely
 self-contained in the \code{loop} body.
 Also, these 7 lines of code \emph{only} detect the double click, leaving the
-actual effect to happen outside the loop (ln. 12) as well as all unrelated
+actual effect to happen outside the loop (ln. 12), as well as all unrelated
 code such as redrawing the button.
@@
 
-In Pingus, the player may assign actions to specific pingus, as illustrated in
+The player may assign actions to specific pingus, as illustrated in
 Figure~\ref{fig.bomber.action}.
@@
 movement on every frame regardless of its current state (ln. 15--16).
-When the animation reaches the 10th frame, it plays the ``Bomb!'' sound and 
-switches to \emph{state-2} (ln. 18--22).
+When the animation reaches the 10th frame, it switches to \emph{state-2} and
+plays the ``Bomb!'' sound (ln. 18--22).
 The state variable \code{sound\_played} is required because the sprite frame
@@
 The same reasoning and technique applies to \emph{state-3} (ln. 24--32 and
-43--44).
+41--46).
 The explosion sprite appears in a single frame in \emph{state-4} (ln. 45).
@@
 %
-Note that a single numeric state variable suffices to track the states, but the
-original authors probably chose to encode each state in an independent boolean 
+Note that a single numeric state variable suffices to track the states
+(Figure~\ref{fig.bomber.fsm}), but the original developers probably chose to
+encode each state in an independent boolean 
 variable to rearrange and experiment with them during development.
@@
 unavoidable and are actually the essence of object-oriented programming
-(i.e., \emph{methods with mutable state}).
-%
-Like double click detection in C++, note that the state machine is encoded
+(i.e., methods with mutable state).
+%
+Like the double click detection in C++, note that the state machine is encoded
 across 3 different methods, each intermixing code with unrelated functionality
@@
 The equivalent code in \CEU for the \emph{Bomber action} in
-Figure~\ref{lst.bomber}.b does not use state variables and reflects the
-sequential state machine implicitly, using \code{await} statements in direct
-style to separate the effects.
+Figure~\ref{lst.bomber}.b does not rely on state variables and reflects the
+sequential state machine implicitly, using \code{await} statements to separate
+the effects in direct style.
 %
@@
 The event \code{game.update} (ln. 12,16,24) is analogous to the \code{update}
-callback of C++ and occurs on every frame.
+callback of C++ and occurs on every game frame.
 %
@@
 
-\subsubsection{Summary \& Uses in Pingus}
+\subsubsection{Summary \& Pattern Uses in Pingus}
 
@@
 \item They encode all states with direct sequential code, eliminating shared
-      state variables.
+      state variables for control-flow purposes.
 \item They handle all states (and only them) in the same contiguous block,
@@
 
-Among the 65 implementation files in \CEU, we found 29 cases in 25 files using
+Among all 65 implementation files in \CEU, we found 29 cases in 25 files using
 structured mechanisms to substitute states machines.
-They manifest as \code{await} statements in sequence or in aborting constructs
-such as \code{par/or} and \code{watching}.
+They typically manifest as \code{await} statements in sequence.
+%or in aborting constructs such as \code{par/or} and \code{watching}.
 

### 2017-05-28 paper.tex
...

@@
       line of execution.
-      There's no implicit preemption or real parallelism, resulting in
+      There is no implicit preemption or real parallelism, resulting in
       deterministic execution.
@@
 \item \emph{Continuation Passing}:
-    The completion of a long-lasting activity may carry a continuation in the
-    game, i.e., some action to execute next.
+    The completion of a long-lasting activity in the game may carry a
+    continuation, i.e., some action to execute next.
 \item \emph{Dispatching Hierarchies}:
@@
   <...>
-  spawn Mover(); // movement in the background
+  spawn Mover(); // movement in background
   var Sprite sprite = spawn Sprite(<...>);
-                 // animation in the background
-  watching s do
+               // frame animation in background
+  watching sprite do
     // 1. plays a "Oh no!" sound.
@@
 
-    The completion of a long-lasting activity may carry a continuation in the
-    game, i.e., some action to execute next.
+    The completion of a long-lasting activity in the game may carry a
+    continuation, i.e., some action to execute next.
 
@@
   var int ret = await Worldmap();
-  if ret==STORY_MAP or ret==STORY_CREDITS then
+  if ret==STORY_CREDITS or ret==STORY_BACK then
     <...>
@@
 
-The campaign world map has clickable blue dots for introductory and ending
-ambience stories in the two extremes of the map progress trail.
+The campaign world map has clickable blue dots in the two extremes of the
+progress trail for introductory and ending ambience stories, respectively.
 For introductory stories, the game returns to the world map after displaying 
@@
 
-In C++, the class \code{StoryDot} in Figure~\ref{lst.story}.a (ln. 1--12) reads
-the level file (ln. 5) to check whether its an ending story and should, after
-termination, display the \emph{Credits screen}.
+In C++, the class \code{StoryDot} in Figure~\ref{lst.story}.a (ln. 1--12) first
+reads the level file (ln. 5) to check whether its an ending story and should,
+after termination, display the \emph{Credits screen}.
 %
@@
 it decides where to go next, depending on the continuation flag
-\code{credits} (ln. 33--37).
+\code{credits} (ln. 33).
 
@@
 We first invoke the \code{Worldmap} (ln. 2), which exhibits the map and let
-the player interact with it until a dot is clicked.
+the player interact with it (e.g., walking around) until a dot is clicked.
 If the player selects a story dot (ln. 4--9), we invoke the \code{Story}
@@
     \item C++ uses an explicit stack to push the \code{Worldmap} screen.
-    \item \CEU invokes the \code{WorldMap} screen expecting a return value
+    \item \CEU invokes the \code{Worldmap} screen expecting a return value
           (ln. 2).
@@
 
-In contrast with C++, the screens in \CEU are decoupled and only the
+In contrast with C++, the screens in \CEU are completely decoupled and only the
 \code{Main Loop} touches them: the \code{Worldmap} has no references to
@@
 It also uses the same technique for the loop that switches the pingu actions
-during gameplay.
+during gameplay (e.g., \emph{walking} to \emph{jumping} and back to
+\emph{walking}).
 
@@
 
-TODO: code for Sprite update?
-
-In C++, the class \code{Bomber} presented in Figure~\ref{lst.hier}.a declares a
-\code{sprite} member (ln. 3) to handle its animation frames
-(Figure~\ref{fig.bomber.fsm}).
+In C++, the class \code{Bomber} in Figure~\ref{lst.hier}.a declares a
+\code{sprite} member (ln. 3) to handle its animation frames.
 %
@@
 %\setlength\itemsep{0em}
-\item \code{ScreenManager::display} in the main game loop calls \code{update}.
+\item \code{ScreenManager::display} in the main game loop calls
+      \code{ScreenManager::update}.
 \item \code{ScreenManager::update} calls \code{last\_screen->update} for the
-      active game screen (a \code{GameSession} instance, considering the
+      active game screen (i.e., a \code{GameSession} instance, considering the
       \code{Bomber}).
@@
 \item The \code{World} class manages and dispatches events to all game
-      entities, such as all pingus and traps, with the common interface
-      \code{WorldObj} (step 4).
+      entities with a common interface \code{WorldObj}, such as all pingus and
+      traps (step 4).
 \item Since it is common to iterate only over the pingus (vs. all world
       objects), the container \code{PinguHolder} manages all pingus (step 5).
-\item Since a single pingu can change between actions during lifetime, the
+\item Since a single pingu can change its actions during lifetime, the
       \code{action} member decouples them with another level of indirection
@@
 %
-The \code{draw} callback flows through the same dispatching hierarchy until
-reaching the \code{Sprite} class.
+The \code{draw} callback also flows through a similar dispatching hierarchy
+until reaching the \code{Sprite} class.
 
@@
 In the case of the Pingus engine, we rewrote 9 files from C++ to \CEU, reducing
-them from 515 to 173 LoCs (not shown in Figure~\ref{tab.tree}), mostly due to
+them from 515 to 173 LoCs (not listed in Figure~\ref{tab.tree}), mostly due to
 dispatching code removal.
@@
 
-Figure~\ref{fig.ui} shows the game UI widgets with the action buttons, score
+Figure~\ref{fig.ui} shows the game UI widgets with action buttons, score
 counters, and a small map which all coexist with the game screen during its
@@
 
-In \CEU, the UI entities that coexist just have to be created in the same
-lexical block in Figure~\ref{lst.ui}.b (ln. 3--5).
+In \CEU, the UI entities that coexist are just created in the same lexical
+block of the abstraction \code{Game} in Figure~\ref{lst.ui}.b (ln. 3--5).
 %
@@
 A pingu is a dynamic entity created periodically and destroyed under certain
-conditions, such as falling from a high altitude%
-\footnote{Death of pingus animation: \url{github.com/an000/p/#5} }.
+conditions, such as falling from a high altitude.%
+\footnote{Death of pingus animation: \url{github.com/an000/p/#5} }
 %
@@
         if ((*pingu)->get_status() == PS_DEAD) {
-            pingu = pingus.erase(pingu);
+            pingu = pingus.remove(pingu);
         }
@@
 %
-Entities with dynamic lifespan in C++ require explicit \code{add} and
-\code{remove} calls associated to a container (ln. 4,13).
-Without the \code{erase} call above, a dead pingu would remain in the
-collection still with updates on every frame (ln. 11).
+Without the \code{remove} call, a dead pingu would remain in the collection
+still with updates on every frame (ln. 11).
 Since the \code{draw} behavior for a dead pingu is innocuous, the death could
@@
 CPU time.
+%
 This problem is known as the \emph{lapsed listener}~\cite{games.patterns} and
@@
 reference to it), and the runtime cannot magically detect it as garbage.
+%
+Entities with dynamic lifespan in C++ always require explicit \code{add} and
+\code{remove} calls associated to a container (ln. 4,13).
 
@@
 \CEU supports \code{pool} declarations to hold dynamic abstraction instances.
-Additionally, the \code{spawn} statement supports a pool identifier to
-associate the new instance with a pool.
+In addition, the \code{spawn} statement supports a pool identifier to associate
+the new instance with a pool.
 %
@@
 game state and screen progression~\cite{games.fsms.2006.1,games.fsms.2006.2}, and
-behavior and AI modeling~\cite{games.bts,games.bts.unreal}
-%
-In Pingus, we employed \CEU at the core of the game for event dispatching
+behavior and AI modeling~\cite{games.bts,games.bts.unreal}.
+%
+In Pingus, the adoption of \CEU is not restricted to a specific subsystem.
+We employed \CEU at the very core of the game for event dispatching
 (Section~\ref{sec.pats.dispatching}) and memory management of entities
@@
 %
-On the other hand, describing a sequence of steps in FRP requires to encode 
-explicit state machines so that functions can switch behavior depending on the 
-current state.
+On the other hand, describing a sequence of steps or control-flow dependencies
+in FRP requires to encode explicit state machines so that functions can switch
+behavior depending on the current state.
 %
@@
 context of embedded systems~\cite{ceu.sensys13}.
-Nonetheless, given the tight integration between with C/C++, critical parts of
-games can be preserved in C++ if needed.
+Nonetheless, given the tight integration between \CEU and C/C++, critical parts
+of games can be preserved in C++ if needed.
 

### 2017-05-28 paper.tex
(+) abstract

@@
 
-\abstract{TODO.
-
+\abstract{
+We present a case study of rewriting the video game Pingus from C++ to \CEU, a
+structured synchronous reactive programming language.
+%
+\CEU supports structured primitives that eliminates callbacks, letting
+programmers write code in direct and sequential style, which help describing
+complex control-flow relationships in the game logic more concisely.
+%
+We demonstrate gains in productivity for six behaviors in the game logic of
+Pingus with an in-depth qualitative analysis of the proposed solutions in \CEU
+in comparison to the original implementations in C++.
+%
+We also categorize the behaviors in four recurrent control-flow patterns that
+likely apply to other games.
 \smallskip

### 2017-05-30 paper.tex
(>) figures placement
...

--- 2017-05-28 paper.tex
@@
 
-\abstract{TODO.
-
+\abstract{
+We present a case study of rewriting the video game Pingus from C++ to the
+structured synchronous reactive language \CEU.
+%
+\CEU supports reactive control-flow primitives that eliminate callbacks and let
+programmers write code in direct and sequential style.
+Structured reactivity helps describing complex control-flow relationships in
+the game logic more concisely.
+%
+We demonstrate gains in productivity for six behaviors in the game logic of
+Pingus through a qualitative analysis of the proposed solutions in \CEU in
+comparison to the original implementations in C++.
+%
+We also categorize the behaviors in four recurrent control-flow patterns that
+likely apply to other games.
 \smallskip
 
-\noindent \textbf{Keywords:} TODO, TODO, TODO.
+%\noindent \textbf{Keywords:} \CEU, Game Logic, Event-Driven Programming.
+\noindent \textbf{Keywords:} C++, \CEU, Control Flow, Event-Driven Programming, Game Logic, Synchronous Reactive Programming.
 }
@@
 In this sense, callbacks actually disrupt structured programming, becoming
-``our generation's \code{goto}''.%
+``our generation's \emph{goto}''.%
 \footnote{``Callbacks as our Generations' Go To Statement'':
@@
 
-In this work, we advocate structured synchronous reactive programming as an
-expressive and productive alternative for game logic development.
+In this work, we advocate structured synchronous reactive programming as a more
+productive alternative for game logic development.
 We present a case study of rewriting Pingus from C++ to the programming
@@
 \item \emph{Structured}: programs use structured control-flow mechanisms, such
-      as \code{await} (to suspend a line of execution), and \code{par} (to
-      combine multiple lines of execution).
+      as \code{spawn} and \code{await} (to create and suspend an activity).
 \item \emph{Synchronous}: reactions run atomically and to completion on each
-      line of execution.
+      activity.
       There is no implicit preemption or real parallelism, resulting in
@@
 %
-\CEU provides primitives that help describing complex control-flow
-relationships in the game logic more concisely.
+%\CEU provides primitives that help describing complex control-flow
+%relationships in the game logic more concisely.
 %
@@
 %game.
-
-The rewriting process consisted of identifying sets of callbacks implementing
-control flow in the game and translating them to \CEU using appropriate
-structured constructs.
-%
-As an example, a double mouse click is characterized by a first click, followed
-by a maximum amount of time, followed by a second click.
-This behavior depends on different events (clicks and timers) which have to
-occur in a particular order.
-In C++, the implementation involves callbacks crossing reactions to successive
-events which manipulate state variables explicitly.
-%
-We focus on a qualitative analysis for the programming techniques that we
-applied during the rewriting process.
-Not all techniques result in reduction in LoCs (especially considering the
-verbose syntax of \CEU), but have other effects such as eliminating shared
-variables and dependencies between classes.
-%
-We employed \emph{live code translation}, i.e., starting from the original
-codebase in C++, we reimplemented piece-by-piece without breaking the game
-compilation and execution.
-%
-This was only possible given the seamless integration between \CEU and
-C/C++~\cite{ceu.sensys13}: the type systems are equivalent and the integration
-happens at the source code level.
-This enables trivial sharing of control and data, i.e., accessing C/C++ data
-and calling C/C++ from \CEU and vice-versa.
-
-The rest of the paper is organized as follows:
-Section~\ref{sec.codebase} gives an overview of the Pingus codebases in C++ and
-\CEU and describes our approach to identify and rewrite the control flow in the
-game.
-Section~\ref{sec.pats} discusses six case studies in detail which are
-categorized in four control-flow patterns.
-Section~\ref{sec.related} discusses related work.
-Section~\ref{sec.conclusion} concludes the paper.
-
-\section{The Pingus Codebase}
-\label{sec.codebase}
 
@@
 
+The rewriting process consisted of identifying sets of callbacks implementing
+control flow in the game and translating them to \CEU using appropriate
+structured constructs.
+%
+As an example, a double mouse click is characterized by a first click, followed
+by a maximum amount of time, followed by a second click.
+This behavior depends on different events (clicks and timers) which have to
+occur in a particular order.
+In C++, the implementation involves callbacks crossing reactions to successive
+events which manipulate state variables explicitly.
+%
+We focus on a qualitative analysis for the programming techniques that we
+applied during the rewriting process.
+Not all techniques result in reduction in LoCs (especially considering the
+verbose syntax of \CEU), but have other effects such as eliminating shared
+variables and dependencies between classes.
+%
+We employed \emph{live code translation}, i.e., starting from the original
+codebase in C++, we reimplemented piece-by-piece without breaking the game
+compilation and execution.
+%
+This was only possible given the seamless integration between \CEU and
+C/C++~\cite{ceu.sensys13}: the type systems are equivalent and the integration
+happens at the source code level.
+This enables trivial sharing of control and data, i.e., accessing C/C++ data
+and calling C/C++ from \CEU and vice-versa.
+
+The rest of the paper is organized as follows:
+Section~\ref{sec.codebase} gives an overview of the Pingus codebases in C++ and
+\CEU and describes our approach to identify and rewrite the control flow in the
+game.
+Section~\ref{sec.pats} discusses six case studies in detail which are
+categorized in four control-flow patterns.
+Section~\ref{sec.related} discusses related work.
+Section~\ref{sec.conclusion} concludes the paper.
+
+\section{The Pingus Codebase}
+\label{sec.codebase}
+
 In Pingus, the game logic accounts for almost half the size of the codebase:
@@
 
-%\begin{figure}[t]
-%\centering
-%\includegraphics[width=\columnwidth]{bomber-anim}
-%\caption{ The \emph{Bomber action}.
-%\label{fig.bomber.anim}
-%}
-%\end{figure}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=120px]{bomber-01}
-\caption{ Assigning the \emph{Bomber action} to a pingu.
-\label{fig.bomber.action}
-}
-\end{figure}
-
-\begin{figure}[t]
-\centering
-\includegraphics[width=\columnwidth]{states}
-\caption{ State machine for the \emph{Bomber animation} sequence.
-\label{fig.bomber.fsm}
-}
-\end{figure}
-
-\begin{figure*}[th!]
-\begin{minipage}[t]{0.50\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
-Bomber::Bomber (Pingu* p) :
-  <...>
-  sprite(<...>),          // bomber sprite
-  sound_played(false),    // tracks state 2
-  particle_thrown(false), // tracks state 3
-  colmap_exploded(false), // tracks state 3
-  gfx_exploded(false)     // tracks state 4
-{
-  <...>
-  // 1. plays a "Oh no!" sound.
-  play_sound("ohno");
-}
-
-void Bomber::update () {
-  sprite.update();
-  <...>   // pingu movement
-
-  // 2. plays a "Bomb!" sound.
-  if (sprite.frame()==10 && !sound_played) {
-    sound_played = true;
-    play_sound("plop");
-  }
-
-  // 3. particles, terrain, explosion sprite
-  if (sprite.frame()==13 && !particle_thrown) {
-    particle_thrown = true;
-    get_world()->get_particles()->add(...);
-  }
-  if (sprite.frame()==13 && !colmap_exploded) {
-    colmap_exploded = true;
-    get_world()->remove(bomber_radius, <...>);
-  }
-
-  // 5. kills the Pingu
-  if (sprite.is_finished ()) {
-    pingu->set_status(PS_DEAD);
-  }
-}
-
-void Bomber::draw (SceneContext& gc) {
-  // 3. particles, terrain, explosion sprite
-  // 4. tick: hides the explosion sprite
-  if (sprite.frame()==13 && !gfx_exploded) {
-    gfx_exploded = true;
-    gc.color().draw(explo_surf, <...>);
-  }
-  gc.color().draw(sprite, pingu->get_pos());
-}
-\end{lstlisting}
-\centering\small{\ax Implementation in C++}
-\end{minipage}
-%
-\begin{minipage}[t]{0.50\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
-code/await Bomber (void) -> ActionName
-do
-  <...>
-  spawn Mover(); // movement in background
-  var Sprite sprite = spawn Sprite(<...>);
-               // frame animation in background
-  watching sprite do
-    // 1. plays a "Oh no!" sound.
-    {play_sound("ohno")};
-
-    // 2. plays a "Bomb!" sound.
-    await game.update until sprite.frame == 10;
-    {play_sound("plop")};
-
-    // 3. particles, terrain, explosion sprite
-    await game.update until sprite.frame == 13;
-    spawn PinguParticles(<...>) in particles;
-    call Game_Remove({&bomber_radius}, <...>);
-    do
-      <...>
-      spawn Sprite(<...>); // explosion
-
-      // 4. tick: hides the explosion sprite
-      await game.update;
-    end
-    await FOREVER;
-  end
-
-  // 5. kills the pingu
-  escape DEAD;
-end
-
-.
-\end{lstlisting}
-\centering\small{\bx Implementation in \CEU}
-\end{minipage}
-%\rule{8.4cm}{0.37pt}
-\caption{ The \emph{Bomber action} sequence.
-\label{lst.bomber}
-}
-\end{figure*}
-
 The player may assign actions to specific pingus, as illustrated in
@@
 \end{enumerate}
+
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=\columnwidth]{bomber-anim}
+%\caption{ The \emph{Bomber action}.
+%\label{fig.bomber.anim}
+%}
+%\end{figure}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=120px]{bomber-01}
+\caption{ Assigning the \emph{Bomber action} to a pingu.
+\label{fig.bomber.action}
+}
+\end{figure}
+
+\begin{figure}[t]
+\centering
+\includegraphics[width=\columnwidth]{states}
+\caption{ State machine for the \emph{Bomber animation} sequence.
+\label{fig.bomber.fsm}
+}
+\end{figure}
+
+\begin{figure}[t]
+\begin{verbatim}
+Action       Ceu    C++   Explicit State
+---------   ----   ----   -----------------
+Bomber        23     50   4 state variables
+Bridger       75    100   2 state variables
+Drown          6     15   1 state variable
+Exiter         7     22   2 state variables
+Splashed       6     19   2 state variables
+\end{verbatim}
+\caption{Pingus actions in \CEU and C++.
+\label{tab.actions}
+}
+\end{figure}
+
+\begin{figure*}[!t]
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+Bomber::Bomber (Pingu* p) :
+  <...>
+  sprite(<...>),          // bomber sprite
+  sound_played(false),    // tracks state 2
+  particle_thrown(false), // tracks state 3
+  colmap_exploded(false), // tracks state 3
+  gfx_exploded(false)     // tracks state 4
+{
+  <...>
+  // 1. plays a "Oh no!" sound.
+  play_sound("ohno");
+}
+
+void Bomber::update () {
+  sprite.update();
+  <...>   // pingu movement
+
+  // 2. plays a "Bomb!" sound.
+  if (sprite.frame()==10 && !sound_played) {
+    sound_played = true;
+    play_sound("plop");
+  }
+
+  // 3. particles, terrain, explosion sprite
+  if (sprite.frame()==13 && !particle_thrown) {
+    particle_thrown = true;
+    get_world()->get_particles()->add(...);
+  }
+  if (sprite.frame()==13 && !colmap_exploded) {
+    colmap_exploded = true;
+    get_world()->remove(bomber_radius, <...>);
+  }
+
+  // 5. kills the Pingu
+  if (sprite.is_finished ()) {
+    pingu->set_status(PS_DEAD);
+  }
+}
+
+void Bomber::draw (SceneContext& gc) {
+  // 3. particles, terrain, explosion sprite
+  // 4. tick: hides the explosion sprite
+  if (sprite.frame()==13 && !gfx_exploded) {
+    gfx_exploded = true;
+    gc.color().draw(explo_surf, <...>);
+  }
+  gc.color().draw(sprite, pingu->get_pos());
+}
+\end{lstlisting}
+\centering\small{\ax Implementation in C++}
+\end{minipage}
+%
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+code/await Bomber (void) -> ActionName
+do
+  <...>
+  spawn Mover(); // movement in background
+  var Sprite sprite = spawn Sprite(<...>);
+               // frame animation in background
+  watching sprite do
+    // 1. plays a "Oh no!" sound.
+    {play_sound("ohno")};
+
+    // 2. plays a "Bomb!" sound.
+    await game.update until sprite.frame == 10;
+    {play_sound("plop")};
+
+    // 3. particles, terrain, explosion sprite
+    await game.update until sprite.frame == 13;
+    spawn PinguParticles(<...>) in particles;
+    call Game_Remove({&bomber_radius}, <...>);
+    do
+      <...>
+      spawn Sprite(<...>); // explosion
+
+      // 4. tick: hides the explosion sprite
+      await game.update;
+    end
+    await FOREVER;
+  end
+
+  // 5. kills the pingu
+  escape DEAD;
+end
+
+.
+\end{lstlisting}
+\centering\small{\bx Implementation in \CEU}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ The \emph{Bomber action} sequence.
+\label{lst.bomber}
+}
+\end{figure*}
 
@@
 primitive (ln. 4--5).
+In \CEU, if multiple abstractions react to the same event, the scheduler
+employs lexical order to preserve determinism, i.e., the \code{spawn} that
+appears first in the source code executes first~\cite{ceu.mod15}.
 The event \code{game.update} (ln. 12,16,24) is analogous to the \code{update}
@@
 
-\begin{figure}[t]
-\begin{verbatim}
-Action       Ceu    C++   Explicit State
----------   ----   ----   -----------------
-Bomber        23     50   4 state variables
-Bridger       75    100   2 state variables
-Drown          6     15   1 state variable
-Exiter         7     22   2 state variables
-Splashed       6     19   2 state variables
-\end{verbatim}
-\caption{Pingus actions in \CEU and C++.
-\label{tab.actions}
-}
-\end{figure}
-
 Pingus supports 16 actions in the game.
@@
 
+\begin{figure*}[!t]
+\centering
+\includegraphics[width=\textwidth]{continuation}
+\caption{ Continuation (C++) vs Direct (\CEU) Styles.
+\label{fig.story}
+}
+\end{figure*}
+
 The campaign world map has clickable blue dots in the two extremes of the
@@
 \centering
-\includegraphics[width=\textwidth]{continuation}
-\caption{ Continuation (C++) vs Direct (\CEU) Styles.
-\label{fig.story}
+\includegraphics[width=\textwidth]{hierarchy}
+\caption{Dispatching chain for \code{update}.
+\label{fig.hier}
+}
+\end{figure*}
+
+\begin{figure*}[t]
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+class Bomber : public PinguAction {
+    <...>
+    Sprite sprite;
+}
+
+Bomber::Bomber (<...>) : <...> {
+    sprite.load(<...>);
+    <...>
+}
+
+void Bomber::update () {
+    sprite.update();
+}
+
+void Bomber::draw (SceneContext& gc) {
+    <...>
+    gc.color().draw(sprite, <...>);
+}
+\end{lstlisting}
+\centering\small{\ax Implementation in C++}
+\end{minipage}
+%
+\begin{minipage}[t]{0.50\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+code/await Bomber (void) -> ActionName do
+    <...>
+    var Sprite sprite = spawn Sprite(<...>);
+    <...>
+end
+
+.
+\end{lstlisting}
+\centering\small{\bx Implementation in \CEU}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ \emph{Bomber action} \code{draw} and \code{update} dispatching.
+\label{lst.hier}
 }
@@
 
-\begin{figure*}[t]
-\centering
-\includegraphics[width=\textwidth]{hierarchy}
-\caption{Dispatching chain for \code{update}.
-\label{fig.hier}
-}
-\end{figure*}
-
-\begin{figure*}[t]
-\begin{minipage}[t]{0.50\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
-class Bomber : public PinguAction {
-    <...>
-    Sprite sprite;
-}
-
-Bomber::Bomber (<...>) : <...> {
-    sprite.load(<...>);
-    <...>
-}
-
-void Bomber::update () {
-    sprite.update();
-}
-
-void Bomber::draw (SceneContext& gc) {
-    <...>
-    gc.color().draw(sprite, <...>);
-}
-\end{lstlisting}
-\centering\small{\ax Implementation in C++}
-\end{minipage}
-%
-\begin{minipage}[t]{0.50\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
-code/await Bomber (void) -> ActionName do
-    <...>
-    var Sprite sprite = spawn Sprite(<...>);
-    <...>
-end
-
-.
-\end{lstlisting}
-\centering\small{\bx Implementation in \CEU}
-\end{minipage}
-%\rule{8.4cm}{0.37pt}
-\caption{ \emph{Bomber action} \code{draw} and \code{update} dispatching.
-\label{lst.hier}
-}
-\end{figure*}
-
 In C++, the class \code{Bomber} in Figure~\ref{lst.hier}.a declares a
@@
 
+\begin{figure*}[t]
+\begin{minipage}[t]{0.54\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+GameSession::GameSession(<...>) :
+    <...>
+{
+    <...>      // these widgets are always active...
+    btpanel  = new ButtonPanel(<...>);
+    pcounter = new PingusCounter(<...>);
+    smallmap = new SmallMap(<...>);
+    <...>
+    uimgr->add(btpanel);  // ...but are added
+    uimgr->add(pcounter); //  dynamically to the
+    uimgr->add(smallmap); //  dispatching hierarchy
+    <...>
+}
+\end{lstlisting}
+\centering\small{\ax Implementation in C++}
+\end{minipage}
+%
+\begin{minipage}[t]{0.46\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+code/await Game (void) do
+    <...> // other coexisting functionality
+    spawn ButtonPanel(<...>);
+    spawn PingusCounter(<...>);
+    spawn SmallMap(<...>);
+    <...> // other coexisting functionality
+end
+
+.
+\end{lstlisting}
+\centering\small{\bx Implementation in \CEU}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ Managing the UI widgets lifecycle.
+\label{lst.ui}
+}
+\end{figure*}
+
 In \CEU, the \code{Bomber} abstraction presented in Figure~\ref{lst.hier}.b
@@
 \end{figure}
-
-\begin{figure*}[t]
-\begin{minipage}[t]{0.54\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
-GameSession::GameSession(<...>) :
-    <...>
-{
-    <...>      // these widgets are always active...
-    btpanel  = new ButtonPanel(<...>);
-    pcounter = new PingusCounter(<...>);
-    smallmap = new SmallMap(<...>);
-    <...>
-    uimgr->add(btpanel);  // ...but are added
-    uimgr->add(pcounter); //  dynamically to the
-    uimgr->add(smallmap); //  dispatching hierarchy
-    <...>
-}
-\end{lstlisting}
-\centering\small{\ax Implementation in C++}
-\end{minipage}
-%
-\begin{minipage}[t]{0.46\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
-code/await Game (void) do
-    <...> // other coexisting functionality
-    spawn ButtonPanel(<...>);
-    spawn PingusCounter(<...>);
-    spawn SmallMap(<...>);
-    <...> // other coexisting functionality
-end
-
-.
-\end{lstlisting}
-\centering\small{\bx Implementation in \CEU}
-\end{minipage}
-%\rule{8.4cm}{0.37pt}
-\caption{ Managing the UI widgets lifecycle.
-\label{lst.ui}
-}
-\end{figure*}
 
@@
 
-TODO: non reactive, C++ integration
-- TODO: OO state + methods
-- eliminar estados explicitos com estruturas de controle apropriadas
-
-We promote the \emph{structured synchronous reactive} programming model of
-\CEU for the development of games.
-We present in-depth use cases categorized in four control-flow patterns applied
-to \emph{Pingus} (an open-source \emph{Lemmings} clone) that likely apply to
-other games.
-
-We show how the standard way to program games with objects and callbacks in C++
-hinders structured programming techniques, such as support for sequential
-execution, long-lasting loops, and persisting local variables.
-In this sense, callbacks actually disrupt structured programming, becoming
-["our generation’s goto"][goto] according to Miguel de Icaza.
-
-Overall, we believe that most difficulties in implementing control behavior in 
-game logic is not inherent to this domain, but a result of accidental
-complexity due to the lack of structured abstractions and an appropriate
-concurrency model to handle event-based applications.
-
-TODO: rever summaries por advantage qualitativa vs LoCs
-
-[goto]: tirania.org/blog/archive/2013/Aug-15.html
+We advocate \emph{Structured Synchronous Reactive Programming} as a productive
+alternative for game logic development.
+%
+We compare the implementation of six game behaviors in C++ and \CEU and discuss
+how structured reactive mechanisms can eliminate callbacks and let programmers
+write code in direct style.
+
+We categorize the behaviors in four recurrent control-flow patterns:
+%
+\emph{State machines} are the workhorses of the game logic, appearing in
+animations, AI behaviors, and input handling.
+\CEU can encode states implicitly with sequential statements in the same
+contiguous block, eliminating shared state variables and improving code
+encapsulation.
+%
+\emph{Continuation passing} controls the overall structure of the game, such as
+screen transitions and level progressions.
+Similarly to state machines, \CEU can describe the flow of the game with
+sequential statements in a self-contained block, eliminating explicit
+data structures and continuation variables (e.g., stacks and boolean flags)
+spread across multiple classes.
+%decouple activities by
+%
+\emph{Dispatching hierarchies} spread input events through the game entities
+and serve as a broadcast communication mechanism.
+Event broadcasting is at the core of the semantics of \CEU, allowing entities
+to react directly to inputs and bypass the program hierarchy entirely.
+In C++, a considerable amount of interleavning code in the game logic and
+engine is just destinated to event dispatching.
+%
+\emph{Lifespan hierarchies} manage the memory and visibility of game entities
+through class fields and containers.
+In \CEU, all entities have an associated lexical scope, similarly to local
+variables with automatic memory management.
+This reduces explicit manipulations of pointers and references considerably.
+
+Overall, we believe that most difficulties in implementing control-flow
+behavior in game logic is not inherent to this domain, but a result of
+accidental complexity due to the lack of structured abstractions and an
+appropriate concurrency model to handle event-based applications.
 
@@
 
-We would like to thank Leonardo Kaplan and Alexander Tkachov for early
+The author would like to thank Leonardo Kaplan and Alexander Tkachov for early
 explorations and prototypes of the game rewrite.

### 2017-06-01 paper.tex
...

@@
 %
-We demonstrate gains in productivity for six behaviors in the game logic of
-Pingus through a qualitative analysis of the proposed solutions in \CEU in
-comparison to the original implementations in C++.
+We show gains in productivity for six behaviors in Pingus through a qualitative
+analysis of the proposed solutions in \CEU in comparison to the original
+implementations in C++.
 %
@@
 ``our generation's \emph{goto}''.%
-\footnote{``Callbacks as our Generations' Go To Statement'':
+\footnote{``Callbacks as our Generations' \emph{goto} Statement'':
 \url{tirania.org/blog/archive/2013/Aug-15.html}}
@@
 
-Our case study demonstrates gains in productivity for six selected behaviors in
-the game logic of Pingus through idiomatic code in \CEU.
+Our case study shows gains in productivity for six selected behaviors in the
+game logic of Pingus through idiomatic code in \CEU.
 We present an in-depth qualitative analysis of the proposed solutions in
@@
 and \code{draw} events, bypassing the program hierarchy entirely.
-External events in \CEU are broadcasted to the entire application.
+Events in \CEU are broadcasted to the entire application in lexical order,
+i.e., an abstraction that appears first in the source code (e.g., ln. 3) reacts
+before another that appears second (e.g., ln. 4).
+As discussed in Section~\ref{sec.pats.fsms.2}, this rule preserves determinism
+and also respects the program static hierarchy.
 While (\emph{and only while}) the bomber abstraction is alive, the sprite

### 2017-06-24 paper.tex
...

@@
 
-\author{Francisco Sant'Anna
-        \\ Departamento de Inform\'atica e Ci\^encia da Computa\c{c}\~ao, UERJ
-        \\ francisco@ime.uerj.br}
+%\author{Francisco Sant'Anna
+        %\\ Departamento de Inform\'atica e Ci\^encia da Computa\c{c}\~ao, UERJ
+        %\\ francisco@ime.uerj.br}
 
@@
 \centering
-\includegraphics[width=\columnwidth]{pingus}
+\includegraphics[width=\columnwidth]{pingus-annotations}
 \caption{Pingus gameplay.
@@
 The codebase%
-\footnote{Pingus repository: \url{github.com/Pingus/pingus/}}
+\footnote{Official Pingus repository: \url{github.com/Pingus/pingus/}}
 is about 40.000 lines of code (LoCs), divided into
@@
 Object-oriented games rely on the \emph{observer pattern}~\cite{games.patterns}
-to handle events from the environment (e.g., key presses and timers) and also
+to handle events from the environment (e.g., key presses and timers), and also
 as a notification mechanism between entities in the game logic.
@@
 We present a case study of rewriting Pingus from C++ to the programming
-language \CEU~\cite{ceu.sensys13,ceu.mod15}, which has the following
-characteristics:
+language \CEU~\cite{ceu.sensys13,ceu.mod15}.
+\CEU has the following general characteristics:
 %
@@
 %\setlength\itemsep{0em}
-\item \emph{Reactive}: code only executes in reactions to events.
-\item \emph{Structured}: programs use structured control-flow mechanisms, such
+\item \emph{Reactive}: code executes in reactions to events.
+\item \emph{Structured}: code uses structured control-flow mechanisms, such
       as \code{spawn} and \code{await} (to create and suspend an activity).
@@
 Structured reactive programming eliminates callbacks, letting programmers write
-code in direct and sequential style and recover from the inversion of control
-imposed by the observer pattern~\cite{rp.deprecating}.
+code in direct and sequential style in order to recover from the inversion of
+control imposed by the observer pattern~\cite{rp.deprecating}.
 %
@@
 terms of memory and CPU usage~\cite{ceu.sensys13}.
-The runtime is single threaded and does not rely on garbage collection.
+The runtime is single threaded and does not rely on garbage collection for
+memory management.
 
 Our case study shows gains in productivity for six selected behaviors in the
-game logic of Pingus through idiomatic code in \CEU.
+game logic of Pingus rewritten in \CEU.
 We present an in-depth qualitative analysis of the proposed solutions in
@@
 %
-A control-flow pattern is a recurring technique to describe dependencies and
-explicit orders between statements (or groups of statements) execution in a
-program.
+A control-flow pattern is a recurring technique to describe execution
+dependency and explicit ordering between statements (or groups of statements).
 %For instance, consider how a key press stimulus propagates through the game
@@
 %game.
+We focus on a qualitative analysis for the programming techniques that we
+applied during the rewriting process.
+Not all techniques result in reduction in LoCs (especially considering the
+verbose syntax of \CEU), but have other effects such as eliminating shared
+variables and dependencies between classes.
+%
+We employed a \emph{live code rewrite}, i.e., starting from the original
+codebase in C++, we reimplemented it piece-by-piece in \CEU without breaking
+the game compilation and execution.
 
@@
       ------------     ----     ----     ----     --------------------------------------------
-game/            2064     2268     0.91     the main gameplay
-  ./              710      679     1.05         main functionality
-  objs/           470      478     0.98         world objects (tiles, traps, etc)
-  pingu/          884     1111     0.80         pingu behaviors
-    ./            343      458     0.75             main functionality
-    actions/      541      653     0.83             pingu actions (bomber, climber, etc)
-worldmap/         468      493     0.95     campaign worldmap
-screens/         1109     1328     0.84     menus and screens
-    option/       347      357     0.97         option menu
-    others/       762      971     0.78         other menus and screens
-misc/              56       46     1.22     miscellaneous functionality
+ 1    game/            2064     2268     0.91     the main gameplay
+ 2      ./              710      679     1.05         main functionality
+ 3      objs/           470      478     0.98         world objects (tiles, traps, etc)
+ 4      pingu/          884     1111     0.80         pingu behaviors
+ 5        ./            343      458     0.75             main functionality
+ 6        actions/      541      653     0.83             pingu actions (bomber, climber, etc)
+ 7    worldmap/         468      493     0.95     campaign worldmap
+ 8    screens/         1109     1328     0.84     menus and screens
+ 9        option/       347      357     0.97         option menu
+10        others/       762      971     0.78         other menus and screens
+11    misc/              56       46     1.22     miscellaneous functionality
                      ----     ----     ----
@@
 \end{figure*}
-
-The rewriting process consisted of identifying sets of callbacks implementing
-control flow in the game and translating them to \CEU using appropriate
-structured constructs.
-%
-As an example, a double mouse click is characterized by a first click, followed
-by a maximum amount of time, followed by a second click.
-This behavior depends on different events (clicks and timers) which have to
-occur in a particular order.
-In C++, the implementation involves callbacks crossing reactions to successive
-events which manipulate state variables explicitly.
-%
-We focus on a qualitative analysis for the programming techniques that we
-applied during the rewriting process.
-Not all techniques result in reduction in LoCs (especially considering the
-verbose syntax of \CEU), but have other effects such as eliminating shared
-variables and dependencies between classes.
-%
-We employed \emph{live code translation}, i.e., starting from the original
-codebase in C++, we reimplemented piece-by-piece without breaking the game
-compilation and execution.
-%
-This was only possible given the seamless integration between \CEU and
-C/C++~\cite{ceu.sensys13}: the type systems are equivalent and the integration
-happens at the source code level.
-This enables trivial sharing of control and data, i.e., accessing C/C++ data
-and calling C/C++ from \CEU and vice-versa.
 
@@
 
-\section{The Pingus Codebase}
+\section{The Pingus Codebase and Rewriting Process}
 \label{sec.codebase}
 
-In Pingus, the game logic accounts for almost half the size of the codebase:
+% cd /opt/pingus/src
+% sloccount .           -> 39975 -> (39362?)
+% cd pingus/
+% sloccount .           -> 18173
+
+In Pingus, the game logic accounts for almost half the size of the codebase%
+\footnote{
+We used \emph{SLOCCount} to count only non-blank, non-comment lines in the
+codebase: \url{www.dwheeler.com/sloccount/}
+}:
 18.173 from 39.362 LoCs (46\%) spread across 272 files.
@@
 dealing with configurations and options, saved games and serialization, maps
-and levels descriptions, string formatting, collision detection, graph
+and level descriptions, string formatting, collision detection, graph
 algorithms, etc.
@@
 Therefore, we only rewrote 9.186 LoCs spread across 126 files%
-\footnote{Complete codebase: \url{github.com/an000/p/tree/master/cpp}}.
-%
-In order to only consider effective code in the analysis, we then removed all
+\footnote{\label{codebase} Effective codebase: \url{github.com/an000/p/tree/master/}}.
+%
+In order to only consider relevant code in the analysis, we then removed all
 headers, declarations, trivial getters \& setters, and other innocuous
@@
 originally written in C++%
-\footnote{C++ codebase: \url{github.com/an000/p/tree/master/all}}.
+\footnotemark[\ref{codebase}].
+%\footnote{C++ codebase: \url{github.com/an000/p/tree/master/all}}.
 We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs%
-\footnote{\CEU codebase: \url{github.com/an000/p/tree/master/all}}.
+\footnotemark[\ref{codebase}].
+%\footnote{\CEU codebase: \url{github.com/an000/p/tree/master/all}}.
 %
@@
 
-Although our analysis is mostly qualitative, the lines in Figure~\ref{tab.tree}
+Although our analysis is mostly qualitative, the rows in Figure~\ref{tab.tree}
 with lower ratio numbers correlate to the parts of the game logic that we
 consider more susceptible to structured reactive programming.
-For instance, the \emph{Pingu} behavior (\emph{ratio 0.80}) contains complex
-animations that are affected by timers, game rules, and user interaction.
-In contrast, the \emph{Option screen} (\emph{ratio 0.97}) is a simple UI grid
-with trivial mouse interactions.
+For instance, the \emph{Pingu} behavior (row 4, \emph{ratio 0.80}) contains
+complex animations that are affected by timers, game rules, and user
+interaction.
+In contrast, the \emph{Option screen} (row 9, \emph{ratio 0.97}) is a simple UI
+grid with trivial mouse interactions.
+
+The rewriting process consisted of identifying sets of callbacks implementing
+control flow in the game and translating them to \CEU using appropriate
+structured constructs.
+%
+As an example, a double mouse click is characterized by a first click, followed
+by a maximum amount of time, followed by a second click.
+This behavior depends on different events (clicks and timers) which have to
+occur in a particular order.
+In C++, the implementation involves callbacks crossing reactions to successive
+events which manipulate state variables explicitly.
+%
+This was only possible given the seamless integration between \CEU and
+C/C++~\cite{ceu.sensys13}: the type systems are equivalent and the integration
+happens at the source code level.
+This enables trivial sharing of control and data, i.e., accessing C/C++ data
+and calling C/C++ from \CEU and vice-versa.
 
@@
 
-During the course of the rewriting process, we have identified four abstract
-control-flow patterns which likely apply to other games as well:
+During the rewriting process, we have identified four abstract control-flow
+patterns which likely apply to other games as well:
 
@@
 patterns and analyse their implementations in C++ and \CEU.%
-\footnote{Due to space constraints, we omit a fifth pattern
-\emph{Signaling Mechanisms} and three other game behaviors.}
+\footnote{Due to space constraints, we omit three other game behaviors and a
+fifth pattern \emph{Signaling Mechanisms}.}
 
@@
 mouse and timer events.
-The code focus on the double click detection and hides unrelated parts with
-\code{<...>}.
+The code in the figure focus on the double click detection and hides unrelated
+parts with \code{<...>}.
 %
@@
 behavior in C++ as a state machine.
-The circles represent the state of the variables in the class, while the arrows 
+The circles represent the state of the variables in the class, and the arrows 
 represent the callbacks manipulating state.
@@
 self-contained in the \code{loop} body.
-Also, these 7 lines of code \emph{only} detect the double click, leaving the
+Also, those 7 lines of code \emph{only} detect the double click, leaving the
 actual effect to happen outside the loop (ln. 12), as well as all unrelated
@@
 In C++, the class \code{Bomber} in Figure~\ref{lst.bomber}.a defines the
-callbacks \code{draw} and \code{update} to manage the state machine described
-above.
+callbacks \code{draw} and \code{update} to manage the state machine of
+Figure~\ref{fig.bomber.fsm}.
 %
@@
 The \code{update} callback (ln. 14--38) first updates the pingu animation and
-movement on every frame regardless of its current state (ln. 15--16).
+movement on every frame, regardless of its current state (ln. 15--16).
 When the animation reaches the 10th frame, it switches to \emph{state-2} and
@@
 %
-Note that a single numeric state variable suffices to track the states
-(Figure~\ref{fig.bomber.fsm}), but the original developers probably chose to
+Note that a single numeric state variable would suffice to track the states as
+in Figure~\ref{fig.bomber.fsm}, but the original developers probably chose to
 encode each state in an independent boolean 
@@
 Pingus supports 16 actions in the game.
-As Figure~\ref{tab.actions} shows, 5 of them implement at least one state
-machine and are considerable smaller in \CEU in terms of LoCs after eliminating
-the state variables.
-%
-Considering the other 11 actions, the reduction in LoCs is negligible.
-This asymmetry in the implementation of actions illustrates the gains in
-expressiveness when describing state machines in direct style.
+5 of them, enumerated in Figure~\ref{tab.actions}, implement at least one state
+machine and are considerable smaller in \CEU in terms of LoCs (after
+eliminating the state variables).
+%
+For the other 11 actions without state machines, the reduction in LoCs is
+negligible.
+%
+This asymmetry illustrates the gains in expressiveness when describing state
+machines in direct style.
 
@@
 
-Among all 65 implementation files in \CEU, we found 29 cases in 25 files using
-structured mechanisms to substitute states machines.
+Among all 65 implementation files in \CEU, we found 29 cases in 25 files that
+use structured mechanisms to substitute states machines.
 They typically manifest as \code{await} statements in sequence.
@@
 
-\subsubsection{ Transition to the \emph{Credits Screen} from the \emph{Story Screen} }
+\subsubsection{ Transition from the \emph{Story Screen}  to the \emph{Credits Screen}}
 
@@
 StoryDot::StoryDot(const FileReader& reader) :
-  credits(false), // do not display by default
+  show_credits(false), // do not show by default
 {
   <...>
-  reader.read("credits", credits); // from file
-}
+  reader.read("credits", show_credits);
+}             // from file
 
@@
   <...>
-  push_screen(<StoryScreen>(<...>, credits));
+  push_screen(<StoryScreen>(show_credits));
   <...>
@@
 StoryScreenComp::StoryScreenComp (<...>) :
-  credits(credits),
+  show_credits(show_credits),
   <...>
@@
     } else {
-      if (credits) {
+      if (show_credits) {
         replace_screen(<Credits>(<...>));
@@
 progress trail for introductory and ending ambience stories, respectively.
-For introductory stories, the game returns to the world map after displaying 
+For introductory stories, the game returns to the world map after showing 
 the story pages.
-For ending stories, the game also displays a \emph{Credits screen} before
+For ending stories, the game also shows a \emph{Credits screen} before
 returning to the world map.%
@@
 reads the level file (ln. 5) to check whether its an ending story and should,
-after termination, display the \emph{Credits screen}.
+after termination, show the \emph{Credits screen}.
 %
@@
 \code{StoryScreen} (ln. 10) and represents the screen continuation, i.e., what
-to do after displaying the story.
+to do after showing the story.
 The class \code{StoryScreen} (not shown) then forwards the continuation even
@@
 In \CEU, the \code{loop} of Figure~\ref{lst.story}.b controls the flow between
-the screens to display as a direct sequence of statements.
+the screens to show as a direct sequence of statements.
 %
@@
 and await its termination (ln. 5).
-Finally, we check the returned values (ln. 6) to perhaps display the
-\code{Credits} screen (ln. 8).
+After showing the story, we check the returned values (ln. 6) to perhaps show
+the \code{Credits} screen (ln. 8).
 The enclosing loop restores the \code{Worldmap} and repeats the process.
@@
 \begin{lstlisting}[numbers=left,xleftmargin=3em]
-class Bomber : public PinguAction {
+class Bomber : public Action {
     <...>
@@
 
-void Bomber::draw (SceneContext& gc) {
-    <...>
-    gc.color().draw(sprite, <...>);
+void Bomber::draw () {
+    <...>
+    sprite.draw();
 }
@@
 
-In contrast with C++, the screens in \CEU are completely decoupled and only the
+In contrast with C++, the screens in \CEU are decoupled and only the
 \code{Main Loop} touches them: the \code{Worldmap} has no references to
@@
 It also uses the same technique for the loop that switches the pingu actions
-during gameplay (e.g., \emph{walking} to \emph{jumping} and back to
+during gameplay (e.g., \emph{walking} to \emph{falling} and back to
 \emph{walking}).
@@
 \item \code{ScreenManager::display} in the main game loop calls
-      \code{ScreenManager::update}.
-\item \code{ScreenManager::update} calls \code{last\_screen->update} for the
-      active game screen (i.e., a \code{GameSession} instance, considering the
-      \code{Bomber}).
+      \code{ScreenManager::update} when starting a new frame.
+\item \code{ScreenManager::update} calls \code{screen->update} for the active
+      game screen (i.e., a \code{GameSession} instance, considering the screen
+      in which the \code{Bomber} appears).
 \item \code{GameSession::update} calls \code{world->update}.
-\item \code{World::update} calls \code{obj->update} for each object in the
+\item \code{World::update} calls \code{objs->update} for each object in the
       world.
@@
 %\setlength\itemsep{0em}
-\item With a single assignment to \code{last\_screen}, we can easily deactivate
-      the current screen and redirect all dispatches to a new screen (step 2).
+\item With a single assignment to \code{screen}, one can easily deactivate the
+current screen and redirect all dispatches to a new screen (step 2).
 \item The \code{World} class manages and dispatches events to all game
-      entities with a common interface \code{WorldObj}, such as all pingus and
-      traps (step 4).
+      entities with a common interface \code{WorldObj}, such as all pingus
+      together and traps (step 4).
 \item Since it is common to iterate only over the pingus (vs. all world
@@
 i.e., an abstraction that appears first in the source code (e.g., ln. 3) reacts
-before another that appears second (e.g., ln. 4).
+before another one that appears second (e.g., ln. 4).
 As discussed in Section~\ref{sec.pats.fsms.2}, this rule preserves determinism
@@
 eliminates dispatching chains entirely.
-Note that one can still declare a local event and restrict its visibility like
+Note that one can still declare a local event to restrict its visibility like
 a local variable.
@@
 
-\subsubsection{Case Study: Game UI Widgets}
+\subsubsection{Case Study: Static Game UI Widgets}
 
@@
 
-\subsubsection{Case Study: Managing the Pingus Lifecycle}
+\subsubsection{Case Study: Dynamic Pingus Lifecycle}
 \label{sec.pats.lifespan.2}
@@
 %
-Without the \code{remove} call, a dead pingu would remain in the collection
-still with updates on every frame (ln. 11).
+If the programmer disregards the call to \code{remove}, a dead pingu would
+remain in the collection and still update on every frame (ln. 11).
 Since the \code{draw} behavior for a dead pingu is innocuous, the death could
@@
 %
-Entities with dynamic lifespan in C++ always require explicit \code{add} and
-\code{remove} calls associated to a container (ln. 4,13).
+Entities with dynamic lifespan in C++ always require explicit matching
+\code{add} and \code{remove} calls associated to a container (ln. 4,13).
 
@@
 Pools are used to hold explosion particles, levels and level sets from files,
-gameplay \& worldmap objects, and UI widgets.
+gameplay \& worldmap objects, and also UI widgets.
 
@@
 The control-flow patterns closely relate to the \emph{GoF} behavioral
-patterns~\cite{gof}, which some previous work discuss in the context of video
-games~\cite{games.patterns,games.gof.2015,games.gof.2007}.
+patterns~\cite{gof}, which are discussed in the context of video games in
+previous work~\cite{games.patterns,games.gof.2015,games.gof.2007}.
 %
@@
     \emph{observer} (to handle events in general)
-patterns as implementation details to achieve the desired higher-level
+patterns as implementation techniques to achieve the desired higher-level
 control-flow patterns.
@@
 %
-Instead, we rewrote an existing game and did it in small steps while keeping it
-working.
+%Instead, we rewrote an existing game and did it in small steps while keeping it
+%working.
 Although we do not provide a performance evaluation (Pingus is not performance
@@
 write code in direct style.
+%
+In the end, we rewrote about 1/4 of the whole codebase (9.186 from 39.362 lines
+of code) which comprises the core of the game logic that is susceptible to
+structured reactive programming.
+%We employed a \emph{live code rewrite}, i.e., starting from the original
+%codebase in C++, we reimplemented it piece-by-piece in \CEU without breaking
+%the game compilation and execution.
 
@@
 screen transitions and level progressions.
-Similarly to state machines, \CEU can describe the flow of the game with
-sequential statements in a self-contained block, eliminating explicit
-data structures and continuation variables (e.g., stacks and boolean flags)
-spread across multiple classes.
+Similarly to state machines, \CEU describes the flow of the game as sequential
+statements in self-contained blocks, eliminating explicit data structures and
+continuation variables spread across multiple classes (e.g., stacks and boolean
+flags).
 %decouple activities by
 %
-\emph{Dispatching hierarchies} spread input events through the game entities
-and serve as a broadcast communication mechanism.
+\emph{Dispatching hierarchies} disseminate input events through the game
+entities and serve as a broadcast communication mechanism.
 Event broadcasting is at the core of the semantics of \CEU, allowing entities
 to react directly to inputs and bypass the program hierarchy entirely.
-In C++, a considerable amount of interleavning code in the game logic and
-engine is just destinated to event dispatching.
+%In C++, a considerable amount of interleaving code in the game logic and
+%engine is just destined to event dispatching.
 %
@@
 variables with automatic memory management.
-This reduces explicit manipulations of pointers and references considerably.
+This reduces explicit manipulation of pointers and references considerably.
 
@@
 
-The author would like to thank Leonardo Kaplan and Alexander Tkachov for early
+The authors would like to thank Leonardo Kaplan and Alexander Tkachov for early
 explorations and prototypes of the game rewrite.

### 2017-06-30 paper.tex
...

@@
 \newcommand{\CEU}{\textsc{C\'{e}u}\xspace}
+\newcommand{\locs}{\emph{locs}\xspace}
 \newcommand{\code}[1] {{\small{\texttt{#1}}}}
@@
 We show gains in productivity for six behaviors in Pingus through a qualitative
-analysis of the proposed solutions in \CEU in comparison to the original
-implementations in C++.
+analysis of the proposed implementations in \CEU in comparison to the originals
+in C++.
 %
@@
 %
-Pingus is developed in standard object-oriented C++, the \emph{lingua franca}
-of game development \cite{games.patterns}.
+Pingus is developed in standard object-oriented C++, ``the lingua franca of
+game development'' \cite{games.patterns}.
 The codebase%
 \footnote{Official Pingus repository: \url{github.com/Pingus/pingus/}}
-is about 40.000 lines of code (LoCs), divided into
+is about 40.000 lines of code (\locs), divided into
 the engine, level editor, auxiliary libraries, and the game logic itself.
@@
 game development resides in \emph{simulation} (aka \emph{game logic}), but
-which accounts for only 10\% of the CPU budget~\cite{games.sweeney}.
+which only accounts for 10\% of the CPU budget~\cite{games.sweeney}.
 %(Figure~\ref{fig.sweeney}).
@@
 Object-oriented games rely on the \emph{observer pattern}~\cite{games.patterns}
-to handle events from the environment (e.g., key presses and timers), and also
+to handle events from the environment (e.g., key presses and timers) and also
 as a notification mechanism between entities in the game logic.
@@
       as \code{spawn} and \code{await} (to create and suspend an activity).
-\item \emph{Synchronous}: reactions run atomically and to completion on each
-      activity.
+\item \emph{Synchronous}: event reactions never overlap and run atomically and
+      to completion on each activity.
       There is no implicit preemption or real parallelism, resulting in
@@
 %
-Structured reactive programming eliminates callbacks, letting programmers write
-code in direct and sequential style in order to recover from the inversion of
-control imposed by the observer pattern~\cite{rp.deprecating}.
+Structured reactive programming recovers from the inversion of control imposed
+by callbacks and the observer pattern~\cite{rp.deprecating}, letting
+programmers write code in direct and sequential style.
 %
@@
 comparison to the original implementations in C++.
-We also identify four recurrent control-flow patterns that likely apply to
-other games:
+We also identify four control-flow patterns that likely apply to other games:
         \emph{Finite State Machines},
@@
 A control-flow pattern is a recurring technique to describe execution
-dependency and explicit ordering between statements (or groups of statements).
+dependency and/or explicit ordering between statements.
+% (or groups of statements).
 %For instance, consider how a key press stimulus propagates through the game
@@
 applied during the rewriting process.
-Not all techniques result in reduction in LoCs (especially considering the
+Not all techniques result in reduction of \locs (especially considering the
 verbose syntax of \CEU), but have other effects such as eliminating shared
@@
 }:
-18.173 from 39.362 LoCs (46\%) spread across 272 files.
+18.173 from 39.362 \locs (46\%) spread across 272 files.
 %
@@
 algorithms, etc.
-This part remains unchanged and relies on the integration between \CEU and
-C/C++.
-%
-Therefore, we only rewrote 9.186 LoCs spread across 126 files%
+This part remains unchanged and relies on the seamless integration between \CEU
+and C/C++~\cite{ceu.sensys13}: the type systems are equivalent and the
+integration happens at the source code level.
+This enables accessing data and calling C/C++ from \CEU and vice-versa.
+%
+Therefore, we only rewrote 9.186 \locs spread across 126 files%
 \footnote{\label{codebase} Effective codebase: \url{github.com/an000/p/tree/master/}}.
@@
 headers, declarations, trivial getters \& setters, and other innocuous
-statements, resulting 4.135 dense LoCs spread across 70 implementation files
+statements, resulting 4.135 dense \locs spread across 70 implementation files
 originally written in C++%
@@
 %\footnote{C++ codebase: \url{github.com/an000/p/tree/master/all}}.
-We did the same with the implementation in \CEU, resulting in 3.697 dense LoCs%
+We did the same with the implementation in \CEU, resulting in 3.697 dense \locs%
 \footnotemark[\ref{codebase}].
@@
 
-Although our analysis is mostly qualitative, the rows in Figure~\ref{tab.tree}
-with lower ratio numbers correlate to the parts of the game logic that we
-consider more susceptible to structured reactive programming.
+Although the analysis in this work is mostly qualitative, the rows with lower
+ratio numbers in Figure~\ref{tab.tree} do correlate with the parts of the game
+logic that we consider more susceptible to structured reactive programming.
 For instance, the \emph{Pingu} behavior (row 4, \emph{ratio 0.80}) contains
@@
 
-The rewriting process consisted of identifying sets of callbacks implementing
-control flow in the game and translating them to \CEU using appropriate
-structured constructs.
+The rewriting process consisted of identifying sets of callbacks in C++
+implementing control flow in the game and translating them to \CEU using
+appropriate structured constructs.
 %
@@
 %
-This was only possible given the seamless integration between \CEU and
-C/C++~\cite{ceu.sensys13}: the type systems are equivalent and the integration
-happens at the source code level.
-This enables trivial sharing of control and data, i.e., accessing C/C++ data
-and calling C/C++ from \CEU and vice-versa.
-
-As a general rewriting rule, we could identify control-flow behaviors in the
-C++ codebase by looking for class members with identifiers resembling verbs,
+As a general rewriting rule, we identify control-flow behaviors in the C++
+codebase by looking for class state members with identifiers resembling verbs,
 statuses, and counters (e.g.,
@@
 
-During the rewriting process, we have identified four abstract control-flow
-patterns which likely apply to other games as well:
+During the rewriting process, we have identified four abstract cause/effect
+control-flow patterns which likely apply to other games as well:
 
@@
         if (press_time > 1.0f) {
-            pressed = false; // give up, 1st click
-            press_time = 0;  // was too long ago
+            pressed = false; // give up, 1st click was
+            press_time = 0;  //         too long ago
         }
@@
 do
-    var RectComponent r = <...>;
+    var RectComponent but = <...>;
     <...>
     loop do
-        await r.on_click;
+        await but.on_click;
         watching 1s do
-            await r.on_click;
+            await but.on_click;
             break;
@@
 behavior in C++ as a state machine.
-The circles represent the state of the variables in the class, and the arrows 
-represent the callbacks manipulating state.
-%
-Note in Figure~\ref{lst.armageddon} how the accesses to the state variables are
-spread across the entire class.
-For instance, the distance between the initialization of \code{pressed} (ln.
-3) and the last access to it (ln. 32) is over 40 lines in the original file.
+The circles represent the state of the variable \code{pressed}, and the arrows 
+represent the callbacks manipulating it.
+%
+Note in Figure~\ref{lst.armageddon}.a how the accesses to the state variables
+are spread across the entire class: the distance between the initialization of
+\code{pressed} (ln.  3) and the last access to it (ln. 32) is over 40 lines in
+the original file.
 Arguably, this dispersion of code across methods makes the understanding and 
@@
 Also, those 7 lines of code \emph{only} detect the double click, leaving the
-actual effect to happen outside the loop (ln. 12), as well as all unrelated
-code such as redrawing the button.
+actual effect (ln. 12) as well as all unrelated code (such as redrawing the
+button) to happen outside the loop.
+
+The \code{await} allows for nested control-flow structures to suspend execution
+while retaining all enclosing state alive.
+Then, a subsequent reaction to an event resumes execution normally.
+In contrast, method callbacks have a single entry point at the top level of the
+class, in which only instance members remain active between invocations.
 
@@
 employs lexical order to preserve determinism, i.e., the \code{spawn} that
-appears first in the source code executes first~\cite{ceu.mod15}.
+appears first in the source code reacts first~\cite{ceu.mod15}.
 The event \code{game.update} (ln. 12,16,24) is analogous to the \code{update}
@@
 %\setlength\itemsep{0em}
-\item They encode all states with direct sequential code, eliminating shared
-      state variables for control-flow purposes.
+\item They encode all states with direct sequential code, eliminating callbacks
+      and shared state variables for control-flow purposes.
 \item They handle all states (and only them) in the same contiguous block,
@@
 However, this approach is not fundamentally different from Pingus' use of
-\code{switch} or \code{if} branches for each possible state.
+\code{switch} or \code{if} branches to decode state.
 
 Pingus supports 16 actions in the game.
-5 of them, enumerated in Figure~\ref{tab.actions}, implement at least one state
-machine and are considerable smaller in \CEU in terms of LoCs (after
-eliminating the state variables).
-%
-For the other 11 actions without state machines, the reduction in LoCs is
+Five of them implement at least one state machine and are considerable smaller
+in \CEU in terms of \locs (Figure~\ref{tab.actions}).
+%(after eliminating the state variables).
+%
+For the other 11 actions without state machines, the reduction in \locs is
 negligible.
@@
 The campaign world map has clickable blue dots in the two extremes of the
-progress trail for introductory and ending ambience stories, respectively.
+progress trail to show introductory and closing ambience stories, respectively.
 For introductory stories, the game returns to the world map after showing 
 the story pages.
-For ending stories, the game also shows a \emph{Credits screen} before
+For closing stories, the game also shows a \emph{Credits screen} before
 returning to the world map.%
@@
 In C++, the class \code{StoryDot} in Figure~\ref{lst.story}.a (ln. 1--12) first
-reads the level file (ln. 5) to check whether its an ending story and should,
+reads the level file (ln. 5) to check whether its a closing story and should,
 after termination, show the \emph{Credits screen}.
 %
-The boolean variable \code{credits} is passed to the class
+The boolean variable \code{show\_credits} (ln. 2,5,10) is passed to the class
 \code{StoryScreen} (ln. 10) and represents the screen continuation, i.e., what
@@
 %
-When the method \code{next\_text} has no pages left to display (ln. 32--38),
-it decides where to go next, depending on the continuation flag
-\code{credits} (ln. 33).
+When the method \code{next\_text} has no story pages left to display
+(ln. 32--38), it decides where to go next, depending on the continuation flag
+\code{show\_credits} (ln. 33).
 
@@
 %
-We first invoke the \code{Worldmap} (ln. 2), which exhibits the map and let
+We first invoke the \code{Worldmap} (ln. 2), which shows the map and lets
 the player interact with it (e.g., walking around) until a dot is clicked.
@@
     %\setlength\itemsep{0em}
-    \item C++ uses an explicit stack to push the \code{Worldmap} screen.
+    \item C++ uses an explicit stack to push the \code{Worldmap} screen (not
+          shown in Figure~\ref{lst.hier}.a).
     \item \CEU invokes the \code{Worldmap} screen expecting a return value
-          (ln. 2).
+          (Figure~\ref{lst.hier}.b, ln. 2).
     \end{itemize}
@@
     \item C++ pushes the \code{Story} screen passing the continuation flag
-          (ln. 10).
+          (Figure~\ref{lst.hier}.a, ln. 10).
     \item \CEU stores the \code{Worldmap} return value and invokes the \code{Story} screen
-          (ln. 2,5).
+          (Figure~\ref{lst.hier}.b, ln. 2,5).
     \end{itemize}
@@
     \item C++ replaces the current \code{Story} screen with the \code{Credits}
-          screen (ln. 34).
+          screen (Figure~\ref{lst.hier}.a, ln. 34).
     \item \CEU invokes the \code{Credits} screen after the \code{await Story}
-          returns (ln. 8).
+          returns (Figure~\ref{lst.hier}.b, ln. 8).
     \end{itemize}
@@
     \item C++ pops the \code{Credits} screen, going back to the \code{Worldmap}
-          screen.
-          \CEU uses an enclosing \code{loop} to restart the process (ln. 1--13).
+          screen (not shown in Figure~\ref{lst.hier}.a).
+    \item \CEU uses an enclosing \code{loop} to restart the process
+          (Figure~\ref{lst.hier}.b, ln. 1--13).
     \end{itemize}
@@
 
-In contrast with C++, the screens in \CEU are decoupled and only the
-\code{Main Loop} touches them: the \code{Worldmap} has no references to
-\code{Story}, which has no references to \code{Credits}.
+In contrast with C++, the screens in \CEU are decoupled from each other and
+only the \code{Main Loop} touches them: the \code{Worldmap} has no references
+to \code{Story}, which has no references to \code{Credits}.
+Changing the screen arrangements is a matter of adjusting the main loop only.
 
@@
 The direct style of \CEU has some advantages in comparison to the 
-continuation-passing style:
+continuation-passing style of C++:
 %
@@
 
-Continuation passing typically controls the overall structure of the game,
-such as screen transitions in menus and level progressions.
-%
-\CEU uses the direct style techniques in five cases involving screen
+Continuation passing typically controls the overall structure of the game in
+C++, such as screen transitions in menus and level progressions.
+%
+\CEU adopts the direct style technique in five cases involving screen
 transitions:
@@
 %
-It also uses the same technique for the loop that switches the pingu actions
-during gameplay (e.g., \emph{walking} to \emph{falling} and back to
+It also uses the same technique for the loop that switches between pingu
+actions during gameplay (e.g., \emph{walking} to \emph{falling} and back to
 \emph{walking}).
@@
 \subsubsection{Case Study: \emph{Bomber Action} \Code{draw} and \Code{update} Dispatching}
+
+\begin{figure*}[t]
+\begin{minipage}[t]{0.54\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+GameSession::GameSession(<...>) :
+    <...>
+{
+    <...>      // these widgets are always active...
+    btpanel  = new ButtonPanel(<...>);
+    pcounter = new PingusCounter(<...>);
+    smallmap = new SmallMap(<...>);
+    <...>
+    uimgr->add(btpanel);  // ...but are added
+    uimgr->add(pcounter); //  dynamically to the
+    uimgr->add(smallmap); //  dispatching hierarchy
+    <...>
+}
+\end{lstlisting}
+\centering\small{\ax Implementation in C++}
+\end{minipage}
+%
+\begin{minipage}[t]{0.46\linewidth}
+\begin{lstlisting}[numbers=left,xleftmargin=3em]
+code/await Game (void) do
+    <...> // other coexisting functionality
+    spawn ButtonPanel(<...>);
+    spawn PingusCounter(<...>);
+    spawn SmallMap(<...>);
+    <...> // other coexisting functionality
+end
+
+.
+\end{lstlisting}
+\centering\small{\bx Implementation in \CEU}
+\end{minipage}
+%\rule{8.4cm}{0.37pt}
+\caption{ Managing the UI widgets lifecycle.
+\label{lst.ui}
+}
+\end{figure*}
 
@@
 \item The \code{World} class manages and dispatches events to all game
-      entities with a common interface \code{WorldObj}, such as all pingus
-      together and traps (step 4).
+      entities with a common interface \code{WorldObj}, such as the pingus and
+      traps (step 4).
 \item Since it is common to iterate only over the pingus (vs. all world
@@
 
-\begin{figure*}[t]
-\begin{minipage}[t]{0.54\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
-GameSession::GameSession(<...>) :
-    <...>
-{
-    <...>      // these widgets are always active...
-    btpanel  = new ButtonPanel(<...>);
-    pcounter = new PingusCounter(<...>);
-    smallmap = new SmallMap(<...>);
-    <...>
-    uimgr->add(btpanel);  // ...but are added
-    uimgr->add(pcounter); //  dynamically to the
-    uimgr->add(smallmap); //  dispatching hierarchy
-    <...>
-}
-\end{lstlisting}
-\centering\small{\ax Implementation in C++}
-\end{minipage}
-%
-\begin{minipage}[t]{0.46\linewidth}
-\begin{lstlisting}[numbers=left,xleftmargin=3em]
-code/await Game (void) do
-    <...> // other coexisting functionality
-    spawn ButtonPanel(<...>);
-    spawn PingusCounter(<...>);
-    spawn SmallMap(<...>);
-    <...> // other coexisting functionality
-end
-
-.
-\end{lstlisting}
-\centering\small{\bx Implementation in \CEU}
-\end{minipage}
-%\rule{8.4cm}{0.37pt}
-\caption{ Managing the UI widgets lifecycle.
-\label{lst.ui}
-}
-\end{figure*}
-
 In \CEU, the \code{Bomber} abstraction presented in Figure~\ref{lst.hier}.b
@@
 As discussed in Section~\ref{sec.pats.fsms.2}, this rule preserves determinism
-and also respects the program static hierarchy.
+and also conforms to the program static hierarchy.
 While (\emph{and only while}) the bomber abstraction is alive, the sprite
-animation remains alive.
+animation remains alive and reacts to the \code{update} and \code{draw} events.
 The radical decoupling between the program hierarchy and reactions to events
@@
 games also rely heavily on this control-flow pattern.
-In the case of the Pingus engine, we rewrote 9 files from C++ to \CEU, reducing
-them from 515 to 173 LoCs (not listed in Figure~\ref{tab.tree}), mostly due to
-dispatching code removal.
+In the case of the Pingus engine, we rewrote 9 files with a reduction from 515
+to 173 \locs (not listed in Figure~\ref{tab.tree}), mostly due to dispatching
+code removal.
 
@@
 Figure~\ref{fig.ui} shows the game UI widgets with action buttons, score
-counters, and a small map which all coexist with the game screen during its
+counters, and a small map, all coexisting with the game screen during its
 whole lifespan.
@@
       missing calls to \code{remove} lead to memory and CPU leaks (to be
-      discussed as the \emph{lapsed listener} problem in
+      discussed as the \emph{lapsed listener problem} in
       Section~\ref{sec.pats.lifespan.2}).
 \end{itemize}
-
-In \CEU, the UI entities that coexist are just created in the same lexical
-block of the abstraction \code{Game} in Figure~\ref{lst.ui}.b (ln. 3--5).
-%
-Since abstractions can react independently, they do not require a dispatching
-container.
-%
-Lexical lifespan never requires containers, allocation and deallocation, or
-explicit references.
-In addition, all required memory is known at compile time, similarly to
-stack-allocated local variables.
-%
-The \emph{Bomber action} of Section~\ref{sec.pats.fsms.2} also relies on
-lexical scope to delimit the lifespan of the explosion sprite to a single
-frame (Figure~\ref{lst.bomber}, ln. 19--25).
-
-\subsubsection{Case Study: Dynamic Pingus Lifecycle}
-\label{sec.pats.lifespan.2}
-
-%\begin{figure}[t]
-%\centering
-%\includegraphics[width=\columnwidth]{pingus_create_die-anim}
-%\caption{Creation and death of pingus.
-%\label{fig.pingus_create_die}
-%}
-%\end{figure}
-
-A pingu is a dynamic entity created periodically and destroyed under certain
-conditions, such as falling from a high altitude.%
-\footnote{Death of pingus animation: \url{github.com/an000/p/#5} }
-%
 
@@
 
+In \CEU, the UI entities that coexist are simply created in the same lexical
+block of the \code{Game} abstraction in Figure~\ref{lst.ui}.b (ln. 3--5).
+%
+Since abstractions can react independently, they do not require a dispatching
+container.
+%
+Lexical lifespan never requires containers, allocation and deallocation, or
+explicit references.
+In addition, all required memory is known at compile time, similarly to
+stack-allocated local variables.
+%
+The \emph{Bomber action} of Section~\ref{sec.pats.fsms.2} also relies on
+lexical scope to delimit the lifespan of the explosion sprite into a single
+frame (Figure~\ref{lst.bomber}, ln. 19--25).
+
+\subsubsection{Case Study: Dynamic Pingus Lifecycle}
+\label{sec.pats.lifespan.2}
+
+%\begin{figure}[t]
+%\centering
+%\includegraphics[width=\columnwidth]{pingus_create_die-anim}
+%\caption{Creation and death of pingus.
+%\label{fig.pingus_create_die}
+%}
+%\end{figure}
+
+A pingu is a dynamic entity created periodically and destroyed under certain
+conditions, such as falling from a high altitude.%
+\footnote{Death of pingus animation: \url{github.com/an000/p/#5} }
+%
+
 In C++, the class \code{PinguHolder} in Figure~\ref{lst.pingus}.a is a
@@
 In addition, the \code{spawn} statement supports a pool identifier to associate
-the new instance with a pool.
+a new instance with a pool.
 %
@@
 (ln. 1--9) instead of the enclosing scope of the \code{spawn} statement
-(ln. 4--7).
+(ln. 4--7), surviving the call to \code{Pingu\_Spawn}.
 Since pools are also subject to lexical scope, the lifespan of all dynamically
@@
       to immediate reclamation.
-\item Abstraction instances (static or dynamic) never require explicit
-      manipulation of pointers/references.
+\item Instances (static or dynamic) never require explicit manipulation of
+      pointers/references.
 \end{itemize}
@@
 instantiations across 37 files.
-Pools are used to hold explosion particles, levels and level sets from files,
-gameplay \& worldmap objects, and also UI widgets.
+Pools are used to hold explosion particles, levels and level sets loaded from
+files, gameplay \& worldmap objects, and also UI widgets.
 
@@
 patterns as implementation techniques to achieve the desired higher-level
-control-flow patterns.
-%
-\CEU overcomes the need of behavioral patterns with a semantics that supports
-structured control-flow mechanisms and event-based communication via broadcast.
-%
-As an example, the \emph{state pattern} for the bomber animation in C++ in
-Section~\ref{sec.pats.fsms} becomes a series of blocks separated by
-\code{await} statements in \CEU.
+control-flow patterns described in the paper.
+%
+\CEU overcomes the need of behavioral patterns with support for structured
+control-flow mechanisms and event-based communication via broadcast.
+%
+%As an example, the \emph{state pattern} for the bomber animation in C++ in
+%Section~\ref{sec.pats.fsms} becomes a series of blocks separated by
+%\code{await} statements in \CEU.
 
@@
 %
-In the end, we rewrote about 1/4 of the whole codebase (9.186 from 39.362 lines
+Ultimately, we rewrote about 1/4 of the whole codebase (9.186 from 39.362 lines
 of code) which comprises the core of the game logic that is susceptible to
@@
 animations, AI behaviors, and input handling.
-\CEU can encode states implicitly with sequential statements in the same
-contiguous block, eliminating shared state variables and improving code
-encapsulation.
+\CEU can encode states implicitly with sequential statements,
+% in the same contiguous block,
+eliminating shared state variables and improving code encapsulation.
 %
@@
 statements in self-contained blocks, eliminating explicit data structures and
-continuation variables spread across multiple classes (e.g., stacks and boolean
-flags).
+continuation variables.
+%spread across multiple classes (e.g., stacks and boolean flags).
 %decouple activities by
@@
 variables with automatic memory management.
-This reduces explicit manipulation of pointers and references considerably.
+%This reduces explicit manipulation of pointers and references considerably.
 

### 2017-07-20 paper.tex
...

@@
 \begin{verbatim}
-      Path              Ceu      C++   Ceu/C++    Descritpion
+      Path              Ceu      C++   Ceu/C++    Description
       ------------     ----     ----     ----     --------------------------------------------
@@
 We describe six representative game behaviors in detail distributed in the four
-patterns and analyse their implementations in C++ and \CEU.%
+patterns and analyze their implementations in C++ and \CEU.%
 \footnote{Due to space constraints, we omit three other game behaviors and a
@@
 
-The \code{await} allows for nested control-flow structures to suspend execution
-while retaining all enclosing state alive.
+The \code{await} allows for nested control-flow statements to suspend execution
+while retaining all enclosing state alive, such as local variables and next
+statement to execute.
 Then, a subsequent reaction to an event resumes execution normally.
@@
 class, in which only instance members remain active between invocations.
+In particular, locals and loops cannot persist across invocations.
 
@@
 \end{verbatim}
-\caption{Pingus actions in \CEU and C++.
+\caption{Pingus actions in \CEU and C++ in terms of \locs and state variables.
 \label{tab.actions}
@@
 eliminates dispatching chains entirely.
-Note that one can still declare a local event to restrict its visibility like
-a local variable.
+%Note that one can still declare a local event to restrict its visibility like
+%a local variable.
 
@@
 %
-If the programmer disregards the call to \code{remove}, a dead pingu would
-remain in the collection and still update on every frame (ln. 11).
+Note that if the programmer disregards the call to \code{remove}, a dead pingu
+would remain in the collection and still update on every frame (ln. 11).
 Since the \code{draw} behavior for a dead pingu is innocuous, the death could
@@
 %
-Entities with dynamic lifespan in C++ always require explicit matching
+Hence, entities with dynamic lifespan always require explicit matching
 \code{add} and \code{remove} calls associated to a container (ln. 4,13).
@@
 %
-The original Pingus in C++ uses variations of the
+The original Pingus in C++ uses variations of the patterns
     \emph{state} (Sections~\ref{sec.pats.fsms}~and~\ref{sec.pats.cps}),
@@
     \emph{observer} (to handle events in general)
-patterns as implementation techniques to achieve the desired higher-level
+as implementation techniques to achieve the desired higher-level
 control-flow patterns described in the paper.
 %
-\CEU overcomes the need of behavioral patterns with support for structured
-control-flow mechanisms and event-based communication via broadcast.
+\CEU overcomes the need of behavioral patterns with support, at the language
+level, for structured control-flow mechanisms and event-based communication via
+broadcast.
 %
@@
 FRP has been successfully used to implement a 3D first person shooting game
-from scratch, but with performance considerations~\cite{games.frag}.
+from scratch, but with some performance considerations~\cite{games.frag}.
 %
@@
 %
+We use the video game \emph{Pingus} as a case study.
 We compare the implementation of six game behaviors in C++ and \CEU and discuss
@@
 accidental complexity due to the lack of structured abstractions and an
-appropriate concurrency model to handle event-based applications.
+appropriate concurrency model to develop event-based applications.
 

### 2017-08-03 paper.tex
(>) small adjusts

@@
 use structured mechanisms to substitute states machines.
-They typically manifest as \code{await} statements in sequence.
+They typically manifest as \code{await} statements in sequence (e.g.,
+ln. 5,7 in Figure~\ref{lst.armageddon} and ln. 12,16,24 in
+Figure~\ref{lst.bomber}).
 %or in aborting constructs such as \code{par/or} and \code{watching}.
@@
 
-\subsubsection{ Transition from the \emph{Story Screen}  to the \emph{Credits Screen}}
+\subsubsection{ Transition from \emph{Story}  to \emph{Credits} Screen}
 
@@
 The campaign world map has clickable blue dots in the two extremes of the
-progress trail to show introductory and closing ambience stories, respectively.
+map road to show introductory and closing ambience stories, respectively.
 For introductory stories, the game returns to the world map after showing 
@@
 In C++, the class \code{StoryDot} in Figure~\ref{lst.story}.a (ln. 1--12) first
-reads the level file (ln. 5) to check whether its a closing story and should,
+reads the level file (ln. 5) to check whether it is a closing story and should,
 after termination, show the \emph{Credits screen}.
@@
 eliminates dispatching chains entirely.
+% TODO: hidden in the language runtime
 %Note that one can still declare a local event to restrict its visibility like

### 2017-08-03 paper.tex
(>) ln. -> ln

@@
 %
-The methods \code{update} (ln. 14--26) and \code{on\_click} (ln. 28--34) are
+The methods \code{update} (ln 14--26) and \code{on\_click} (ln 28--34) are
 examples of \emph{short-lived callbacks}, which are pieces of code that execute
@@
 The callback \code{on\_click} reacts to mouse clicks detected by the base class
-\code{RectComponent} (ln. 2), while the callback \code{update} continuously
+\code{RectComponent} (ln 2), while the callback \code{update} continuously
 reacts to the passage of time, frame by frame.
@@
 %
-The class first initializes the variable \code{pressed} (ln. 3) to track the
-first click (ln. 32).
-It also initializes the variable \code{press\_time} (ln. 4) to count the time
-since the first click (ln. 16--17).
+The class first initializes the variable \code{pressed} (ln 3) to track the
+first click (ln 32).
+It also initializes the variable \code{press\_time} (ln 4) to count the time
+since the first click (ln 16--17).
 If another click occurs within 1 second, the class signals the double click to
-the application (ln. 29--30).
+the application (ln 29--30).
 Otherwise, the \code{pressed} and \code{press\_time} state variables are reset
-(ln. 18--21). 
+(ln 18--21).
 %
@@
 are spread across the entire class: the distance between the initialization of
-\code{pressed} (ln.  3) and the last access to it (ln. 32) is over 40 lines in
+\code{pressed} (ln  3) and the last access to it (ln 32) is over 40 lines in
 the original file.
@@
 Also, even though the state variables are private, unrelated methods such as 
-\code{draw}, which is defined in middle of the class (ln. 10--12), can
+\code{draw}, which is defined in middle of the class (ln 10--12), can
 potentially access them.
@@
 %
-In Figure~\ref{lst.armageddon}.b, the loop to detect double clicks (ln. 4--10)
-awaits the first click (ln. 5) and then, while watching 1 second (ln. 6--9),
-awaits the second click (ln. 7).
+In Figure~\ref{lst.armageddon}.b, the loop to detect double clicks (ln 4--10)
+awaits the first click (ln 5) and then, while watching 1 second (ln 6--9),
+awaits the second click (ln 7).
 If the second click occurs within 1 second, the \code{break} terminates the
-loop (ln. 8) and the \code{emit} in sequence signals the double click to the
-application (ln. 12).
+loop (ln 8) and the \code{emit} in sequence signals the double click to the
+application (ln 12).
 Otherwise, the \code{watching} block as a whole aborts after 1 second  and the
-loop restarts, falling back to the first click \code{await} (ln. 5).
+loop restarts, falling back to the first click \code{await} (ln 5).
 %
@@
 Also, those 7 lines of code \emph{only} detect the double click, leaving the
-actual effect (ln. 12) as well as all unrelated code (such as redrawing the
+actual effect (ln 12) as well as all unrelated code (such as redrawing the
 button) to happen outside the loop.
@@
 The class first defines one state variable for each effect to perform
-(ln. 4--7).
+(ln 4--7).
 The ``Oh no!'' sound plays as soon as the object starts in \emph{state-1} 
-(ln. 11).
-The \code{update} callback (ln. 14--38) first updates the pingu animation and
-movement on every frame, regardless of its current state (ln. 15--16).
+(ln 11).
+The \code{update} callback (ln 14--38) first updates the pingu animation and
+movement on every frame, regardless of its current state (ln 15--16).
 When the animation reaches the 10th frame, it switches to \emph{state-2} and
-plays the ``Bomb!'' sound (ln. 18--22).
+plays the ``Bomb!'' sound (ln 18--22).
 The state variable \code{sound\_played} is required because the sprite frame
@@
 \code{update} may execute twice during the 10th frame).
-The same reasoning and technique applies to \emph{state-3} (ln. 24--32 and
+The same reasoning and technique applies to \emph{state-3} (ln 24--32 and
 41--46).
-The explosion sprite appears in a single frame in \emph{state-4} (ln. 45).
-Finally, the pingu dies after the animation frames terminate (ln. 34--37).
+The explosion sprite appears in a single frame in \emph{state-4} (ln 45).
+Finally, the pingu dies after the animation frames terminate (ln 34--37).
 %
@@
 \code{code/await} abstractions and execute in separate through the \code{spawn}
-primitive (ln. 4--5).
+primitive (ln 4--5).
 In \CEU, if multiple abstractions react to the same event, the scheduler
@@
 appears first in the source code reacts first~\cite{ceu.mod15}.
-The event \code{game.update} (ln. 12,16,24) is analogous to the \code{update}
+The event \code{game.update} (ln 12,16,24) is analogous to the \code{update}
 callback of C++ and occurs on every game frame.
 %
-The code tracks the animation aliveness (ln. 7--27) and, on termination,
-performs the last bomber effect, killing the pingu (ln. 30).
-As soon as the animation starts, the code performs the first effect (ln. 9).
+The code tracks the animation aliveness (ln 7--27) and, on termination,
+performs the last bomber effect, killing the pingu (ln 30).
+As soon as the animation starts, the code performs the first effect (ln 9).
 The intermediate effects are performed when the corresponding conditions occur
-(ln. 12,16,24).
-The \code{do-end} block (ln. 19--25), restricts the lifespan of the
-single-frame explosion sprite (ln. 21): after the next game tick (ln. 24), the
+(ln 12,16,24).
+The \code{do-end} block (ln 19--25), restricts the lifespan of the
+single-frame explosion sprite (ln 21): after the next game tick (ln 24), the
 block terminates and automatically destroys the spawned abstraction (removing
@@
 In contrast with the implementation in C++, all effects occur in a contiguous
-chunk of code (ln. 7--30), which handles no extra functionality.
+chunk of code (ln 7--30), which handles no extra functionality.
 
@@
 They typically manifest as \code{await} statements in sequence (e.g.,
-ln. 5,7 in Figure~\ref{lst.armageddon} and ln. 12,16,24 in
+ln 5,7 in Figure~\ref{lst.armageddon} and ln 12,16,24 in
 Figure~\ref{lst.bomber}).
@@
 
-In C++, the class \code{StoryDot} in Figure~\ref{lst.story}.a (ln. 1--12) first
-reads the level file (ln. 5) to check whether it is a closing story and should,
+In C++, the class \code{StoryDot} in Figure~\ref{lst.story}.a (ln 1--12) first
+reads the level file (ln 5) to check whether it is a closing story and should,
 after termination, show the \emph{Credits screen}.
 %
-The boolean variable \code{show\_credits} (ln. 2,5,10) is passed to the class
-\code{StoryScreen} (ln. 10) and represents the screen continuation, i.e., what
+The boolean variable \code{show\_credits} (ln 2,5,10) is passed to the class
+\code{StoryScreen} (ln 10) and represents the screen continuation, i.e., what
 to do after showing the story.
 The class \code{StoryScreen} (not shown) then forwards the continuation even
-further to the auxiliary class \code{StoryScreenComp} (ln. 16--40).
+further to the auxiliary class \code{StoryScreenComp} (ln 16--40).
 %
 When the method \code{next\_text} has no story pages left to display
-(ln. 32--38), it decides where to go next, depending on the continuation flag
-\code{show\_credits} (ln. 33).
+(ln 32--38), it decides where to go next, depending on the continuation flag
+\code{show\_credits} (ln 33).
 
@@
 %
-We first invoke the \code{Worldmap} (ln. 2), which shows the map and lets
+We first invoke the \code{Worldmap} (ln 2), which shows the map and lets
 the player interact with it (e.g., walking around) until a dot is clicked.
-If the player selects a story dot (ln. 4--9), we invoke the \code{Story}
-and await its termination (ln. 5).
-After showing the story, we check the returned values (ln. 6) to perhaps show
-the \code{Credits} screen (ln. 8).
+If the player selects a story dot (ln 4--9), we invoke the \code{Story}
+and await its termination (ln 5).
+After showing the story, we check the returned values (ln 6) to perhaps show
+the \code{Credits} screen (ln 8).
 The enclosing loop restores the \code{Worldmap} and repeats the process.
@@
     \item \CEU invokes the \code{Worldmap} screen expecting a return value
-          (Figure~\ref{lst.hier}.b, ln. 2).
+          (Figure~\ref{lst.hier}.b, ln 2).
     \end{itemize}
@@
     \item C++ pushes the \code{Story} screen passing the continuation flag
-          (Figure~\ref{lst.hier}.a, ln. 10).
+          (Figure~\ref{lst.hier}.a, ln 10).
     \item \CEU stores the \code{Worldmap} return value and invokes the \code{Story} screen
-          (Figure~\ref{lst.hier}.b, ln. 2,5).
+          (Figure~\ref{lst.hier}.b, ln 2,5).
     \end{itemize}
@@
     \item C++ replaces the current \code{Story} screen with the \code{Credits}
-          screen (Figure~\ref{lst.hier}.a, ln. 34).
+          screen (Figure~\ref{lst.hier}.a, ln 34).
     \item \CEU invokes the \code{Credits} screen after the \code{await Story}
-          returns (Figure~\ref{lst.hier}.b, ln. 8).
+          returns (Figure~\ref{lst.hier}.b, ln 8).
     \end{itemize}
@@
     \item \CEU uses an enclosing \code{loop} to restart the process
-          (Figure~\ref{lst.hier}.b, ln. 1--13).
+          (Figure~\ref{lst.hier}.b, ln 1--13).
     \end{itemize}
@@
 In C++, the class \code{Bomber} in Figure~\ref{lst.hier}.a declares a
-\code{sprite} member (ln. 3) to handle its animation frames.
+\code{sprite} member (ln 3) to handle its animation frames.
 %
@@
 \code{draw} requests from the game and forward them to the sprite
-(ln. 11--13 and 15--18).
+(ln 11--13 and 15--18).
 %
@@
 In \CEU, the \code{Bomber} abstraction presented in Figure~\ref{lst.hier}.b
-spawns a \code{Sprite} animation instance on its body (ln. 3).
+spawns a \code{Sprite} animation instance on its body (ln 3).
 %
@@
 Events in \CEU are broadcasted to the entire application in lexical order,
-i.e., an abstraction that appears first in the source code (e.g., ln. 3) reacts
-before another one that appears second (e.g., ln. 4).
+i.e., an abstraction that appears first in the source code (e.g., ln 3) reacts
+before another one that appears second (e.g., ln 4).
 As discussed in Section~\ref{sec.pats.fsms.2}, this rule preserves determinism
@@
 In C++, the widgets are created in the constructor of the class
-\code{GameSession} in Figure~\ref{lst.ui}.a (ln. 5--7), added to a UI container
-(ln. 9--11), and are never removed since they must always be visible.
+\code{GameSession} in Figure~\ref{lst.ui}.a (ln 5--7), added to a UI container
+(ln 9--11), and are never removed since they must always be visible.
 Arguably, to better express the intent of making them coexist with the game
@@
 In \CEU, the UI entities that coexist are simply created in the same lexical
-block of the \code{Game} abstraction in Figure~\ref{lst.ui}.b (ln. 3--5).
+block of the \code{Game} abstraction in Figure~\ref{lst.ui}.b (ln 3--5).
 %
@@
 lexical scope to delimit the lifespan of the explosion sprite into a single
-frame (Figure~\ref{lst.bomber}, ln. 19--25).
+frame (Figure~\ref{lst.bomber}, ln 19--25).
 
@@
 %
-The method \code{PinguHolder::create\_pingu} (ln. 1--6) is called periodically
+The method \code{PinguHolder::create\_pingu} (ln 1--6) is called periodically
 to create a new \code{Pingu} and add it to the \code{pingus} collection
-(ln. 3--4).
-The method \code{PinguHolder::update} (ln. 8--18) checks the state of all
-pingus on every frame, removing those with the dead status (ln. 12--14).
+(ln 3--4).
+The method \code{PinguHolder::update} (ln 8--18) checks the state of all
+pingus on every frame, removing those with the dead status (ln 12--14).
 %
 Note that if the programmer disregards the call to \code{remove}, a dead pingu
-would remain in the collection and still update on every frame (ln. 11).
+would remain in the collection and still update on every frame (ln 11).
 Since the \code{draw} behavior for a dead pingu is innocuous, the death could
@@
 Hence, entities with dynamic lifespan always require explicit matching
-\code{add} and \code{remove} calls associated to a container (ln. 4,13).
+\code{add} and \code{remove} calls associated to a container (ln 4,13).
 
@@
 The game screen in Figure~\ref{lst.pingus}.b spawns a new \code{Pingu} on every
-invocation of \code{Pingu\_Spawn} (ln. 4--7).
-%
-The \code{spawn} statement (ln. 6) specifies the pool declared at the top-level
-block of the game screen (ln. 3).
+invocation of \code{Pingu\_Spawn} (ln 4--7).
+%
+The \code{spawn} statement (ln 6) specifies the pool declared at the top-level
+block of the game screen (ln 3).
 In this case, the lifespan of the new instances follows the scope of the pool
-(ln. 1--9) instead of the enclosing scope of the \code{spawn} statement
-(ln. 4--7), surviving the call to \code{Pingu\_Spawn}.
+(ln 1--9) instead of the enclosing scope of the \code{spawn} statement
+(ln 4--7), surviving the call to \code{Pingu\_Spawn}.
 Since pools are also subject to lexical scope, the lifespan of all dynamically
@@
 %
-The \code{escape} statement (ln. 17) aborts the execution block of the
+The \code{escape} statement (ln 17) aborts the execution block of the
 \code{Pingu} instance, removing it from its associated pool automatically.
